---
title: "Cell-to-cell communication analysis"
author: "Sara Wernig-Zorc"
date: "`r Sys.Date()`"
output: 
  html_document: 
    toc: yes
    toc_depth: 5
    toc_float: yes
    collapsed: false
    code_folding: hide
    highlight: pygments
    theme: spacelab
    df_print: paged
params:
  input: "/home/data/MapMet/"
  out: "/home/out/MapMet/"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      tidy = TRUE, 
                      error = FALSE, 
                      messages = FALSE,
                      warning = FALSE)

library(parallel)
future::plan(strategy = 'future::multicore', workers = 12)
options(future.globals.maxSize = 8 * 10^9)
```

## __Project description__ 
<br>
MapMet project: IMC data (Lazic et al. 2024)
<br>
Analysis done on scRNA-seq data from Fetahu et al., 2023
<br>
1. heatmaps: hierarhical clustering with column normalization 
(by subtracting the column mean)
<br>
2. label transfer from IMC data to scRNA-seq data
<br>
3. CCC analysis 
<br>
4. T-cell exhustion and activation markers
<br>
We also check if PD1+ T-cells co-express the exhustion and activation markers
to further suppport our hypothesis that T-cells are exhusted in NB tumors. 
Meaning they have identified the tumor cells as tumor cells, but they can no
longer attack them/eradicate them, as they are exhusted.
<br>
5. Label transfer from MONACO immune cells dataset to T-cell clusters 
<br>
Here we would like to identify T-cells sub-types. We first look at the 
CD4/CD8/S100B/GZMB expression per T-cell cluster. Second, as the CD4 RNA
expression is known to be low, we also do a label transfer based on closest
expression profile (with SingleR). 
<br>
Analysis done in R v4.3.2 in a docker (mapmet_v2024-01-19.Dockerfile) container
<br>
docker exec -it --user root d04b26a70393 /bin/bash
<br>

# __Load libraries__

```{r R packages: install, eval=FALSE, include=FALSE, echo=FALSE}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("OmnipathR",force = TRUE)
BiocManager::install("ComplexHeatmap",force = TRUE)
BiocManager::install("presto",force = TRUE)
BiocManager::install("Biobase",force = TRUE)
BiocManager::install("BiocGenerics",force = TRUE)
BiocManager::install("BiocNeighbors",force = TRUE)
BiocManager::install("SingleR")
devtools::install_github("immunogenomics/presto")

install.packages("tidyverse", clean = TRUE)
install.packages("stringr", clean = TRUE)

install.packages('igraph', repos = c(igraph = 'https://igraph.r-universe.dev',
                           CRAN = 'https://cloud.r-project.org'))

devtools::install_github("jinworks/CellChat")

install.packages('Cairo', repo='https://RForge.net')

devtools::install_github('VPetukhov/ggrastr')

devtools::install_version('speedglm', version = '0.3-4', 
                          repos = 'https://packagemanager.rstudio.com/cran/2023-03-31')

remotes::install_github(repo = 'cole-trapnell-lab/monocle3')


install.packages("magick")
BiocManager::install("SpatialExperiment",force = TRUE)
```

```{r R packages: load, include=FALSE}
library("Seurat")
library("SingleCellExperiment")
library("tidyverse")
library("OmnipathR")
library("ComplexHeatmap")
library("stringr")
library("tidyverse")
library("CellChat")
library("monocle3")
library("SpatialExperiment")
library("SingleR")
library("magrittr")
library("CellChat")
library("RColorBrewer")
```

# __Input data__

Results of CCC analysis from Fetahu et al., 2023

```{r input data Fetahu et al - published, eval=FALSE, include=FALSE}
# Already analysed cell to cell communication using CellChat v 1.1.0
cellchat <- readRDS("/home/data/MapMet/fetahu.scRNA.2023/data/ccc_cellchat_object.rds")
signaling <- readRDS("/home/data/MapMet/fetahu.scRNA.2023/data/ccc_signaling_data.rds")
```

```{r input data Fetahu et al - new}
# New analysis with CellChat v2.1.1
scrna_data <- readRDS("/home/data/MapMet/fetahu.scRNA.2023/data/rna_decontaminated.rds")
metadata <- readRDS("/home/data/MapMet/fetahu.scRNA.2023/data/metadata.rds")

nb_metadata <- 
  metadata %>% 
  select(
    cell, library_size, n_features, sample,
    group, cellont_name, cellont_abbr, 
  ) %>% 
  filter(cellont_abbr != "other", group != "I") %>% 
  mutate(
    cellont_abbr =
      cellont_abbr %>% 
      fct_drop() %>% 
      fct_relevel("NB", "pDC", "M", "B", "T", "NK", "SC", "E")
  ) %>% 
  column_to_rownames("cell")

expr_data <- 
  scrna_data %>% 
  assay("soupx_counts") %>% 
  normalizeData(scale.factor = 1e6) %>% 
  extract(, rownames(nb_metadata))
expr_data 
```

# __Subset RNA data by IMC markers__

```{r input data Lazic et al}
protein2gene <- read.csv("/home/data/MapMet/lazic.imc.2024/gene_list/protein2gene.csv", 
                         sep=";", header = TRUE)
imc_data <- readRDS("/home/data/MapMet/lazic.imc.2024/Rds/imc_seurat.rds")

head(protein2gene)

IMC_markers <- protein2gene %>% filter(used_for_clustering == "yes")

gene.list <- IMC_markers$gene_name
```


```{r subset Fetahu et al., 2023 data by Lazic et al., 2024 gene markers}
# Filter gene expression matrix by the 42 genes used in IMC
expr_data_39markers <- expr_data[rownames(expr_data) %in% gene.list, ]

#saveRDS(expr_data_39markers, 
#        file = "/home/out/MapMet/Rds/fetahu2023.expr_data.39markers.IMC.Rds")

expr_data_39markers <- as.matrix(expr_data_39markers)
```


# __CCC analysis__

CellChat v2.1.1

## All metaclusters

```{r CellChat: meta clusters}
# New analysis with CellChat v2.1.1
cellchat <- createCellChat(
  expr_data,
  meta = nb_metadata,
  group.by = "cellont_abbr",
)
cellchat@DB <- CellChatDB.human

# only keep genes in CellChatDB
cellchat <- subsetData(cellchat)

# find overexpressed ligands/receptors
cellchat <- identifyOverExpressedGenes(cellchat)

# find overexpressed interactions
cellchat <- identifyOverExpressedInteractions(cellchat)

# calculate communication probability between cell groups
cellchat <- computeCommunProb(cellchat, seed.use = 1)

# filter communications occurring with less than 10 cells in a cell group
cellchat <- filterCommunication(cellchat)

# calculate communication probability on the signaling pathway level
cellchat <- computeCommunProbPathway(cellchat)

# calculate aggregated communication network
cellchat <- aggregateNet(cellchat)

# determine network centrality pathways
cellchat <- netAnalysis_computeCentrality(cellchat)

# single LR dotplot data
df_net <-
  cellchat %>% 
  subsetCommunication(thresh = NA) %>% 
  as_tibble() %>% 
  group_by(interaction_name_2) %>% 
  mutate(prob.norm = prob / max(prob)) %>% 
  ungroup() %>% 
  mutate(
    source = factor(source, c("NB", "pDC", "M", "B", "T", "NK", "SC", "E")),
    target = factor(target, c("NB", "pDC", "M", "B", "T", "NK", "SC", "E")),
  )

df_net

cellchat %>% saveRDS(paste0(params$out,"Rds/fetahu2023.cellchat.rds"))
df_net %>% saveRDS(paste0(params$out,"/home/out/MapMet/Rds/fetahu2023.signaling.rds"))

```

Reproduce data from the Fetahu et al., 2023 paper (to check that the new version 
of cellChat gives similar results)

```{r CellChat figures}
library(latex2exp)
source("/home/data/MapMet/fetahu.scRNA.2023/scripts/csbg-neuroblastoma-d976641/styling.R")

#cellchat <- readRDS("/home/out/MapMet/Rds/fetahu2023.cellchat.rds")
#sig_data <- readRDS("/home/out/MapMet/Rds/fetahu2023.signaling.rds")


## 3a ----

plot_n_interactions <- function() {
  rc_order <- names(CELL_TYPE_ABBREVIATIONS)
  mat <- cellchat@net$count[rc_order, rc_order]
  total_target <- colSums(mat)
  total_source <- rowSums(mat)
  bar_ylim <- c(0, max(total_source, total_target))
  
  Heatmap(
    mat,
    col = RColorBrewer::brewer.pal(9, "YlOrBr"),
    name = "number of\ninteractions",
    
    cluster_rows = FALSE, 
    row_names_gp = gpar(fontsize = BASE_TEXT_SIZE_PT), 
    row_names_side = "left",
    row_title = "source (ligand)", 
    row_title_gp = gpar(fontsize = BASE_TEXT_SIZE_PT), 
    
    cluster_columns = FALSE, 
    column_names_gp = gpar(fontsize = BASE_TEXT_SIZE_PT), 
    column_title = "target (receptor)",
    column_title_side = "bottom",
    column_title_gp = gpar(fontsize = BASE_TEXT_SIZE_PT), 
    
    width = unit(20, "mm"),
    height = unit(20, "mm"),
    
    top_annotation = HeatmapAnnotation(
      count_bar = anno_barplot(
        total_target,
        ylim = bar_ylim,
        border = FALSE, 
        axis = FALSE,
        gp = gpar(fill = "gray70", col = "gray70")
      ),
      count_text = anno_text(
        total_target, 
        gp = gpar(fontsize = BASE_TEXT_SIZE_PT),
      ), 
      simple_anno_size_adjust = TRUE,
      show_annotation_name = FALSE
    ), 
    
    right_annotation = rowAnnotation(
      count_text = anno_text(
        total_source, 
        gp = gpar(fontsize = BASE_TEXT_SIZE_PT),
      ), 
      count_bar = anno_barplot(
        total_source,
        ylim = bar_ylim,
        border = FALSE, 
        axis = FALSE,
        gp = gpar(fill = "gray70", col = "gray70")
      ),
      simple_anno_size_adjust = TRUE,
      show_annotation_name = FALSE
    ), 
    
    heatmap_legend_param = list(
      at = c(min(mat), max(mat)),
      title_gp = gpar(fontsize = BASE_TEXT_SIZE_PT), 
      title_position = "topleft", 
      border = NA, 
      legend_height = unit(20, "mm"), 
      labels_gp = gpar(fontsize = BASE_TEXT_SIZE_PT), 
      grid_width = unit(2, "mm")
    )
  )
}

(p <- plot_n_interactions())
ggsave(paste0(params$out,"/figures/3a_n_interactions.png"), plot = p, width = 6, height = 5)



## 3b ----

plot_contribution_celltype <- function(cell_type = "NB", signif = 0.05) {
   sig_data %>%
    filter(
      pathway_name %in% cellchat@netP$pathways, 
      source == {{cell_type}}, 
      target != {{cell_type}},
      pval <= signif
    ) %>%
    group_by(interaction = interaction_name_2) %>%
    summarise(
      prob.norm = sum(prob.norm),
      count = n(),
      pathway_name = first(pathway_name)
    ) %>%
    ungroup() %>%
    mutate(prob.avg = prob.norm / count) %>% 
    mutate(interaction = fct_reorder(interaction, prob.norm)) %>% 
    ggplot(aes(interaction, prob.norm)) +
    geom_col(fill = "#ffb03e", width = 0.85) +
    geom_hline(yintercept = 0, size = BASE_LINE_SIZE) +
    geom_text(
      aes(y = 2.68, label = pathway_name),
      size = BASE_TEXT_SIZE_MM,
      fontface = "italic"
    ) +
    xlab("ligand-receptor pair") +
    scale_y_continuous(
      "total outgoing communication score from NB",
      expand = expansion(0)
    ) +
    coord_flip() +
    theme_nb(grid = FALSE) +
    theme(
      axis.title.x = element_text(hjust = 1),
      axis.ticks.y = element_blank(),
      panel.border = element_blank(),
      panel.grid.major.x = element_line(
        color = "grey92",
        size = BASE_LINE_SIZE
      )
    )
}

plot_contribution_celltype()
ggsave(paste0(params$out,"3b_outgoing_comm_score.png"), width = 5.5, height = 5)


## 3c ----

plot_selected_dots <- function(interactions, source_type = "NB") {
  vis_data <- 
    sig_data %>% 
    filter(
      interaction_name %in% {{interactions}}, 
      source == {{source_type}}
    ) %>%
    mutate(
      pathway_name = factor(pathway_name),
      source = fct_recode(source, "from NB (source) to" = "NB"),
      target = factor(target, levels = names(CELL_TYPE_ABBREVIATIONS))
    )
  
  ggplot(vis_data, aes(target, interaction_name_2)) +
    geom_point(
      aes(color = prob.norm, size = pmin(5, -log10(pval)))
    ) + 
    scale_radius(
      TeX("-log_{10} p-Value_{cap.}"),
      range = c(1, 5),
      guide = "none"
    ) +
    scale_color_distiller(
      "relative communication score",
      type = "seq",
      palette = "YlOrBr",
      direction = 1,
      breaks = round(range(vis_data$prob.norm), 2),
      guide = guide_colorbar(
        barheight = unit(2, "mm"),
        barwidth = unit(15, "mm"),
        ticks = FALSE,
        title.position = "top"
      )
    ) +
    xlab("cell type (target)") +
    ylab("ligand-receptor pair\nand pathway") +
    facet_grid(
      vars(pathway_name), 
      vars(source),
      scales = "free",
      space = "free",
      switch = "both"
    ) + 
    theme_nb(grid = FALSE) + 
    theme(
      axis.text.x = element_text(
        size = BASE_TEXT_SIZE_PT, 
        margin = margin(t = 1, b = 1, r = 1, l = 1, unit = "mm")
      ),
      axis.text.y = element_text(size = BASE_TEXT_SIZE_PT),
      axis.ticks.length.x = unit(2, "mm"),
      axis.ticks.x = element_line(
        arrow = arrow(length = unit(1,"mm"), ends = "first", type = "closed")
      ),
      axis.title = element_text(size = BASE_TEXT_SIZE_PT),
      
      panel.border = element_rect(color = "black", size = BASE_LINE_SIZE),
      panel.spacing = unit(-BASE_LINE_SIZE / 2, "pt"),
      
      strip.background.y = element_rect(
        fill = "gray95",
        size = BASE_LINE_SIZE
      ),
      strip.background.x = element_rect(
        fill = "gray95",
        size = BASE_LINE_SIZE
      ),
      strip.text.x =  element_text(
        size = BASE_TEXT_SIZE_PT, 
        margin = margin(t = 1, b = 1, r = 4, l = 4, unit = "mm"),
        angle = 0
      ),
      strip.text.y.left = element_text(
        size = BASE_TEXT_SIZE_PT, 
        margin = margin(t = 0.5, b = 0.5, r = 1, l = 1, unit = "mm"),
        angle = 0
      ),
      strip.switch.pad.grid = unit(0, "pt"),
      
      legend.position = c(-.4, -.2),
      legend.direction = "horizontal",
      legend.text = element_text(size = BASE_TEXT_SIZE_PT),
      legend.title = element_text(size = BASE_TEXT_SIZE_PT),
    )
}

plot_selected_dots(c("MDK_NCL", "MDK_LRP1", "MIF_CD74_CXCR4", "MIF_CD74_CD44"))
ggsave(paste0(params$out,"/figures/3c_dots.png"), height = 3.4, width = 7)



```


## __Clustering based on IMC markers__

```{r heatmap: all cell types}
# Plotting the gene expression
fetahu.39.IMC.markers <- CreateSeuratObject(counts = expr_data_39markers, 
                                            meta.data = nb_metadata)

# Convert count data for the 39 markers to a matrix
mat <- fetahu.39.IMC.markers[["RNA"]]@data %>% as.matrix()

## scale the rows
mat <- t(scale(t(mat)))

cluster_anno <- fetahu.39.IMC.markers@meta.data$cellont_abbr

quantile(mat, c(0.1, 0.95))

## make the black color map to 0. the yellow map to highest and the purle map to the lowest
col_fun = circlize::colorRamp2(c(-1, 0, 1, 1.5), c("purple", "white", "lightgreen", "darkgreen"))

marker.expression <- 
  Heatmap(mat, name = "IMC markers gene expression",  
        column_split = factor(cluster_anno),
        cluster_columns = TRUE,
        show_column_dend = TRUE,
        cluster_column_slices = TRUE,
        column_title_gp = gpar(fontsize = 8),
        column_gap = unit(0.5, "mm"),
        cluster_rows = TRUE,
        show_row_dend = TRUE,
        col = col_fun,
        row_names_gp = gpar(fontsize = 4),
        column_title_rot = 90,
        top_annotation = HeatmapAnnotation(foo = anno_block(gp = gpar(fill = scales::hue_pal()(9)))),
        show_column_names = FALSE,
        use_raster = TRUE,
        raster_quality = 4)

marker.expression

pdf(file = paste0(params$out,"/figures/fetahu.exp.IMC.markers.pdf"))
marker.expression
dev.off()

marker.expression.v2 <- 
  Heatmap(mat, name = "IMC markers gene expression",  
        column_split = factor(cluster_anno),
        cluster_columns = TRUE,
        show_column_dend = TRUE,
        cluster_column_slices = TRUE,
        column_title_gp = gpar(fontsize = 8),
        column_gap = unit(0.5, "mm"),
        cluster_rows = TRUE,
        show_row_dend = TRUE,
        row_names_gp = gpar(fontsize = 4),
        column_title_rot = 90,
        top_annotation = HeatmapAnnotation(foo = anno_block(gp = gpar(fill = scales::hue_pal()(9)))),
        show_column_names = FALSE,
        use_raster = TRUE,
        raster_quality = 4)

pdf(file = paste0(params$out,"/figures/fetahu.exp.IMC.markers.v2.pdf"))
marker.expression.v2
dev.off()


```

## __Tumor cells only__

```{r heatmap: NB cells only, eval=FALSE, include=FALSE}

Idents(rna) <- rna@meta.data$orig.ident
tumor_cells <- "NB (8)"
fetahu.39.IMC.markers.NB <- subset(rna, idents = tumor_cells, invert = FALSE)
fetahu.39.IMC.markers.NB

data <- fetahu.39.IMC.markers.NB[["RNA"]]@data
filter <- rownames(data) == gene.list

data <- fetahu.39.IMC.markers.NB[["RNA"]]@data %>% as.data.frame() 

data %>% 
  filter(rownames(data) == gene.list) 

# Convert count data for the 39 markers to a matrix
mat <- fetahu.39.IMC.markers.NB[["RNA"]]@data %>% as.matrix() 


## scale the rows
mat <- t(scale(t(mat)))

quantile(mat, c(0.1, 0.95), na.rm = T)


## make the black color map to 0. the yellow map to highest and the purle map to the lowest
col_fun = circlize::colorRamp2(c(-0.5, 0,0.5, 1), c("purple", "white", "lightgreen","darkgreen"))

marker.expression <- 
  Heatmap(mat, name = "IMC markers gene expression",  
        cluster_columns = TRUE,
        show_column_dend = TRUE,
        cluster_column_slices = TRUE,
        column_title_gp = gpar(fontsize = 8),
        column_gap = unit(0.5, "mm"),
        cluster_rows = TRUE,
        show_row_dend = TRUE,
        col = col_fun,
        row_names_gp = gpar(fontsize = 4),
        column_title_rot = 90,
        top_annotation = HeatmapAnnotation(foo = anno_block(gp = gpar(fill = scales::hue_pal()(9)))),
        show_column_names = FALSE,
        use_raster = TRUE,
        raster_quality = 4,
        na_col = "black",)

marker.expression

pdf(file = paste0(params$out,"/figures/fetahu.exp.IMC.markers.NBonly.v1.pdf"))
marker.expression
dev.off()

marker.expression.v2 <- 
  Heatmap(mat, name = "IMC markers gene expression",  
        cluster_columns = TRUE,
        show_column_dend = TRUE,
        cluster_column_slices = TRUE,
        column_title_gp = gpar(fontsize = 8),
        column_gap = unit(0.5, "mm"),
        cluster_rows = TRUE,
        show_row_dend = TRUE,
        row_names_gp = gpar(fontsize = 4),
        column_title_rot = 90,
        top_annotation = HeatmapAnnotation(foo = anno_block(gp = gpar(fill = scales::hue_pal()(9)))),
        show_column_names = FALSE,
        use_raster = TRUE,
        raster_quality = 4)

pdf(file = paste0(params$out,"/figures/fetahu.exp.IMC.markers.NBonly.v2.pdf"))
marker.expression.v2
dev.off()


```

```{r sub-cluster the tumor cluster}
filter <- nb_metadata$cellont_abbr == "NB"
tumor.metadata <- nb_metadata[filter,]


fetahu <- CreateSeuratObject(counts = expr_data, metadata = nb_metadata)
fetahu@meta.data$orig.ident <- nb_metadata$cellont_abbr
Idents(fetahu) <- fetahu@meta.data$orig.ident
tumor_cells <- c("NB (8)")
fetahu.NB <- subset(fetahu, 
                    idents = tumor_cells, 
                    invert = FALSE)

fetahu.NB
fetahu.NB <- AddMetaData(fetahu.NB,metadata = tumor.metadata)

fetahu.NB[["percent.mt"]] <- PercentageFeatureSet(object = fetahu.NB, 
                                                  pattern = "^MT-")

DefaultAssay(fetahu.NB) <- "RNA"
fetahu.NB <- fetahu.NB %>%
    NormalizeData() %>%
    FindVariableFeatures(selection.method = "vst", nfeatures = 2000) %>% 
    ScaleData() %>%
    SCTransform(vars.to.regress = c("percent.mt"))

# Calculate PCs using variable features determined by SCTransform
fetahu.NB <- RunPCA(fetahu.NB, assay = "SCT", seed.use = 28, npcs = 50)
PCAPlot(object = fetahu.NB)


top2000 <- head(VariableFeatures(fetahu.NB), 4000)
fetahu.NB <- fetahu.NB[top2000]

fetahu.NB <- RunPCA(fetahu.NB, assay = "SCT", npcs = 10)
fetahu.NB <- FindNeighbors(fetahu.NB, reduction = "pca", 
                           dims = 1:5,
                           k.param = 50)

# tested k nearest neighbors: 10, 50, 80
# tested number fo dimensions: 5, 10, 20, 50

# define neighbours based on gene expression rather than dimensionlity reduction
# Compute an SNN on the gene expression level
#fetahu.NB <- FindNeighbors(fetahu.NB, 
#                           features = VariableFeatures(object = fetahu.NB))

fetahu.NB <- FindClusters(fetahu.NB, resolution = 0.2,
                          random.seed = 28,
                          group.singletons = FALSE) # 4 = Leiden, default = Louvain 
fetahu.NB <- RunUMAP(fetahu.NB, 
                     reduction = "pca", 
                     dims = 1:5,
                     n.epochs = 500,
                     min.dist = 0.1,
                     seed.use = 28)

# Tested minimum distance 0.5, 0.1, 0.01, 0.001

# Visualization
p1 <- DimPlot(fetahu.NB, reduction = "umap", 
              group.by = "sample")
p2 <- DimPlot(fetahu.NB, reduction = "umap",
              group.by = "group")

p1 + p2

p3 <- DimPlot(fetahu.NB, reduction = "umap", 
              label = TRUE, repel = TRUE)
p3

saveRDS(fetahu.NB, 
        file = paste0(params$out,"Rds/fetahu2023.scRNA.tumor.cluster.Rds"))

```

# __Sub-setting the tumor cluster markers__

```{r}
fetahu.NB <- readRDS(file = 
                       paste0(params$out,"Rds/fetahu.scRNA.tumor.cluster.Rds"))

# find cluster markers with Seurat
fetahu.NB.markers <- FindAllMarkers(fetahu.NB, 
                                      min.pct  = 0.1, 
                                      logfc.threshold = 0.25,
                                      min.cells.group = 50,
                                      random.seed = 28)
fetahu.NB.markers %>%
    group_by(cluster) %>%
    slice_max(n = 10, order_by = avg_log2FC)

fetahu.NB.markers %>%
    group_by(cluster) %>%
    top_n(n = 10, wt = avg_log2FC) -> top10
DoHeatmap(fetahu.NB, features = fetahu.NB.markers$gene) + NoLegend()

library(DElegate)

# find cluster markers with DElegate 
de_results <- DElegate::findDE(object = fetahu.NB,
                               group_column = "seurat_clusters")
marker_results <- DElegate::FindAllMarkers2(object = fetahu.NB,
                                            group_column = "seurat_clusters")
dplyr::filter(marker_results, feature_rank < 6)

DoHeatmap(marker_results)


fetahu.NB.markers %>%
    group_by(cluster) %>%
    top_n(n = 10, wt = avg_log2FC) -> top10


```

```{r}

NB.marker.genes <- c("NCAM1", "B4GALNT1", "L1CAM", "MYCN", "GATA2", 
                     "DBH", "CHGB", "CHGA", "SYP", "TH", "PHOX2A", "PHOX2B")

DoHeatmap(fetahu.NB, features = NB.marker.genes,assay = "RNA")

top100 <- head(VariableFeatures(fetahu.NB), 100)
top100
DoHeatmap(fetahu.NB, features = top100) + NoLegend()


var.genes <- as.data.frame(x =  VariableFeatures(fetahu.NB))

colnames(var.genes) <- "var_genes"

overlap = var.genes$var_genes[var.genes$var_genes %in% gene.list]
overlap
# I find 6 genes from the IMC markers to be among the 2000 most variable genes
# in the tumor cluster

```




# __Label transfer__

Transfer cell type annotations from Lazic et al. 2024 to Fetahu et al. 2023


```{r label transfer}
library(BiocParallel)
library(SingleR)
library(tidyr)

metadata <- readRDS("/home/data/MapMet/fetahu.scRNA.2023/data/metadata.rds")
scrna_data <- readRDS("/home/data/MapMet/fetahu.scRNA.2023/data/rna_decontaminated.rds")

rna <- as.Seurat(scrna_data, data=NULL)
rna <- RenameAssays(object = rna, originalexp = 'RNA')


rna@active.ident <- setNames(pull(metadata[match(metadata$cell,colnames(rna)),"cellont_cluster"], "cellont_cluster"),pull(metadata[match(metadata$cell,colnames(rna)), "cell"], "cell"))
DefaultAssay(rna) <- "RNA"


rna_mat <- GetAssayData(rna, assay = "RNA", slot = "data")

rna_clusters <- rna@active.ident
common_genes_rna <- rownames(rna_mat) %in% gene.list

rna_mat <- GetAssayData(rna, assay = "RNA", slot = "data")[common_genes_rna,]

protein_mat <- GetAssayData(imc_data, assay = "originalexp", slot = "counts")
common_genes_protein <- rownames(protein_mat) %in% rownames(rna_mat)
protein_mat <- GetAssayData(imc_data, assay = "originalexp", slot = "counts")[common_genes_protein,]
imc_clusters <- imc_data@meta.data$celltype
predicted_labels <- SingleR(test = rna_mat, # Transfer labels to RNA data
                            ref = protein_mat, # Original cell types from IMC data
                            labels = imc_clusters, # Cell types defined by IMC data
                            BPPARAM = MulticoreParam(workers=10, RNGseed=20231030)) 

saveRDS(predicted_labels,
        file = paste0(params$out,"Rds/20240126_predicted_labels_IMCtoRNA.rds"))

################################################################################

# Take only the BM samples (because some cell types were only present in the PT)
imc_data@meta.data$tissue
imc_BM <- subset(imc_data, tissue == "BM")
imc_BM 

protein_mat <- GetAssayData(imc_BM, assay = "originalexp", slot = "counts")
common_genes_protein <- rownames(protein_mat) %in% rownames(rna_mat)
protein_mat <- GetAssayData(imc_BM, assay = "originalexp", slot = "counts")[common_genes_protein,]
imc_clusters <- imc_BM@meta.data$celltype
predicted_labels <- SingleR(test = rna_mat, # Transfer labels to RNA data
                            ref = protein_mat, # Original cell types from IMC data
                            labels = imc_clusters, # Cell types defined by IMC data
                            BPPARAM = MulticoreParam(workers=10, RNGseed=20231030)) 

saveRDS(predicted_labels,
        file = paste0(params$out,"Rds/20240126_predicted_labels_IMC-BMtoRNA.rds"))

```


```{r match colors, message=FALSE, fig.height=17}
col_clusters <- c(#"GD2+MPO+"=rgb(255,255,162, maxColorValue = 255),
                #"DTC"=rgb(255,221,57, maxColorValue = 255),
                #"DTC"=rgb(246,246,0, maxColorValue = 255),
                "other"=rgb(255,255,209, maxColorValue = 255),
                "LIN- Ki67+"=rgb(254,217,166, maxColorValue = 255),
                "HSPC"=rgb(253,180,98, maxColorValue = 255),
                #"GZMB+ S100B- DC/NK"=rgb(255,137,0, maxColorValue = 255),
                "GZMB+ S100B- DC/NK"=rgb(28,146,255, maxColorValue = 255),
                "B cell"=rgb(168,128,250, maxColorValue = 255),
                "GZMB+ S100B+ DC/NK"=rgb(0,255,255, maxColorValue = 255),
                #"HLA-DR+ CXCR4+"=rgb(152,78,163, maxColorValue = 255),
                #"B cell"=rgb(82,75,181, maxColorValue = 255),
                "CD8+ naive T cell"=rgb(188,128,189, maxColorValue = 255),
                "DN GZMB+ T cell"=rgb(190,174,212, maxColorValue = 255),
                "CD8+ S100B+ T cell"=rgb(253,218,236, maxColorValue = 255),
                "CD8+ GZMB+ PD1lo T cell"=rgb(251,154,153, maxColorValue = 255),
                "CD4+ naive T cell"=rgb(242,3,137, maxColorValue = 255),
                "DN T cell"=rgb(251,128,114, maxColorValue = 255),
                "CD4+ PD1+ T cell"=rgb(254,142,175, maxColorValue = 255),
                "CD8+ PDL1lo T cell"=rgb(240,0,240, maxColorValue = 255),
                "CD4+ S100B+ T cell"=rgb(227,26,28, maxColorValue = 255),
                "dense T cell region"=rgb(145,2,144, maxColorValue = 255),
                "CD24+ marker-lo TC"=rgb(230,245,201, maxColorValue = 255),
                "early SYM-like TC"=rgb(204,235,197, maxColorValue = 255),
                "neural progenitor"=rgb(238,118,0, maxColorValue = 255),
                "bridge-like TC"=rgb(198,255,72, maxColorValue = 255),
                "GATA3hi TC"=rgb(65,208,57, maxColorValue = 255),
                "CD24- marker-lo TC"=rgb(60,119,97, maxColorValue = 255),
                "CD14+ PDL1- MO"=rgb(104,180,238, maxColorValue = 255),
                "monoblast-like"=rgb(128,177,211, maxColorValue = 255),
                "CD14+ PDL1+ MO"=rgb(51,180,180, maxColorValue = 255),
                "CHGAhi TC"=rgb(102,166,30, maxColorValue = 255),
                "Ki67hi TC"=rgb(149,144,89, maxColorValue = 255),
                "CXCR4hi TC"=rgb(210,205,126, maxColorValue = 255),
                "CD44+ mes-like TC"=rgb(27,158,119, maxColorValue = 255),
                "GD2lo TC"=rgb(102,194,165, maxColorValue = 255),
                "CD14- PDL1- MO/DC"=rgb(39,130,187, maxColorValue = 255),
                "CD14- PDL1+ MO/DC"=rgb(166,206,227, maxColorValue = 255),
                "pDC"=rgb(212,241,240, maxColorValue = 255),
                "band cell"=rgb(169,189,210, maxColorValue = 255),
                "myelocyte"=rgb(217,217,217, maxColorValue = 255),          
                #"granulocyte"=rgb(251,241,232, maxColorValue = 255),
                "myeloblast"=rgb(2,0,138, maxColorValue = 255),
                "neutrophil"=rgb(113,113,113, maxColorValue = 255),
                "fibroblast/endothel"=rgb(212,89,67, maxColorValue = 255),
                "schwann cell"=rgb(156,76,24, maxColorValue = 255),
                "MSC"=rgb(174,129,37, maxColorValue = 255),
                "HPC"=rgb(212,143,75, maxColorValue = 255),
                "Ki67+ CXCR4+ cell"=rgb(254,215,0, maxColorValue = 255))
 
pie(rep(1, length(col_clusters)), col=col_clusters, labels=names(col_clusters))
 
col_clusters

```

using BM and PT samples (IMC data) for label transfer
```{r heatmap of transferred labels}
library(viridis)
library(RColorBrewer)
set.seed(20230517)

predicted_labels <- readRDS(file = paste0(params$out,"Rds/20240126_predicted_labels_IMCtoRNA.rds"))

plotScoreHeatmap(predicted_labels[sample(seq_len(nrow(predicted_labels)), 2000),], 
                 annotation_legend=T, treeheight_row=0) 

pdf(paste0(params$out,"figures/labeltransfer_score_heatmap.pdf"))
plotScoreHeatmap(predicted_labels[sample(seq_len(nrow(predicted_labels)), 2000),], 
                 annotation_legend=T, treeheight_row=0) 
dev.off()

true_label <- rna@active.ident
predicted_label <- predicted_labels$labels

n <- length(unique(predicted_label))

table(predicted_label)



### downsample

rna.small <- subset(rna, downsample = 500)
rna_mat.small <- GetAssayData(rna.small, assay = "RNA", slot = "data")[common_genes_rna,]

rna_mat_scaled <-  t(scale(t(rna_mat.small)))

qual_col_pals = brewer.pal.info[brewer.pal.info$category == 'qual',]
col_vector_74 = unlist(mapply(brewer.pal, qual_col_pals$maxcolors, rownames(qual_col_pals)))
col_vector_433 = colors()[grep('gr(a|e)y', grDevices::colors(), invert = T)]

col_predicted_labels = setNames(sample(col_vector_433, length(unique(predicted_label))), unique(predicted_label))
col_true_labels <- metadata$cellont_abbr

column_ha = HeatmapAnnotation(predicted_label = predicted_label[rna_mat_scaled],
                              true_cluster = true_label[rna_mat_scaled],
                              col=list(predicted_label=col_predicted_labels, 
                                       true_label=col_true_labels
                                       ), 
                              show_legend=T)


cluster_anno <- rna.small@meta.data$predicted_celltype

quantile(rna_mat_scaled, c(0.1, 0.95))

## make the white color map to 0. the green map to highest and the purle map to the lowest
col_fun = circlize::colorRamp2(c(-0.5, 0, 0.5, 1), 
                               c("purple", "white", "lightgreen", "darkgreen"))

rna <- SetIdent(rna, value = "predicted_celltype")

x <- Heatmap(rna_mat_scaled, name = "IMC markers gene expression",  
             column_split = factor(cluster_anno),
             cluster_columns = TRUE,
             show_column_dend = TRUE,
             cluster_column_slices = TRUE,
             column_title_gp = gpar(fontsize = 8),
             column_gap = unit(0.5, "mm"),
             cluster_rows = TRUE,
             show_row_dend = TRUE,
             col = col_fun,
             row_names_gp = gpar(fontsize = 4),
             column_title_rot = 90,
             top_annotation = HeatmapAnnotation(foo = anno_block(gp = gpar(fill = scales::hue_pal()(9)))),
             show_column_names = FALSE,
             use_raster = TRUE,
             raster_quality = 4)

pdf(paste0(params$out,"figures/labeltransfer_expression_heatmap.pdf"))
x
dev.off()

pdf(paste0(params$out,"figures/labeltransfer_IMC_marker_expression_dotplot.pdf"))
DotPlot(rna, features = gene.list) + RotatedAxis()
dev.off()


tumor_markers <- c("ELAVL4", "CXCR4", "MME", "CD74", "HLA-A", "CD274", "GATA3", 
                   "VIM", "MKI67", "CD44", "NCAM1", "ST8SIA1", "PRPH", "LUM", 
                   "CHGA", "SOX10", "S100B")

pdf(paste0(params$out,"figures/labeltransfer_tumor_marker_expression_dotplot.pdf"))
DotPlot(rna, features = tumor_markers) + RotatedAxis()
dev.off()

DoHeatmap(subset(rna, downsample = 500), 
          features = gene.list, 
          size = 3, 
          slot = "counts")


# Plot in the UMAP published and predicted cell types (based on IMC protein data)

library(scater)
library(patchwork)
library(cowplot)
library(viridis)

data <- data.frame(IMC_markers=predicted_label, 
                   original_clusters=rna@active.ident)

data$freq <- 1
data_wide <- data %>% pivot_wider(names_from = c("original_clusters"), 
                                  values_from = "freq", 
                                  values_fn = sum, 
                                  values_fill = 0)

data_long <- data_wide %>% pivot_longer(cols = !IMC_markers, 
                                        names_to = "original_clusters", 
                                        values_to = "count") 


IMC_cell_types <- data_wide$IMC_markers
data <- data_wide
data <- data[,-1]
rownames(data) <- IMC_cell_types
data <- (prop.table(as.matrix(data), margin=2))
original_clusters <- names(unlist(apply(data, 2, which.max)))
predicted_clust_1 <- data_wide[unname(unlist(apply(data, 2, which.max))), "IMC_markers"]
proportion_1 <- unname(unlist(apply(data, 2, max)))
predicted_clust_2 <- unname(apply(data, 2, function(x){rownames(data)[order(x, decreasing=T)][2]}))
proportion_2 <- unname(apply(data, 2, function(x){sort(x, decreasing=T)[2]}))

assignments <- as.data.frame(cbind("original_clusters" = original_clusters, 
                                   "predicted_cluster_1" = predicted_clust_1, 
                                   "proportion_1" = proportion_1, 
                                   "predicted_cluster_2" = predicted_clust_2, 
                                   "proportion_2" = proportion_2))

assignments

rna$original_celltype <- rna@active.ident
rna$predicted_celltype <- predicted_label

bar_plot <- ggplot(data_long, aes(fill=IMC_markers, y=count, x=original_clusters)) + 
  ggtitle("") +
  geom_bar(position="fill", stat="identity", show.legend = T) + 
  scale_fill_manual(values = col_clusters) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) 

pdf(paste0(params$out,"figures/barplot_predicted_celltypes_proportion.pdf"),
    height = 10,width = 20)
bar_plot
dev.off()


## UMAP colored by cell type and expression - dittoDimPlot
p1 <- DimPlot(rna, 
             reduction = "UMAP", 
             pt.size = 0.3, 
             seed = 28, 
             label =T, 
             group.by = "predicted_celltype")



pdf(file=paste0(params$out,"figures/umap.pdf"),height = 10, width = 20)
p1
dev.off()


table(rna$predicted_celltype)
table(rna$orig.ident)

rna$original_celltype

compare_cell_types <- cbind(rna$orig.ident,rna$predicted_celltype)


write.table(compare_cell_types, 
             file = paste0(params$out,"tables/fetahu.scRNA.original.vs.predicted.celltypes.csv"),
             sep = "\t", quote = F)

```

using only BM samples (IMC data) for label transfer
```{r heatmap of transferred labels 2}
library(viridis)
library(RColorBrewer)
set.seed(20230517)

predicted_labels <- 
  readRDS(file = paste0(params$out,"Rds/20240126_predicted_labels_IMC-BMtoRNA.rds"))

plotScoreHeatmap(predicted_labels[sample(seq_len(nrow(predicted_labels)), 2000),], 
                 annotation_legend=T, treeheight_row=0) 

pdf(paste0(params$out,"figures/labeltransfer_IMC-BM_score_heatmap.pdf"))
plotScoreHeatmap(predicted_labels[sample(seq_len(nrow(predicted_labels)), 2000),], 
                 annotation_legend=T, treeheight_row=0) 
dev.off()

rna@meta.data$original_celltype <- rna@active.ident
true_label <- rna@meta.data$original_celltype

predicted_label <- predicted_labels$labels
rna@meta.data$predicted_celltype <- predicted_label


n <- length(unique(predicted_label))

table(predicted_label)

rna_mat_scaled <-  t(scale(t(rna_mat)))

qual_col_pals = brewer.pal.info[brewer.pal.info$category == 'qual',]
col_vector_74 = unlist(mapply(brewer.pal, qual_col_pals$maxcolors, rownames(qual_col_pals)))
col_vector_433 = colors()[grep('gr(a|e)y', grDevices::colors(), invert = T)]

col_predicted_labels = setNames(sample(col_vector_433, length(unique(predicted_label))), unique(predicted_label))
col_true_labels <- metadata$cellont_abbr

column_ha = HeatmapAnnotation(predicted_label = predicted_label[rna_mat_scaled],
                              true_cluster = true_label[rna_mat_scaled],
                              col=list(predicted_label=col_predicted_labels, 
                                       true_label=col_true_labels
                                       ), 
                              show_legend=T)



quantile(rna_mat_scaled, c(0.1, 0.95))

## make the white color map to 0. the green map to highest and the purle map to the lowest
col_fun = circlize::colorRamp2(c(-0.5, 0, 0.5, 1), 
                               c("purple", "white", "lightgreen", "darkgreen"))

rna <- SetIdent(rna, value = "predicted_celltype")

x <- Heatmap(rna_mat_scaled, name = "IMC markers gene expression",  
             #column_split = factor(predicted_label),
             cluster_columns = TRUE,
             show_column_dend = TRUE,
             cluster_column_slices = TRUE,
             column_title_gp = gpar(fontsize = 8),
             column_gap = unit(0.5, "mm"),
             cluster_rows = TRUE,
             show_row_dend = TRUE,
             col = col_fun,
             row_names_gp = gpar(fontsize = 4),
             column_title_rot = 90,
             top_annotation = HeatmapAnnotation(foo = anno_block(gp = gpar(fill = scales::hue_pal()(9)))),
             show_column_names = FALSE,
             use_raster = TRUE,
             raster_quality = 4)

pdf(paste0(params$out,"figures/labeltransfer_IMC-BM_expression_heatmap.pdf"))
x
dev.off()

DotPlot(rna, features = gene.list) + RotatedAxis()

pdf(paste0(params$out,"figures/labeltransfer_IMC-BM_marker_expression_dotplot.pdf"))
DotPlot(rna, features = gene.list) + RotatedAxis()
dev.off()


tumor_markers <- c("ELAVL4", "CXCR4", "MME", "CD74", "HLA-A", "CD274", "GATA3", 
                   "VIM", "MKI67", "CD44", "NCAM1", "ST8SIA1", "PRPH", "LUM", 
                   "CHGA", "SOX10", "S100B")

pdf(paste0(params$out,"figures/labeltransfer_IMC-BM_tumor_marker_expression_dotplot.pdf"))
DotPlot(rna, features = tumor_markers) + RotatedAxis()
dev.off()

DoHeatmap(subset(rna, downsample = 500), 
          features = gene.list, 
          size = 3, 
          slot = "counts")


# Plot in the UMAP published and predicted cell types (based on IMC protein data)

library(scater)
library(patchwork)
library(cowplot)
library(viridis)

data <- data.frame(IMC_cell_types=predicted_label, 
                   original_clusters=rna@meta.data$original_celltype)

data$freq <- 1
data_wide <- data %>% pivot_wider(names_from = c("original_clusters"), 
                                  values_from = "freq", 
                                  values_fn = sum, 
                                  values_fill = 0)

data_long <- data_wide %>% pivot_longer(cols = !IMC_cell_types, 
                                        names_to = "original_clusters", 
                                        values_to = "count") 


IMC_cell_types <- data_wide$IMC_cell_types
data <- data_wide
data <- data[,-1]
rownames(data) <- IMC_cell_types
data <- (prop.table(as.matrix(data), margin=2))
original_clusters <- names(unlist(apply(data, 2, which.max)))
predicted_clust_1 <- data_wide[unname(unlist(apply(data, 2, which.max))), "IMC_cell_types"]
proportion_1 <- unname(unlist(apply(data, 2, max)))
predicted_clust_2 <- unname(apply(data, 2, function(x){rownames(data)[order(x, decreasing=T)][2]}))
proportion_2 <- unname(apply(data, 2, function(x){sort(x, decreasing=T)[2]}))

assignments <- as.data.frame(cbind("original_clusters" = original_clusters, 
                                   "predicted_cluster_1" = predicted_clust_1, 
                                   "proportion_1" = proportion_1, 
                                   "predicted_cluster_2" = predicted_clust_2, 
                                   "proportion_2" = proportion_2))

assignments


#rna <- SetIdent(rna, value = "predicted_celltype")


bar_plot <- ggplot(data_long, aes(fill=IMC_cell_types, y=count, x=original_clusters)) + 
  ggtitle("") +
  geom_bar(position="fill", stat="identity", show.legend = T) + 
  scale_fill_manual(values = col_clusters) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) 

bar_plot

pdf(paste0(params$out,"figures/barplot_predicted_celltypes_IMC-BM_proportion.pdf"),
    height = 10,width = 20)
bar_plot
dev.off()


## UMAP colored by cell type and expression - dittoDimPlot
p1 <- DimPlot(rna, 
             reduction = "UMAP", 
             pt.size = 0.3, 
             seed = 28, 
             label =T, 
             group.by = "predicted_celltype")



pdf(file=paste0(params$out,"figures/umap_IMC-BM_labelTransfer.pdf"),height = 10, width = 20)
p1
dev.off()


table(rna$predicted_celltype)
table(rna$orig.ident)

rna$original_celltype

compare_cell_types <- c(rna$orig.ident,rna$predicted_celltype)

```


# __CCC analysis__

To infer the cell state-specific communications, CellChat v2 identifies over-expressed 
ligands or receptors in one cell group and then identifies over-expressed 
ligand-receptor interactions if either ligand or receptor are over-expressed.

## Sub-clusters (from Lazic et al., 2024)

```{r CellChat: sub-clusters}

Idents(rna) <- rna@meta.data$predicted_celltype
#nb_metadata$predicted_celltype <- Idents(rna)
rna_mat <- GetAssayData(rna, assay = "RNA", slot = "data")

# New analysis with CellChat v2.1.1
cellchat <- createCellChat(
  rna_mat,
  meta = rna@meta.data,
  group.by = "predicted_celltype",
)
cellchat@DB <- CellChatDB.human

# only keep genes in CellChatDB
cellchat <- subsetData(cellchat)

# find overexpressed ligands/receptors
cellchat <- identifyOverExpressedGenes(cellchat)

# find overexpressed interactions
cellchat <- identifyOverExpressedInteractions(cellchat)

# calculate communication probability between cell groups
cellchat <- computeCommunProb(cellchat, seed.use = 1)

# filter communications occurring with less than 10 cells in a cell group
cellchat <- filterCommunication(cellchat)

# calculate communication probability on the signaling pathway level
cellchat <- computeCommunProbPathway(cellchat)

# calculate aggregated communication network
cellchat <- aggregateNet(cellchat)

# determine network centrality pathways
cellchat <- netAnalysis_computeCentrality(cellchat)

# define which clusters to compare
all_clusters <- c("other",
                  "LIN- Ki67+",
                  "HSPC",
                  "GZMB+ S100B- DC/NK",
                  "B cell", 
                  "GZMB+ S100B+ DC/NK",
                  "CD8+ naive T cell",
                  "DN GZMB+ T cell",
                  "CD8+ S100B+ T cell",
                  "CD8+ GZMB+ PD1lo T cell",
                  "CD4+ naive T cell",
                  "DN T cell",
                  "CD4+ PD1+ T cell",
                  "CD8+ PDL1lo T cell", 
                  "CD4+ S100B+ T cell", 
                  "dense T cell region", 
                  "CD24+ marker-lo TC", 
                  "early SYM-like TC",
                  "neural progenitor",
                  "bridge-like TC",
                  "GATA3hi TC","
                  CD24- marker-lo TC",
                  "CD14+ PDL1- MO",
                  "monoblast-like",
                  "CD14+ PDL1+ MO", 
                  "CHGAhi TC", 
                  "Ki67hi TC", 
                  "CXCR4hi TC", 
                  "CD44+ mes-like TC",
                  "GD2lo TC",
                  "CD14- PDL1- MO/DC", 
                  "CD14- PDL1+ MO/DC", 
                  "pDC",
                  "band cell", 
                  "myelocyte", 
                  "myeloblast",
                  "neutrophil",
                  "fibroblast/endothel",
                  "schwann cell",
                  "MSC", 
                  "HPC",
                  "Ki67+ CXCR4+ cell")

treat_resist <- c("early SYM-like TC","CD44+ mes-like TC",
                  "CD14+ PDL1+ MO","CD4+ PD1+ T cell")

# single LR dotplot data
# only cells associated with treatment resistance and progression
df_net_int <-
  cellchat %>% 
  subsetCommunication(thresh = NA) %>% 
  as_tibble() %>% 
  group_by(interaction_name_2) %>% 
  mutate(prob.norm = prob / max(prob)) %>% 
  ungroup() %>% 
  mutate(
    source = factor(source, treat_resist),
    target = factor(target, treat_resist),
  )


df_net_int

# all cells / clusters
df_net_all <-
  cellchat %>% 
  subsetCommunication(thresh = NA) %>% 
  as_tibble() %>% 
  group_by(interaction_name_2) %>% 
  mutate(prob.norm = prob / max(prob)) %>% 
  ungroup() %>% 
  mutate(
    source = factor(source, all_clusters),
    target = factor(target, all_clusters),
  )


df_net_all

cellchat %>% 
  saveRDS(paste0(params$out,"Rds/fetahu2023.cellchat.predictedCellTypes.fromIMC.rds"))
df_net_all %>% 
  saveRDS(paste0(params$out,"Rds/fetahu2023.signaling.predictedCellTypes.fromIMC.rds"))
df_net_int %>% 
  saveRDS(paste0(params$out,"Rds/fetahu2023.signaling.predictedCellTypes.fromIMC.TreatResistCellsOnly.rds"))

```


```{r CellChat figures: sub-clusters}
library(latex2exp)
source("/home/data/MapMet/fetahu.scRNA.2023/scripts/csbg-neuroblastoma-d976641/styling.R")

#cellchat <- readRDS(paste0(params$out,"Rds/fetahu2023.signaling.predictedCellTypes.fromIMC.TreatResistCellsOnly.rds"))
#sig_data <- readRDS(paste0(params$out,"Rds/fetahu2023.cellchat.predictedCellTypes.fromIMC.rds"))

summary_treat_resist <- drop_na(df_net_int)
write.table(summary_treat_resist, 
            file = paste0(params$out,"figures/CCC_treatResist.csv"),
            sep = "\t", quote = FALSE)

mat <- cellchat@net$count
total_target <- colSums(mat)
total_source <- rowSums(mat)
bar_ylim <- c(0, max(total_source, total_target))

## plot: Number of interactions between all cell types
BASE_TEXT_SIZE_PT = 20

int_plot <- Heatmap(mat,
    col = RColorBrewer::brewer.pal(9, "YlOrBr"),
    name = "number of\ninteractions",
    
    cluster_rows = FALSE, 
    row_names_gp = gpar(fontsize = BASE_TEXT_SIZE_PT), 
    row_names_side = "left",
    row_title = "source (ligand)", 
    row_title_gp = gpar(fontsize = BASE_TEXT_SIZE_PT), 
    
    cluster_columns = FALSE, 
    column_names_gp = gpar(fontsize = BASE_TEXT_SIZE_PT), 
    column_title = "target (receptor)",
    column_title_side = "bottom",
    column_title_gp = gpar(fontsize = BASE_TEXT_SIZE_PT))

pdf(paste0(params$out,"figures/number_of_interactions_all_celltypes.pdf"), 
    height = 20, width = 20)
int_plot
dev.off()


## plot: one cell type to all

# 1. early SYM-like TC
treat_resist[1]

plot_contribution_celltype <- function(cell_type = treat_resist[1], signif = 0.05) {
   df_net_int %>%
    filter(
      pathway_name %in% cellchat@netP$pathways, 
      source == {{cell_type}}, 
      target != {{cell_type}},
      pval <= signif
    ) %>%
    group_by(interaction = interaction_name_2) %>%
    summarise(
      prob.norm = sum(prob.norm),
      count = n(),
      pathway_name = first(pathway_name)
    ) %>%
    ungroup() %>%
    mutate(prob.avg = prob.norm / count) %>% 
    mutate(interaction = fct_reorder(interaction, prob.norm)) %>% 
    ggplot(aes(interaction, prob.norm)) +
    geom_col(fill = "#ffb03e", width = 0.85) +
    geom_hline(yintercept = 0, size = BASE_LINE_SIZE) +
    geom_text(
      aes(y = 2.68, label = pathway_name),
      size = BASE_TEXT_SIZE_MM,
      fontface = "italic"
    ) +
    xlab("ligand-receptor pair") +
    scale_y_continuous(
      "total outgoing communication score from treatment resistant cells",
      expand = expansion(0)
    ) +
    coord_flip() +
    theme_nb(grid = FALSE) +
    theme(
      axis.title.x = element_text(hjust = 1),
      axis.ticks.y = element_blank(),
      panel.border = element_blank(),
      panel.grid.major.x = element_line(
        color = "grey92",
        size = BASE_LINE_SIZE
      )
    )
}

plot_contribution_celltype()
ggsave(paste0(params$out,"figures/early_SYM-like_TC_outgoing_comm_score_allCellTypes.png"))

plot_contribution_celltype <- function(cell_type = treat_resist[1], signif = 0.05) {
   df_net_all %>%
    filter(
      pathway_name %in% cellchat@netP$pathways, 
      source == {{cell_type}}, 
      target != {{cell_type}},
      pval <= signif
    ) %>%
    group_by(interaction = interaction_name_2) %>%
    summarise(
      prob.norm = sum(prob.norm),
      count = n(),
      pathway_name = first(pathway_name)
    ) %>%
    ungroup() %>%
    mutate(prob.avg = prob.norm / count) %>% 
    mutate(interaction = fct_reorder(interaction, prob.norm)) %>% 
    ggplot(aes(interaction, prob.norm)) +
    geom_col(fill = "#ffb03e", width = 0.85) +
    geom_hline(yintercept = 0, size = BASE_LINE_SIZE) +
    geom_text(
      aes(y = 2.68, label = pathway_name),
      size = BASE_TEXT_SIZE_MM,
      fontface = "italic"
    ) +
    xlab("ligand-receptor pair") +
    scale_y_continuous(
      "total outgoing communication score from treatment resistant cells",
      expand = expansion(0)
    ) +
    coord_flip() +
    theme_nb(grid = FALSE) +
    theme(
      axis.title.x = element_text(hjust = 1),
      axis.ticks.y = element_blank(),
      panel.border = element_blank(),
      panel.grid.major.x = element_line(
        color = "grey92",
        size = BASE_LINE_SIZE
      )
    )
}

plot_contribution_celltype()
ggsave(paste0(params$out,"figures/early_SYM-like_TC_outgoing_comm_score_treatResistCells.png"))

# 2. CD44+ mes-like TC
treat_resist[2]

plot_contribution_celltype <- function(cell_type = treat_resist[2], signif = 0.05) {
   df_net_int %>%
    filter(
      pathway_name %in% cellchat@netP$pathways, 
      source == {{cell_type}}, 
      target != {{cell_type}},
      pval <= signif
    ) %>%
    group_by(interaction = interaction_name_2) %>%
    summarise(
      prob.norm = sum(prob.norm),
      count = n(),
      pathway_name = first(pathway_name)
    ) %>%
    ungroup() %>%
    mutate(prob.avg = prob.norm / count) %>% 
    mutate(interaction = fct_reorder(interaction, prob.norm)) %>% 
    ggplot(aes(interaction, prob.norm)) +
    geom_col(fill = "#ffb03e", width = 0.85) +
    geom_hline(yintercept = 0, size = BASE_LINE_SIZE) +
    geom_text(
      aes(y = 2.68, label = pathway_name),
      size = BASE_TEXT_SIZE_MM,
      fontface = "italic"
    ) +
    xlab("ligand-receptor pair") +
    scale_y_continuous(
      "total outgoing communication score from treatment resistant cells",
      expand = expansion(0)
    ) +
    coord_flip() +
    theme_nb(grid = FALSE) +
    theme(
      axis.title.x = element_text(hjust = 1),
      axis.ticks.y = element_blank(),
      panel.border = element_blank(),
      panel.grid.major.x = element_line(
        color = "grey92",
        size = BASE_LINE_SIZE
      )
    )
}

plot_contribution_celltype()
ggsave(paste0(params$out,"figures/CD44+_mes-like_TC_outgoing_comm_score_allCellTypes.png"))

plot_contribution_celltype <- function(cell_type = treat_resist[2], signif = 0.05) {
   df_net_all %>%
    filter(
      pathway_name %in% cellchat@netP$pathways, 
      source == {{cell_type}}, 
      target != {{cell_type}},
      pval <= signif
    ) %>%
    group_by(interaction = interaction_name_2) %>%
    summarise(
      prob.norm = sum(prob.norm),
      count = n(),
      pathway_name = first(pathway_name)
    ) %>%
    ungroup() %>%
    mutate(prob.avg = prob.norm / count) %>% 
    mutate(interaction = fct_reorder(interaction, prob.norm)) %>% 
    ggplot(aes(interaction, prob.norm)) +
    geom_col(fill = "#ffb03e", width = 0.85) +
    geom_hline(yintercept = 0, size = BASE_LINE_SIZE) +
    geom_text(
      aes(y = 2.68, label = pathway_name),
      size = BASE_TEXT_SIZE_MM,
      fontface = "italic"
    ) +
    xlab("ligand-receptor pair") +
    scale_y_continuous(
      "total outgoing communication score from treatment resistant cells",
      expand = expansion(0)
    ) +
    coord_flip() +
    theme_nb(grid = FALSE) +
    theme(
      axis.title.x = element_text(hjust = 1),
      axis.ticks.y = element_blank(),
      panel.border = element_blank(),
      panel.grid.major.x = element_line(
        color = "grey92",
        size = BASE_LINE_SIZE
      )
    )
}

plot_contribution_celltype()
ggsave(paste0(params$out,"figures/CD44+_mes-like_TC_outgoing_comm_score_treatResistCells.png"))

# 3. CD14+ PDL1+ MO
treat_resist[3]

plot_contribution_celltype <- function(cell_type = treat_resist[3], signif = 0.05) {
   df_net_int %>%
    filter(
      pathway_name %in% cellchat@netP$pathways, 
      source == {{cell_type}}, 
      target != {{cell_type}},
      pval <= signif
    ) %>%
    group_by(interaction = interaction_name_2) %>%
    summarise(
      prob.norm = sum(prob.norm),
      count = n(),
      pathway_name = first(pathway_name)
    ) %>%
    ungroup() %>%
    mutate(prob.avg = prob.norm / count) %>% 
    mutate(interaction = fct_reorder(interaction, prob.norm)) %>% 
    ggplot(aes(interaction, prob.norm)) +
    geom_col(fill = "#ffb03e", width = 0.85) +
    geom_hline(yintercept = 0, size = BASE_LINE_SIZE) +
    geom_text(
      aes(y = 2.68, label = pathway_name),
      size = BASE_TEXT_SIZE_MM,
      fontface = "italic"
    ) +
    xlab("ligand-receptor pair") +
    scale_y_continuous(
      "total outgoing communication score from treatment resistant cells",
      expand = expansion(0)
    ) +
    coord_flip() +
    theme_nb(grid = FALSE) +
    theme(
      axis.title.x = element_text(hjust = 1),
      axis.ticks.y = element_blank(),
      panel.border = element_blank(),
      panel.grid.major.x = element_line(
        color = "grey92",
        size = BASE_LINE_SIZE
      )
    )
}

plot_contribution_celltype()
ggsave(paste0(params$out,"figures/CD14+_PDL1+_MO_outgoing_comm_score_allCellTypes.png"))

plot_contribution_celltype <- function(cell_type = treat_resist[3], signif = 0.05) {
   df_net_all %>%
    filter(
      pathway_name %in% cellchat@netP$pathways, 
      source == {{cell_type}}, 
      target != {{cell_type}},
      pval <= signif
    ) %>%
    group_by(interaction = interaction_name_2) %>%
    summarise(
      prob.norm = sum(prob.norm),
      count = n(),
      pathway_name = first(pathway_name)
    ) %>%
    ungroup() %>%
    mutate(prob.avg = prob.norm / count) %>% 
    mutate(interaction = fct_reorder(interaction, prob.norm)) %>% 
    ggplot(aes(interaction, prob.norm)) +
    geom_col(fill = "#ffb03e", width = 0.85) +
    geom_hline(yintercept = 0, size = BASE_LINE_SIZE) +
    geom_text(
      aes(y = 2.68, label = pathway_name),
      size = BASE_TEXT_SIZE_MM,
      fontface = "italic"
    ) +
    xlab("ligand-receptor pair") +
    scale_y_continuous(
      "total outgoing communication score from treatment resistant cells",
      expand = expansion(0)
    ) +
    coord_flip() +
    theme_nb(grid = FALSE) +
    theme(
      axis.title.x = element_text(hjust = 1),
      axis.ticks.y = element_blank(),
      panel.border = element_blank(),
      panel.grid.major.x = element_line(
        color = "grey92",
        size = BASE_LINE_SIZE
      )
    )
}

plot_contribution_celltype()
ggsave(paste0(params$out,"figures/CD14+_PDL1+_MO_outgoing_comm_score_treatResistCells.png"))

# 3. CD4+ PD1+ T cell
treat_resist[4]

plot_contribution_celltype <- function(cell_type = treat_resist[4], signif = 0.05) {
   df_net_int %>%
    filter(
      pathway_name %in% cellchat@netP$pathways, 
      source == {{cell_type}}, 
      target != {{cell_type}},
      pval <= signif
    ) %>%
    group_by(interaction = interaction_name_2) %>%
    summarise(
      prob.norm = sum(prob.norm),
      count = n(),
      pathway_name = first(pathway_name)
    ) %>%
    ungroup() %>%
    mutate(prob.avg = prob.norm / count) %>% 
    mutate(interaction = fct_reorder(interaction, prob.norm)) %>% 
    ggplot(aes(interaction, prob.norm)) +
    geom_col(fill = "#ffb03e", width = 0.85) +
    geom_hline(yintercept = 0, size = BASE_LINE_SIZE) +
    geom_text(
      aes(y = 2.68, label = pathway_name),
      size = BASE_TEXT_SIZE_MM,
      fontface = "italic"
    ) +
    xlab("ligand-receptor pair") +
    scale_y_continuous(
      "total outgoing communication score from treatment resistant cells",
      expand = expansion(0)
    ) +
    coord_flip() +
    theme_nb(grid = FALSE) +
    theme(
      axis.title.x = element_text(hjust = 1),
      axis.ticks.y = element_blank(),
      panel.border = element_blank(),
      panel.grid.major.x = element_line(
        color = "grey92",
        size = BASE_LINE_SIZE
      )
    )
}

plot_contribution_celltype()
ggsave(paste0(params$out,"figures/CD4+_PDL1+_T_outgoing_comm_score_allCellTypes.png"))

plot_contribution_celltype <- function(cell_type = treat_resist[4], signif = 0.05) {
   df_net_all %>%
    filter(
      pathway_name %in% cellchat@netP$pathways, 
      source == {{cell_type}}, 
      target != {{cell_type}},
      pval <= signif
    ) %>%
    group_by(interaction = interaction_name_2) %>%
    summarise(
      prob.norm = sum(prob.norm),
      count = n(),
      pathway_name = first(pathway_name)
    ) %>%
    ungroup() %>%
    mutate(prob.avg = prob.norm / count) %>% 
    mutate(interaction = fct_reorder(interaction, prob.norm)) %>% 
    ggplot(aes(interaction, prob.norm)) +
    geom_col(fill = "#ffb03e", width = 0.85) +
    geom_hline(yintercept = 0, size = BASE_LINE_SIZE) +
    geom_text(
      aes(y = 2.68, label = pathway_name),
      size = BASE_TEXT_SIZE_MM,
      fontface = "italic"
    ) +
    xlab("ligand-receptor pair") +
    scale_y_continuous(
      "total outgoing communication score from treatment resistant cells",
      expand = expansion(0)
    ) +
    coord_flip() +
    theme_nb(grid = FALSE) +
    theme(
      axis.title.x = element_text(hjust = 1),
      axis.ticks.y = element_blank(),
      panel.border = element_blank(),
      panel.grid.major.x = element_line(
        color = "grey92",
        size = BASE_LINE_SIZE
      )
    )
}

plot_contribution_celltype()
ggsave(paste0(params$out,"figures/CD4+_PDL1+_T_outgoing_comm_score_treatResistCells.png"))

## 3c ----

plot_selected_dots <- function(interactions, source_type = "NB") {
  vis_data <- 
    sig_data %>% 
    filter(
      interaction_name %in% {{interactions}}, 
      source == {{source_type}}
    ) %>%
    mutate(
      pathway_name = factor(pathway_name),
      source = fct_recode(source, "from NB (source) to" = "NB"),
      target = factor(target, levels = names(CELL_TYPE_ABBREVIATIONS))
    )
  
  ggplot(vis_data, aes(target, interaction_name_2)) +
    geom_point(
      aes(color = prob.norm, size = pmin(5, -log10(pval)))
    ) + 
    scale_radius(
      TeX("-log_{10} p-Value_{cap.}"),
      range = c(1, 5),
      guide = "none"
    ) +
    scale_color_distiller(
      "relative communication score",
      type = "seq",
      palette = "YlOrBr",
      direction = 1,
      breaks = round(range(vis_data$prob.norm), 2),
      guide = guide_colorbar(
        barheight = unit(2, "mm"),
        barwidth = unit(15, "mm"),
        ticks = FALSE,
        title.position = "top"
      )
    ) +
    xlab("cell type (target)") +
    ylab("ligand-receptor pair\nand pathway") +
    facet_grid(
      vars(pathway_name), 
      vars(source),
      scales = "free",
      space = "free",
      switch = "both"
    ) + 
    theme_nb(grid = FALSE) + 
    theme(
      axis.text.x = element_text(
        size = BASE_TEXT_SIZE_PT, 
        margin = margin(t = 1, b = 1, r = 1, l = 1, unit = "mm")
      ),
      axis.text.y = element_text(size = BASE_TEXT_SIZE_PT),
      axis.ticks.length.x = unit(2, "mm"),
      axis.ticks.x = element_line(
        arrow = arrow(length = unit(1,"mm"), ends = "first", type = "closed")
      ),
      axis.title = element_text(size = BASE_TEXT_SIZE_PT),
      
      panel.border = element_rect(color = "black", size = BASE_LINE_SIZE),
      panel.spacing = unit(-BASE_LINE_SIZE / 2, "pt"),
      
      strip.background.y = element_rect(
        fill = "gray95",
        size = BASE_LINE_SIZE
      ),
      strip.background.x = element_rect(
        fill = "gray95",
        size = BASE_LINE_SIZE
      ),
      strip.text.x =  element_text(
        size = BASE_TEXT_SIZE_PT, 
        margin = margin(t = 1, b = 1, r = 4, l = 4, unit = "mm"),
        angle = 0
      ),
      strip.text.y.left = element_text(
        size = BASE_TEXT_SIZE_PT, 
        margin = margin(t = 0.5, b = 0.5, r = 1, l = 1, unit = "mm"),
        angle = 0
      ),
      strip.switch.pad.grid = unit(0, "pt"),
      
      legend.position = c(-.4, -.2),
      legend.direction = "horizontal",
      legend.text = element_text(size = BASE_TEXT_SIZE_PT),
      legend.title = element_text(size = BASE_TEXT_SIZE_PT),
    )
}

plot_selected_dots(c("MDK_NCL", "MDK_LRP1", "MIF_CD74_CXCR4", "MIF_CD74_CD44"))
ggsave(paste0(params$out,"/figures/3c_dots.png"), height = 3.4, width = 7)

```


```{r load previous analysis}
metadata <- readRDS("/home/data/MapMet/fetahu.scRNA.2023/data/metadata.rds")
scrna_data <- readRDS("/home/data/MapMet/fetahu.scRNA.2023/data/rna_decontaminated.rds")
predicted_labels <- readRDS(file = paste0(params$out,"Rds/20240126_predicted_labels_IMCtoRNA.rds"))

rna <- as.Seurat(scrna_data, data=NULL)
rna <- RenameAssays(object = rna, originalexp = 'RNA')

rna@active.ident <- setNames(pull(metadata[match(metadata$cell,colnames(rna)),"cellont_cluster"], "cellont_cluster"),pull(metadata[match(metadata$cell,colnames(rna)), "cell"], "cell"))
DefaultAssay(rna) <- "RNA"

true_label <- rna@active.ident
predicted_label <- predicted_labels$labels

rna$original_celltype <- rna@active.ident
rna$predicted_celltype <- predicted_label

rna@meta.data
```


# __PDL-1+ and PD-1+ immune cells__

```{r}
library(viridis)
pal <- viridis(n = 10, option = "C", direction = -1)

# CD274 gene encodes for the Programmed death-ligand 1 (PDL-1)
pdf(paste0(params$out,"figures/CD274_expression.pdf"))
FeaturePlot(rna, label = TRUE,
            features = "CD274", 
            order = T, cols = pal)
dev.off()


FeaturePlot(rna, label = TRUE,
            features = "CD274", 
            order = T, cols = pal)

# PDCD1 gene encodes for the Programmed cell death protein 1 (PD-1)
pdf(paste0(params$out,"figures/PDCD1_expression.pdf"))
FeaturePlot(rna, label = TRUE,
            features = "PDCD1", 
            order = T, cols = pal)
dev.off()
```

Violin plot for PD1 expression in T cell clusters

```{r Violin plot}
features <- "PDCD1"
tcells <- tcells_pd1 %>%
  NormalizeData(normalization.method = "LogNormalize") %>%
  ScaleData()

VlnPlot(tcells,features,log = TRUE)
DoHeatmap(tcells,features)


# PD1+ T cells only
tcells_pd1 <- tcells_pd1 %>%
  NormalizeData(normalization.method = "LogNormalize") %>%
  ScaleData()

VlnPlot(tcells_pd1,features,log = TRUE)
RidgePlot(tcells_pd1,features,log=TRUE)

# Gene set enrichment score

features <- c("PDCD1","CTLA4","BATF","VHL",
              "LAG3","CD244","CD160","HAVCR2",
              "IL10RA","PRDM1","GATA3","IKZF2","TBX21","EOMES")

tcells <- AddModuleScore(tcells, features,
                         name = "exhustion_markers_set", seed = 28)

tcells_pd1 <- AddModuleScore(tcells_pd1, features,
                         name = "exhustion_markers_set", seed = 28)

# Plot scores
FeaturePlot(tcells,
            features = "exhustion_markers_set1", label = TRUE, repel = TRUE, 
            order = TRUE, pt.size = 1) +
            scale_colour_gradientn(colours = brewer.pal(n = 20, name = "RdPu"))


VlnPlot(tcells,features = "exhustion_markers_set1", log = TRUE,pt.size = 0)

VlnPlot(tcells_pd1,features = "exhustion_markers_set1", log = TRUE,pt.size = 0)

RidgePlot(tcells_pd1, features = "exhustion_markers_set1", log = TRUE)

```



CD4, CD8 and GZMB expression in T-cells

The question here is, which T-cells express PD-1, are these CD8+ or CD4+ T-cells
and are the cells also positive for GZNB?

```{r CD4, CD8 and GZMB expression in T-cells}
FeaturePlot(rna, label = TRUE,
            features = "CD274", 
            order = T, cols = pal)

FeaturePlot(rna, label = TRUE,
            features = "PDCD1", 
            order = T, cols = pal)

FeaturePlot(rna, label = TRUE,
            features = "CD8A", 
            order = T, cols = pal)

FeaturePlot(rna, label = TRUE,
            features = "CD4", 
            order = T, cols = pal)

FeaturePlot(rna, label = TRUE,
            features = "S100B", 
            order = T, cols = pal)
 
FeaturePlot(rna, label = TRUE,
            features = "GZMB", 
            order = T, cols = pal)
```

Is PD1 co-expressed with CD8/CD4/GZMB/S100B?

```{r PD1+ T-cells}
# T-cells
tcells <- subset(x = rna, idents = c("T (6)","T (9)","T (5)")) # 19126 
# with "T (18)" -> 19428  cells

# PD1+ T cells
tcells_pd1 <- subset(x = rna, idents = c("T (6)","T (9)","T (5)"),
                 subset = PDCD1 > 0.1, ) 
# with "T (18)" -> 552 cells


# S100B+ cells
tcells_s100b <- subset(x = rna, idents = c("T (6)","T (9)","T (5)"),
                 subset = S100B > 0.1, ) # 897 cells


features <- c("CD8A","CD4","GZMB","S100B")
DotPlot(tcells, features = features) + RotatedAxis()
DotPlot(tcells_pd1, features = features) + RotatedAxis()
```

CD69 - early T cell activation marker
HLA-DR - late T cell activation marker
CD4 - T helper cell marker (CD4+ T cells)
CD8 - cytotoxic T cell marker (CD8+ T cells) 
CD3 - predominantly located on T cells

T cell activation (HLA-DR and CD38), 
senescence and exhaustion (CD57 and PD1), 
CD57 = B3GAT1 
CD25 = IL2RA
differentiation (CD45RA, CCR7, CD28, and CD27), 
regulatory T cells (Treg, considered as CD25high and CD127low), 

```{r S100B+ T-cells}
features <- c("CD69","HLA-DRA", "CD38", "IL2RA","B3GAT1", 
              "PTPRC","CCR7","CD28","CD27",
              "CD3G","CD8A","CD4")
DotPlot(tcells, features = features) + RotatedAxis()

DotPlot(tcells_s100b, features = features) + RotatedAxis()
```


Co-expression of T-cell exhustion markers

Elevated co-expression of inhibitory checkpoint receptors such as PDCD1, CTLA4, 
HAVCR2, and LAG3, TOX TFs, PRDM1, NR4A TFs, KLRG1, B3GAT1

CXCL13 is a common marker of PD-1 high CD8+ and PD-1 high CD4+ T cells 

High levels of expression of GZMB and FASL but low levels of IFNG and TNF

```{r T-cell exhustion markers, fig.height=10, fig.width=10}
features <- c("PDCD1","CTLA4", "LAG3", "HAVCR2", "CSF1", "CD38",
              "PRDM1", "TOX", "TOX2", "TOX3", "TOX4", "IRF4", "BATF4",
              "NR4A1", "NR4A2", "NR4A3", "KLRG1", "B3GAT1", "MKI67",
              "GZMB" , "FASLG" , "IFNG" , "TNF", "CXCL13","TCF7",
              "IGKC", "HLADRA","CCL3", "CXCR6", "IL13","CLCF1", 
              "OSM", "IL10RA","CCL4","CCL5","CCL16",
               "CCR3", "CCR4", "CCR5", "CCR6", "PTPRC")

#RidgePlot(rna, features = features, ncol = 2)
#VlnPlot(rna, features = features)
#FeaturePlot(rna, features = features)


x <- DotPlot(rna, features = features) + RotatedAxis()
x

DotPlot(tcells, features = features) + RotatedAxis()




pdf(paste0(params$out,"figures/Tcell_exhustion_markers_dotPlot.pdf"))
x
dev.off()

# Tumor reactivity-related genes
tumor_reactivity_genes <- c("TCF7", "PDCD1", "LAG3", "HAVCR2", "CTLA4" , 
                            "TNFRSF9" , "TNFRSF18" , "ENTPD1", "ITGAE", "CXCL13")

tumor_reactivity_genes_2 <- c("PDCD1", "LAG3", "HAVCR2", "CTLA4" , 
                            "TNFRSF9" , "TNFRSF18" , "ENTPD1", "CXCL13")

# IL-8 receptors
IL8_receptors <- c("IL18","IL18BP", "IL1A", "IL1B", "IL1F10", "IL1F3","IL1RA", "IL1F5", "IL1F6",
 "IL1F7", "IL1F8", "IL1RL2", "IL1F9", "IL33")
DotPlot(rna, features = IL8_receptors) + RotatedAxis()

# IL-1 Receptors
IL1_receptors <- c("IL18R1", "IL18RAP", "IL1R1", "IL1R2", "IL1R3", "IL1R8", "IL1R9", "IL1RL1", "SIGIRR")
DotPlot(rna, features = IL1_receptors) + RotatedAxis()

# Interferon (IFN)
IFN_receptors <- c("IFNA1", "IFNA10", "IFNA13", "IFNA14", "IFNA2", "IFNA4", 
                   "IFNA7", "IFNB1", "IFNE", "IFNG", "IFNZ", "IFNA8", "IFNA5", 
                   "IFNW1","IFNAR1" ,"IFNAR2", "IFNGR1", "IFNGR2")
DotPlot(rna, features = IFN_receptors) + RotatedAxis()

# IFN Receptor

# IL6 Family & Receptors
IL6_family_receptors <- c("CLCF1", "CNTF", "IL11", "IL31", "IL6", "Leptin" ,"LIF", 
                   "OSM","CNTFR", "IL11RA", "IL6R", "LEPR", "LIFR", "OSMR", "IL31RA")
DotPlot(rna, features = IL6_family_receptors) + RotatedAxis()


# IL10 Family & Receptors
IL10_family_receptors <- c("IL10", "IL19", "IL20", "IL22", "IL24", "IL28B", 
                           "IL28A", "IL29", "IL10RA", "IL10RB", "IL20RA", 
                           "IL20RB", "IL22RA2", "IL22R")

DotPlot(rna, features = IL10_family_receptors) + RotatedAxis()

# TGF beta Family
#TGF-beta 1/TGFB1TGF-beta 2/TGFB2TGF-beta 3/TGFB3
# TGF beta Family Receptor
#ALK-7ATF2CD105/ENGTGFBR1TGFBR2TGFBR3

# Chemokine
Chemokines <- c("CCL1","CCL11","CCL12","CCL13","CCL14","CCL15","CCL16","CCL17",
                "CCCL18","CCL19","CCL2","CCL20","CCL21","CCL22","CCCL23","CCL24",
                "CCL25","CCL26","CCL27","CCL28","CCL3","CCL3L3","CCL4","CCL4L1",
                "CCL5","CCL6","CCL7","CCL8","CCL9","CX3CL1","CXCL1","CXCL10",
                "CXCL11","CXCL12","CXCL13","CXCL14","CXCL15","CXCL16","CXCL17",
                "CXCL2","CXCL3","CXCL4","CXCL5","CXCL6","CXCL7","CXCL9IL8",
                "CXCL8","XCL1","XCL2","FAM19A1","FAM19A2","FAM19A3","FAM19A4",
                "FAM19A5")
DotPlot(rna, features = Chemokines) + RotatedAxis()

# Chemokine Receptor
Chemokine_receptors <- c("CCR1", "CCR2", "CCR3", "CCR4", "CCR5", "CCR6", "CCR7",
                         "CCR8", "CCRL1", "CXCR3", "CXCR4", "CXCR5", "CXCR6",
                         "CXCR7", "CXCR1", "CXCR2")
DotPlot(rna, features = Chemokine_receptors) + RotatedAxis()


y <- DotPlot(rna, features = tumor_reactivity_genes) + RotatedAxis()
y

DotPlot(rna, features = tumor_reactivity_genes_2) + RotatedAxis()

pdf(paste0(params$out,"figures/Tcell_tumor_reactivity_markers_dotPlot.pdf"))
y
dev.off()

```

## T-cell exhustion markers from Sevgi and Christoph (CCRI)

```{r}
sevgi_christoph_gene_set <- c("PDCD1","CTLA4","NFATC1","SPRY2","BATF","VHL",
                              "FOXO1","FOXP1","LAG3","CD244","CD160","HAVCR2",
                              "TRAF1","TNFRSF9","IL10RA","IL10","PRDM1","STAT3",
                              "IFNA1","IFNB1","IL21","CXCR5","SOCS3","GATA3",
                              "IKZF2","BCL6","BCL2","TBX21","EOMES")

sevgi_christoph_gene_set <- c("PDCD1","CTLA4","BATF","VHL",
                              "LAG3","CD244","CD160","HAVCR2",
                              "IL10RA","PRDM1","GATA3","IKZF2","TBX21","EOMES")

DotPlot(tcells, features = sevgi_christoph_gene_set,dot.scale = 10,
        cols = c("gray","purple")) + RotatedAxis()

sevgi2024

pdf(paste0(params$out,"figures/Tcell_exhustion_markers_fromSevgi-CCRI_dotPlot.pdf"))
sevgi2024
dev.off()

```

## T-cell exhustion markers from Mei et al., 2024

```{r Mei et al., 2024}
gene_list_mei2024 <- c("TOX", "NR4A1","NR4A2","NR4A3","PDCD1","CTLA4","TIGIT",
                       "HAVCR2","LAG3","BTLA","CD244","TBX21","EOMES",
                       "BATF","VHL","FOXO1", "FOXO3", "FOXO4",  "FOXO6","FOXP1",
                       "ENTPD1","CXCL13",
                       "MYO1E","PRDM1","WARS","CREM","LAYN","PHLDA1",
                       "SNAP47","CD38","MYO7A","PARK7",
                       "NFATC1", "NFATC2", "NFATC3", "NFATC4", "NFAT5")

mei2024 <- DotPlot(tcells, features = gene_list_mei2024) + RotatedAxis()
mei2024_pd1 <- DotPlot(tcells_pd1, features = gene_list_mei2024) + RotatedAxis()

mei2024
mei2024_pd1

pdf(paste0(params$out,"figures/Tcell_exhustion_markers_fromMei2024_dotPlot.pdf"))
mei2024
dev.off()

```

# ___CCC analysis___

# Fetahu et al., 2023
```{r original labels}

true_label <- rna@active.ident
predicted_label <- predicted_labels$labels

rna$original_celltype <- rna@active.ident
rna$predicted_celltype <- predicted_label

Idents(rna) <- rna@meta.data$original_celltype


# CellChat infers the biologically significant cell-cell communication by 
# assigning each interaction with a probability value and peforming a permutation test. 
pathway <- c("CXCL", "CCL")
treat_resist_fetahu_lables <- subset(x = rna, 
                                     idents = c("T (5)","T (6)","T (9)","T (18)",
                                                "NB (8)", 
                                                "M (1)","M (2)","M (10)","M (15)"))
cellchat <- createCellChat(object = treat_resist_fetahu_lables, 
                           group.by = "ident", assay = "RNA")

# set the used database in the object
cellchat@DB <- CellChatDB

# Prepare DB
CellChatDB <- CellChatDB.human 
showDatabaseCategory(CellChatDB)


# subset the expression data of signaling genes for saving computation cost
cellchat <- subsetData(cellchat) # This step is necessary even if using the whole database
cellchat <- identifyOverExpressedGenes(cellchat)
cellchat <- identifyOverExpressedInteractions(cellchat)

# The number of highly variable ligand-receptor pairs used for signaling inference is 2893 

# project gene expression data onto PPI (Use `raw.use = FALSE` in the function 
# computeCommunProb() in order to use the projected data)
cellchat <- projectData(cellchat, PPI.human)

# Changing the parameter for how many cells (per cluster/cell type) have to 
# express the ligand/receptor to be considered in the analysis

computeAveExpr(cellchat, features = c("PDCD1","CD274"), 
               type =  "truncatedMean", trim = 0.01)


# Compute the communication probability and infer cellular communication network
cellchat <- computeCommunProb(cellchat,  
                              #type = "truncatedMean", trim = 0.01,
                              raw.use = FALSE, population.size = TRUE)

cellchat <- filterCommunication(cellchat, min.cells = 50)
# population.size = TRUE <- cell numbers per cell type/cluster 
# used for the probability calculation 

df.net <- subsetCommunication(cellchat)
df.net.pathways <- subsetCommunication(cellchat, slot.name = "netP" )
# slot.name = "netP" for nferred communications at the level of signaling pathways
df.net.pathways

# Infer the cell-cell communication at a signaling pathway level
cellchat <- computeCommunProbPathway(cellchat)

# Calculate the aggregated cell-cell communication network
cellchat <- aggregateNet(cellchat)

groupSize <- as.numeric(table(cellchat@idents))
par(mfrow = c(1,2), xpd=TRUE)
netVisual_circle(cellchat@net$count, vertex.weight = groupSize, 
                 weight.scale = T, label.edge= F, 
                 title.name = "Number of interactions")
netVisual_circle(cellchat@net$weight, vertex.weight = groupSize, 
                 weight.scale = T, label.edge= F, 
                 title.name = "Interaction weights/strength")

# Bubble plot: all the significant interactions (Ligand-Receptor pairs) from 
# one cell groups (defined by 'sources.use') to other cell groups (defined by 'targets.use')

netVisual_bubble(cellchat, sources.use = 3,  remove.isolate = FALSE)
netVisual_chord_gene(cellchat, sources.use = 3, slot.name = "netP", legend.pos.x = 10)

# Compute the network centrality scores
cellchat <- netAnalysis_computeCentrality(cellchat, slot.name = "netP") 
# the slot 'netP' means the inferred intercellular communication network of signaling pathways
# Visualize the computed centrality scores using heatmap, allowing ready identification of major signaling roles of cell groups
netAnalysis_signalingRole_network(cellchat, signaling = pathway, 
                                  width = 8, height = 2.5, font.size = 10)

# Visualize dominant senders (sources) and receivers (targets) in a 2D space

# Signaling role analysis on the aggregated cell-cell communication network from all signaling pathways
gg1 <- netAnalysis_signalingRole_scatter(cellchat)
#> Signaling role analysis on the aggregated cell-cell communication network from all signaling pathways
# Signaling role analysis on the cell-cell communication networks of interest
gg2 <- netAnalysis_signalingRole_scatter(cellchat, signaling = pathway)
#> Signaling role analysis on the cell-cell communication network from user's input
gg1 + gg2

# Signaling role analysis on the aggregated cell-cell communication network from all signaling pathways
ht1 <- netAnalysis_signalingRole_heatmap(cellchat, pattern = "outgoing")
ht2 <- netAnalysis_signalingRole_heatmap(cellchat, pattern = "incoming")
ht1
ht2

library(NMF)
library(ggalluvial)

# run selectK to infer the number of patterns.
#selectK(cellchat, pattern = "outgoing") # define the number of patterns/cut-off

nPatterns = 3
cellchat <- identifyCommunicationPatterns(cellchat, pattern = "outgoing", k = nPatterns)

# river plot
netAnalysis_river(cellchat, pattern = "outgoing")

# dot plot
netAnalysis_dot(cellchat, pattern = "outgoing")


#selectK(cellchat, pattern = "incoming") # define the number of patterns/cut-off

nPatterns = 7
cellchat <- identifyCommunicationPatterns(cellchat, pattern = "incoming", k = nPatterns)

saveRDS(cellchat, file = paste0(params$out,"/Rds/CCC_analysis_Fetahu2023.Rds"))


df.net.fetahu <- subsetCommunication(cellchat,
                            slot.name = "net")

df.path.fetahu <- subsetCommunication(cellchat,
                               slot.name = "netP")

# Signalling source: Cluster 9
filt <- df.net.pathways$source == "T (9)"
df.net.pathways[filt,]

# Make sure the object@dr contains a low-dimensional space of the data such as 
# umap and tsne in order to produce the feature plot of signaling genes. 
# A new reduced space can be add via the function addReduction.

library("shiny")
library("bslib")
runCellChatApp(cellchat)

```

# Lazic et al., 2024
```{r predicted labels}
# Change the cell type annotation to predicted labels

true_label <- rna@active.ident
predicted_label <- predicted_labels$labels

rna$original_celltype <- rna@active.ident
rna$predicted_celltype <- predicted_label

Idents(rna) <- rna@meta.data$predicted_celltype



# CellChat infers the biologically significant cell-cell communication by 
# assigning each interaction with a probability value and peforming a permutation test. 

treat_resist <- c("CD14- PDL1+ MO/DC",
                  "CD14+ PDL1+ MO", 
                  "CD14+ PDL1+ MO",
                  "CD4+ PD1+ T cell",
                  "CD8+ PDL1lo T cell", 
                  "CD8+ S100B+ T cell",
                  "CD8+ GZMB+ PD1lo T cell",
                  "early SYM-like TC",
                  "CD44+ mes-like TC")

#treat_resist <- c("early SYM-like TC","CD44+ mes-like TC",
#                 "CD14+ PDL1+ MO","CD4+ PD1+ T cell")

treat_resist_lazic_lables <- subset(x = rna, 
                                     idents = treat_resist) #5814  cells
cellchat_predict <- createCellChat(object = treat_resist_lazic_lables, 
                           group.by = "ident", assay = "RNA")

table(treat_resist_lazic_lables$predicted_celltype)

# early SYM-like TC -> only 132 cells !!

# Prepare DB
cellchatDB <-  CellChatDB.human 
showDatabaseCategory(cellchatDB)

# set the used database in the object
cellchat_predict@DB <- cellchatDB


# subset the expression data of signaling genes for saving computation cost
cellchat_predict <- subsetData(cellchat_predict) # This step is necessary even if using the whole database
cellchat_predict <- identifyOverExpressedGenes(cellchat_predict)
cellchat_predict <- identifyOverExpressedInteractions(cellchat_predict)

# The number of highly variable ligand-receptor pairs used for signaling inference is 2601 

# project gene expression data onto PPI (Use `raw.use = FALSE` in the function 
# `computeCommunProb()` in order to use the projected data)
cellchat_predict <- projectData(cellchat_predict, PPI.human)

# Changing the parameter for how many cells (per cluster/cell type) have to 
# express the ligand/receptor to be considered in the analysis

computeAveExpr(cellchat_predict, features = c("PDCD1","CD274"), 
               type =  "truncatedMean", trim = 0.01)


# Compute the communication probability and infer cellular communication network
cellchat_predict <- computeCommunProb(cellchat_predict,  
                              #type = "truncatedMean", trim = 0.01,
                              raw.use = FALSE, population.size = TRUE)

# population.size = TRUE <- cell numbers per cell type/cluster 
# used for the probability calculation 

cellchat_predict <- filterCommunication(cellchat_predict, min.cells = 10)


df.net <- subsetCommunication(cellchat_predict)
df.net.pathways <- subsetCommunication(cellchat_predict, slot.name = "netP" )
# slot.name = "netP" for nferred communications at the level of signaling pathways
df.net.pathways

# Infer the cell-cell communication at a signaling pathway level
cellchat_predict <- computeCommunProbPathway(cellchat_predict)

# Calculate the aggregated cell-cell communication network
cellchat_predict <- aggregateNet(cellchat_predict)

groupSize <- as.numeric(table(cellchat_predict@idents))
par(mfrow = c(1,2), xpd=TRUE)
netVisual_circle(cellchat_predict@net$count, vertex.weight = groupSize, 
                 weight.scale = T, label.edge= F, 
                 title.name = "Number of interactions")
netVisual_circle(cellchat_predict@net$weight, vertex.weight = groupSize, 
                 weight.scale = T, label.edge= F, 
                 title.name = "Interaction weights/strength")

# Bubble plot: all the significant interactions (Ligand-Receptor pairs) from 
# one cell groups (defined by 'sources.use') to other cell groups (defined by 'targets.use')

netVisual_bubble(cellchat_predict, sources.use = 3,  remove.isolate = FALSE)
netVisual_chord_gene(cellchat_predict, sources.use = 3, 
                     slot.name = "netP", legend.pos.x = 10)

# Compute the network centrality scores
cellchat_predict <- netAnalysis_computeCentrality(cellchat_predict, 
                                                  slot.name = "netP") 
# the slot 'netP' means the inferred intercellular communication network of signaling pathways
# Visualize the computed centrality scores using heatmap, allowing ready identification of major signaling roles of cell groups
netAnalysis_signalingRole_network(cellchat_predict, signaling = pathway, 
                                  width = 8, height = 2.5, font.size = 10)

# Visualize dominant senders (sources) and receivers (targets) in a 2D space
gg1 <- netAnalysis_signalingRole_scatter(cellchat_predict)
# Signaling role analysis on the cell-cell communication networks of interest
gg2 <- netAnalysis_signalingRole_scatter(cellchat_predict, signaling = pathway)
#> Signaling role analysis on the cell-cell communication network from user's input
gg1
gg2

# Signaling role analysis on the aggregated cell-cell communication network from all signaling pathways
ht1 <- netAnalysis_signalingRole_heatmap(cellchat_predict, pattern = "outgoing")
ht2 <- netAnalysis_signalingRole_heatmap(cellchat_predict, pattern = "incoming")
ht1
ht2

library(NMF)
library(ggalluvial)

# run selectK to infer the number of patterns.
#selectK(cellchat_predict, pattern = "outgoing") # define the number of patterns/cut-off

nPatterns = 2
cellchat_predict <- identifyCommunicationPatterns(cellchat_predict, 
                                                  pattern = "outgoing", 
                                                  k = nPatterns)

# river plot
netAnalysis_river(cellchat_predict, pattern = "outgoing")

# dot plot
netAnalysis_dot(cellchat_predict, pattern = "outgoing")


#selectK(cellchat_predict, pattern = "incoming") # define the number of patterns/cut-off

nPatterns = 2
cellchat_predict <- identifyCommunicationPatterns(cellchat_predict, 
                                                  pattern = "incoming", 
                                                  k = nPatterns)


saveRDS(cellchat_predict, file = paste0(params$out,
                                        "/Rds/CCC_analysis_Lazic2024_v2.Rds"))
 

df.net.lazic <- subsetCommunication(cellchat_predict,
                                    slot.name = "net")

df.path.lazic <- subsetCommunication(cellchat_predict,
                                     slot.name = "netP")

# runcellchat_predictApp(cellchat_predict)

```

# __Tumor cluster downstream analysis__

```{r}
fetahu.NB <- readRDS(file = paste0(params$out,"Rds/fetahu.scRNA.tumor.cluster.Rds"))


# Predicted cell types based on IMC data
Idents(fetahu.NB) <- fetahu.NB@meta.data$predicted_celltype

# Remove cell type labels with less than 100 cells
filter <- table(fetahu.NB@meta.data$predicted_celltype) > 100
filter <- subset(filter, filter == TRUE)
TC_clusters <- rownames(filter)
TC_clusters

fetahu.NB.filt <- subset(fetahu.NB, 
                    idents = TC_clusters, 
                    invert = FALSE)

fetahu.NB.filt

treat_resist_TC <- c("early SYM-like TC", "CD44+ mes-like TC")

treat_resist_TC_cells <- fetahu.NB.filt@meta.data %>%
    filter(predicted_celltype == treat_resist_TC[1] | predicted_celltype == treat_resist_TC[2])
treat_resist_TC_cells <- rownames(treat_resist_TC_cells)

treat_resist_TC_early_SYM <- fetahu.NB.filt@meta.data %>%
    filter(predicted_celltype == treat_resist_TC[1])
treat_resist_TC_early_SYM <- rownames(treat_resist_TC_early_SYM)

treat_resist_TC_MES_like <- fetahu.NB.filt@meta.data %>%
    filter(predicted_celltype == treat_resist_TC[2])
treat_resist_TC_MES_like <- rownames(treat_resist_TC_MES_like)

treat_resist_TC_MES_like_col = rgb(27,158,119, maxColorValue = 255)
treat_resist_TC_early_SYM_col = rgb(204,235,197, maxColorValue = 255)

# Visualization
DimPlot(fetahu.NB.filt, reduction = "umap", group.by = "seurat_clusters")
DimPlot(fetahu.NB.filt, reduction = "umap", group.by = "sample")
DimPlot(fetahu.NB.filt, reduction = "umap", group.by = "group")
DimPlot(fetahu.NB.filt, reduction = "umap",  label = TRUE, repel = TRUE)
DimPlot(fetahu.NB.filt, reduction = "umap", repel = TRUE,
        cells.highlight = list("early SYM-like TC" = treat_resist_TC_early_SYM, 
                               "CD44+ MES-like TC" = treat_resist_TC_MES_like), 
        cols.highlight = c(treat_resist_TC_early_SYM_col,
                           treat_resist_TC_MES_like_col), 
        cols= "grey")



# find cluster markers with Seurat
fetahu.NB.filt.markers <- FindAllMarkers(fetahu.NB.filt, 
                                      min.pct  = 0.1, 
                                      logfc.threshold = 0.25,
                                      min.cells.group = 50,
                                      random.seed = 28)
fetahu.NB.filt.markers %>%
    group_by(cluster) %>%
    slice_max(n = 5, order_by = avg_log2FC)

fetahu.NB.filt.markers %>%
    group_by(cluster) %>%
    top_n(n = 5, wt = avg_log2FC) -> top5
DoHeatmap(fetahu.NB.filt, features = top5) + NoLegend()

print(top5, n = 70)


library(DElegate)

# find cluster markers with DElegate 
de_results <- DElegate::findDE(object = fetahu.NB.filt,
                               group_column = "predicted_celltype")
marker_results <- DElegate::FindAllMarkers2(object = fetahu.NB.filt,
                                            group_column = "predicted_celltype")
dplyr::filter(marker_results, feature_rank < 6)


marker_results %>%
    group_by("group1") %>%
    top_n(n = 5, wt = log_fc) -> top5_predict


print(top5_predict, n = 70)


#####################################################################

# Seurat clusters
Idents(fetahu.NB.filt) <- fetahu.NB.filt@meta.data$seurat_clusters

DoHeatmap(fetahu.NB.filt, features = NB.marker.genes, assay = "RNA")

# find cluster markers with DElegate 
de_results <- DElegate::findDE(object = fetahu.NB.filt,
                               group_column = "seurat_clusters")
marker_results <- DElegate::FindAllMarkers2(object = fetahu.NB.filt,
                                            group_column = "seurat_clusters")
top_5 <- dplyr::filter(marker_results, feature_rank < 6)

DoHeatmap(fetahu.NB.filt, features = top_5$feature, assay = "RNA")

```

```{r}
DoHeatmap(fetahu.NB.filt,features = gene.list)

NB.marker.genes <- c("NCAM1", "B4GALNT1", "L1CAM", "MYCN", "GATA2", 
                     "DBH", "CHGB", "CHGA", "SYP", "TH", "PHOX2A", "PHOX2B")

DoHeatmap(fetahu.NB, features = NB.marker.genes,assay = "RNA")

top100 <- head(VariableFeatures(fetahu.NB), 100)
top100
DoHeatmap(fetahu.NB, features = top100) + NoLegend()


saveRDS(fetahu.NB, file = paste0(params$out,"Rds/fetahu.scRNA.tumor.cluster.Rds"))

```
# Dong et al., 2021 dataset

<br>

NB primary tumors at diagnosis

```{r}
dong <- readRDS(file="/home/out/CancReg/scRNA-seq_stage_M/results/published/data/tumor_data_dong.rds")
```


```{r end}
sessionInfo()
```

