---
title: "rarity_test"
output: html_document
date: '2023-04-26'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(ggplot2)
```

In this RMD file, we will perform clustering on the curated data (after excluding cells that swam off between IF and IMC and segmentation artifacts). First, we will read our SPE.

```{r read-data}

path <- "/mnt/Multimodal_Imaging_Daria/_Data_Analysis/_tmp_daria/Image_analysis/20230112_DIMR_Ilastik/20230312_sc_data_combined"

spe <- readRDS(file.path(path,"data/spe_1px_wo-ghosts.rds"))

```

Then we will generate matrices from morphological and intensity features alone and combined. 

```{r generate-dataframes, message=FALSE, fig.height=7}
library(bruceR)

# morph_names <- c("area", 
#                  "area_convex", 
#                  "area_filled", 
#                  "axis_major_length", 
#                  "axis_minor_length", 
#                  "eccentricity", 
#                  "equivalent_diameter_area", 
#                  "perimeter", 
#                  "solidity", 
#                  "assymetry", 
#                  "concavity", 
#                  "fill", 
#                  "aspect_ratio-0", 
#                  "perimeter_ratio-0")

morph_names <- c("area",
                 "eccentricity"
)

robustScaler <- function(data, upper, lower) {
  scaled_data <- (data-median(data))/(quantile(data, probs = c(upper)) - quantile(data, probs = c(lower)))
  return(scaled_data)
}

zscore_Scaler <- function(data) {
  scaled_data <- (data-mean(data))/sd(data)
  return(scaled_data)
}

clipScaler <- function(data, upper, lower, min, max) {
  data[data > quantile(data, probs = c(upper))] <- quantile(data, probs = c(upper))
  data[data < quantile(data, probs = c(lower))] <- quantile(data, probs = c(lower))
  scaled_data <- RESCALE(data, to=min:max)
  return(scaled_data)
}
intensity_mat <- assay(spe, "exprs")
colnames(intensity_mat) <- spe$sample
intensity_mat <- intensity_mat[!grepl("mean-80", rownames(intensity_mat)),]


intensity_mat_hq <- intensity_mat[!grepl("CD274|IDO|DAPI|2IF_GD2|3IF_CD56|DNA1|H4K12Ac|H3K9Ac|CXCR2|PNMT|Fibro", rownames(intensity_mat)),]
intensity_mat_lq <- intensity_mat[grepl("CD274|IDO|DAPI|2IF_GD2|3IF_CD56|DNA1|H4K12Ac|H3K9Ac|CXCR2|PNMT|Fibro", rownames(intensity_mat)),]
#intensity_mat_hq <- intensity_mat[!grepl("DAPI|2IF_GD2|3IF_CD56|DNA1|H4K12Ac|H3K9Ac|PNMT", rownames(intensity_mat)),]
#intensity_mat_lq <- intensity_mat[grepl("DAPI|2IF_GD2|3IF_CD56|DNA1|H4K12Ac|H3K9Ac|PNMT", rownames(intensity_mat)),]

#Make sure to exclude cells that are zero for all markers as PG won't work otherwise
non_zeros <- as.logical(colSums(intensity_mat_hq)!=0)

intensity_mat_hq_wo_zeros <- intensity_mat_hq[,non_zeros]
intensity_mat_lq_wo_zeros <- intensity_mat_lq[,non_zeros]
spe <- spe[,non_zeros]


cell_id <- colnames(spe)
tissue <- spe$tissue

morph_mat <- as.matrix(t(colData(spe)[,morph_names]))
rownames(morph_mat) <- gsub('-0','',rownames(morph_mat))
#morph_mat_scaled <- t(apply(morph_mat, 1, RESCALE, to=0:1))
#morph_mat_scaled <- t(apply(morph_mat, 1, robustScaler, upper=0.95, lower=0.05))
#morph_mat_scaled <- t(apply(morph_mat, 1, zscore_Scaler))
#morph_mat_scaled <- t(apply(morph_mat, 1, clipScaler, upper=0.9995, lower=0.0005, min=0, max=mean(apply(intensity_mat_scaled,1,max))))
morph_mat_scaled <- t(apply(morph_mat, 1, clipScaler, upper=0.995, lower=0.005, min=0, max=1))
colnames(morph_mat_scaled) <- spe$sample


#features_mat <- rbind(intensity_mat_hq, morph_mat_scaled)

```

In a next step, we will plot a single-cell heatmap using complex heatmap. 

```{r single-cell-heatmap, message=FALSE, fig.height=20}
library(ComplexHeatmap)
library(viridis)
library(circlize)
library(RColorBrewer)
library(paletteer)

mat <- intensity_mat_hq_wo_zeros

qual_col_pals = brewer.pal.info[brewer.pal.info$category == 'qual',]
col_vector_74 = unlist(mapply(brewer.pal, qual_col_pals$maxcolors, rownames(qual_col_pals)))
col_vector_433 = colors()[grep('gr(a|e)y', grDevices::colors(), invert = T)]

col_sample = setNames(sample(col_vector_433, length(unique(spe$sample))), unique(spe$sample))

set.seed(221228)
rand_cells <- sample(seq_len(ncol(mat)), 2000)

column_ha = HeatmapAnnotation(sample = colnames(mat[,rand_cells]), col=list(sample=col_sample), show_legend=F)
Heatmap(mat[,rand_cells], 
        column_title = "Single-cell heatmap",
        top_annotation = column_ha, 
        show_column_dend = F,
        show_column_names=F,
        col=viridis(100)
        )
```

### Install reticulate package

```{r libraries-reticulate}
if (!"reticulate" %in% installed.packages())
  install.packages("reticulate")
library(reticulate)
```

### Creating a conda environment with reticulate

```{r create-environment}
# Path to conda binary
options(reticulate.conda_binary = "/home/rstudio/anaconda3/bin/conda")

# I had to indicate the path to python with
RETICULATE_PYTHON = "/home/rstudio/anaconda3/envs/rarity/bin/python3.9"

# Activate the environment
Sys.setenv(PYTHONPATH = "/home/rstudio/anaconda3/envs/rarity/lib/python3.9/site-packages")
use_condaenv(condaenv = "rarity", required = TRUE)
py_config()

```

Subsample the data.

```{r subsample, message=FALSE, fig.height=17}

set.seed(20230317)

#Get same number of PT and BM cells
pt_index <- which(tissue=="PT")
bm_index <- which(tissue=="BM")
rand_cells_pt <- sample(seq_len(length(pt_index)), 8192)
rand_cells_bm <- sample(seq_len(length(bm_index)), 8192)
pt_index_rand <- pt_index[rand_cells_pt]
bm_index_rand <- bm_index[rand_cells_bm]
rand_cells <- c(pt_index_rand, bm_index_rand)

mat <- t(intensity_mat_hq_wo_zeros)
subset <- mat[rand_cells,]
```

### Clustering with rarity

More information can be found [here](https://rstudio.github.io/reticulate/articles/versions.html).

```{r phenograph-clustering, fig.height=33}
library(tidyverse)

use_python("/home/rstudio/anaconda3/envs/rarity/bin/python3")

source("R/core.R")
source("R/helpers.R")

Y <- t(intensity_mat_hq_wo_zeros)

start <- Sys.time()
m <- Rarity_fit(Y, n_iter = 20000L, batch_size = 512)
saveRDS(m, file.path(path, "data/rarity-clusters.rds"))

Sys.time() - start

plot_binary_signatures(m, n_cells = 2500, column_names_gp = grid::gpar(fontsize = 16), row_names_gp = grid::gpar(fontsize = 18), column_names_rot = 45)
# ggsave(filename=file.path(path, "data/rarity_heatmap.tiff"), device='tiff')
# 
plot_heatmap_with_observations(Y, m, n_cells = 5000)

clusters <- m$cluster_allocations

```

Now we will look at these clusters in a single-cell heatmap ordered by clusters. 

```{r single-cell-heatmap-clusters, message=FALSE, fig.height=17}
library(ComplexHeatmap)
library(viridis)
library(circlize)
library(RColorBrewer)
library(paletteer)

#set.seed(20230324)
#set.seed(20230327)
#set.seed(20230342)
set.seed(20230317)

mat <- intensity_mat_hq_wo_zeros
clusters <- m$cluster_allocations
tissue <- spe$tissue

clusters[clusters %in% which(m$counts<2500)] <- 0 #only look at clusters with a minimum of 5000 cells

n <- length(unique(clusters))
col_vector_433 = colors()[grep('gr(a|e)y', grDevices::colors(), invert = T)]
#just for rarity
#col_clusters <- setNames(sample(col_vector_433, n), unique(clusters))
col_clusters <- setNames(sample(rep(col_vector_433, ceiling(n/433)), n), unique(clusters))

col_marker = colorRamp2(c(0, 2.5, 0.5), c("blue", "white", "red"))
col_tissue = c("PT" = "deepskyblue", "BM" = "orange")


mat <- mat[, clusters!=0]

tissue = tissue[clusters!=0]
clusters = clusters[clusters!=0]

#Get same number of PT and BM cells
pt_index <- which(tissue=="PT")
bm_index <- which(tissue=="BM")
rand_cells_pt <- sample(seq_len(length(pt_index)), 4000)
rand_cells_bm <- sample(seq_len(length(bm_index)), 4000)
pt_index_rand <- pt_index[rand_cells_pt]
bm_index_rand <- bm_index[rand_cells_bm]
rand_cells <- c(pt_index_rand, bm_index_rand)

#rand_cells <- sample(seq_len(ncol(mat)), 2000)

column_ha = HeatmapAnnotation(cluster = clusters[rand_cells],
                              tissue = tissue[rand_cells],
                              col=list(cluster=col_clusters, 
                                       tissue=col_tissue 
                                       ), 
                              show_legend=T)


Heatmap(mat[,rand_cells], 
        column_title = "Single-cell heatmap with phenograph clusters (1-px mean)",
        top_annotation = column_ha, 
        cluster_columns=F,
        column_order= order(clusters[rand_cells], decreasing=F),
        show_column_dend = F,
        show_column_names=F,
        col=viridis(100)
        )
```

We will now cluster the data.

```{r phenograph-clustering}
library(FastPG)
library(parallel)

set.seed(221228)
n_threads <- detectCores()

mat <- t(intensity_mat_hq_wo_zeros)

out <- FastPG::fastCluster(data=mat, k=45, num_threads=n_threads)
clusters <- factor(out$communities)
saveRDS(clusters, file.path(path, "data/clusters-int-1px-mean_postcuration.rds"))

```

Now we will look at these clusters in a single-cell heatmap ordered by clusters. 

```{r single-cell-heatmap-clusters, message=FALSE, fig.height=17}
library(ComplexHeatmap)
library(viridis)
library(circlize)
library(RColorBrewer)
library(paletteer)

#set.seed(20230324)
#set.seed(20230327)
set.seed(20230342)
#set.seed(20230317)

mat <- intensity_mat_hq_wo_zeros
clusters <- factor(out$communities)
tissue <- spe$tissue

n <- length(unique(clusters))
col_vector_433 = colors()[grep('gr(a|e)y', grDevices::colors(), invert = T)]
col_clusters <- setNames(sample(col_vector_433, n), unique(clusters))

col_marker = colorRamp2(c(0, 2.5, 0.5), c("blue", "white", "red"))
col_tissue = c("PT" = "deepskyblue", "BM" = "orange")

#Get same number of PT and BM cells
pt_index <- which(tissue=="PT")
bm_index <- which(tissue=="BM")
rand_cells_pt <- sample(seq_len(length(pt_index)), 4000)
rand_cells_bm <- sample(seq_len(length(bm_index)), 4000)
pt_index_rand <- pt_index[rand_cells_pt]
bm_index_rand <- bm_index[rand_cells_bm]
rand_cells <- c(pt_index_rand, bm_index_rand)

#rand_cells <- sample(seq_len(ncol(mat)), 2000)

column_ha = HeatmapAnnotation(cluster = clusters[rand_cells],
                              tissue = tissue[rand_cells],
                              col=list(cluster=col_clusters, 
                                       tissue=col_tissue 
                                       ), 
                              show_legend=T)


Heatmap(mat[,rand_cells], 
        column_title = "Single-cell heatmap with phenograph clusters (1-px mean)",
        top_annotation = column_ha, 
        cluster_columns=F,
        column_order= order(clusters[rand_cells], decreasing=F),
        show_column_dend = F,
        show_column_names=F,
        col=viridis(100)
        )
```