---
title: "rarity_test"
output: html_document
date: '2023-04-26'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(ggplot2)
```

In this RMD file, we will perform clustering on the curated data (after excluding cells that swam off between IF and IMC and segmentation artifacts). First, we will read our SPE.

```{r read-data}

path <- "/mnt/Multimodal_Imaging_Daria/_Data_Analysis/_tmp_daria/Image_analysis/20230112_DIMR_Ilastik/20230312_sc_data_combined"

spe <- readRDS(file.path(path,"data/spe_1px_wo-ghosts.rds"))

```

Then we will generate matrices from morphological and intensity features alone and combined. 

```{r generate-dataframes, message=FALSE, fig.height=7}
library(bruceR)

# morph_names <- c("area", 
#                  "area_convex", 
#                  "area_filled", 
#                  "axis_major_length", 
#                  "axis_minor_length", 
#                  "eccentricity", 
#                  "equivalent_diameter_area", 
#                  "perimeter", 
#                  "solidity", 
#                  "assymetry", 
#                  "concavity", 
#                  "fill", 
#                  "aspect_ratio-0", 
#                  "perimeter_ratio-0")

morph_names <- c("area",
                 "eccentricity"
)

robustScaler <- function(data, upper, lower) {
  scaled_data <- (data-median(data))/(quantile(data, probs = c(upper)) - quantile(data, probs = c(lower)))
  return(scaled_data)
}

zscore_Scaler <- function(data) {
  scaled_data <- (data-mean(data))/sd(data)
  return(scaled_data)
}

clipScaler <- function(data, upper, lower, min, max) {
  data[data > quantile(data, probs = c(upper))] <- quantile(data, probs = c(upper))
  data[data < quantile(data, probs = c(lower))] <- quantile(data, probs = c(lower))
  scaled_data <- RESCALE(data, to=min:max)
  return(scaled_data)
}
intensity_mat <- assay(spe, "exprs")
colnames(intensity_mat) <- spe$sample
intensity_mat <- intensity_mat[!grepl("mean-80", rownames(intensity_mat)),]


intensity_mat_hq <- intensity_mat[!grepl("CD274|IDO|DAPI|2IF_GD2|3IF_CD56|DNA1|H4K12Ac|H3K9Ac|CXCR2|PNMT|Fibro", rownames(intensity_mat)),]
intensity_mat_lq <- intensity_mat[grepl("CD274|IDO|DAPI|2IF_GD2|3IF_CD56|DNA1|H4K12Ac|H3K9Ac|CXCR2|PNMT|Fibro", rownames(intensity_mat)),]
#intensity_mat_hq <- intensity_mat[!grepl("DAPI|2IF_GD2|3IF_CD56|DNA1|H4K12Ac|H3K9Ac|PNMT", rownames(intensity_mat)),]
#intensity_mat_lq <- intensity_mat[grepl("DAPI|2IF_GD2|3IF_CD56|DNA1|H4K12Ac|H3K9Ac|PNMT", rownames(intensity_mat)),]

#Make sure to exclude cells that are zero for all markers as PG won't work otherwise
non_zeros <- as.logical(colSums(intensity_mat_hq)!=0)

intensity_mat_hq_wo_zeros <- intensity_mat_hq[,non_zeros]
intensity_mat_lq_wo_zeros <- intensity_mat_lq[,non_zeros]
spe <- spe[,non_zeros]


cell_id <- colnames(spe)
tissue <- spe$tissue

morph_mat <- as.matrix(t(colData(spe)[,morph_names]))
rownames(morph_mat) <- gsub('-0','',rownames(morph_mat))
#morph_mat_scaled <- t(apply(morph_mat, 1, RESCALE, to=0:1))
#morph_mat_scaled <- t(apply(morph_mat, 1, robustScaler, upper=0.95, lower=0.05))
#morph_mat_scaled <- t(apply(morph_mat, 1, zscore_Scaler))
#morph_mat_scaled <- t(apply(morph_mat, 1, clipScaler, upper=0.9995, lower=0.0005, min=0, max=mean(apply(intensity_mat_scaled,1,max))))
morph_mat_scaled <- t(apply(morph_mat, 1, clipScaler, upper=0.995, lower=0.005, min=0, max=1))
colnames(morph_mat_scaled) <- spe$sample


features_mat <- rbind(intensity_mat_hq, morph_mat_scaled)

```

In a next step, we will plot a single-cell heatmap using complex heatmap. 

```{r single-cell-heatmap, message=FALSE, fig.height=20}
library(ComplexHeatmap)
library(viridis)
library(circlize)
library(RColorBrewer)
library(paletteer)

mat <- intensity_mat_hq

qual_col_pals = brewer.pal.info[brewer.pal.info$category == 'qual',]
col_vector_74 = unlist(mapply(brewer.pal, qual_col_pals$maxcolors, rownames(qual_col_pals)))
col_vector_433 = colors()[grep('gr(a|e)y', grDevices::colors(), invert = T)]

col_sample = setNames(sample(col_vector_433, length(unique(spe$sample))), unique(spe$sample))

set.seed(221228)
rand_cells <- sample(seq_len(ncol(mat)), 2000)

column_ha = HeatmapAnnotation(sample = colnames(mat[,rand_cells]), col=list(sample=col_sample), show_legend=F)
Heatmap(mat[,rand_cells], 
        column_title = "Single-cell heatmap",
        top_annotation = column_ha, 
        show_column_dend = F,
        show_column_names=F,
        col=viridis(100)
        )
```


Now that we have the single-cell matrix, we will test rarity for clustering. 

```{r phenograph-clustering}
library(tidyverse)
library(reticulate)
use_python("/usr/bin/python3")

source("R/core.R")
source("R/helpers.R")

Y <- t(intensity_mat_hq_wo_zeros)

m <- Rarity_fit(Y, n_iter = 20000L, batch_size = 512)

plot_binary_signatures(m, n_cells = 30)

plot_heatmap_with_observations(Y, m, n_cells = 30)

```