---
title: "Differential abundance testing with limma-voom"
output: html_document
date: '2023-08-19'
---

```{r setup, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(echo = TRUE, out.width = "120%")
```

## Load data

First, we will find paired samples. 

```{r selection-paired-sample, message=FALSE}
library(SpatialExperiment)
library(CATALYST)
library(cowplot)
library(ggplot2)
library(diffcyt)
library(flowCore)
library(readxl)
library(scater)
library(dplyr)
library(tidyr)

path <- "/mnt/Multimodal_Imaging_Daria/PIPELINE/NEWEST_RESULTS"
spe <- readRDS(file.path(path,"results/Manuscript/spe_clustered.rds"))

clusternames_ordered <- read.csv(file.path(path, "results/Manuscript/celltypes_ordered_2.csv"), sep=";", colClasses = c("order" = "character"))
codes <- setNames(paste0(clusternames_ordered$order, "_", clusternames_ordered$celltype), clusternames_ordered$celltype)
spe$celltype_ordered <- recode(spe$celltype, !!!codes)

metadata(spe)$color_vectors$col_celltype_ordered <- metadata(spe)$color_vectors$col_celltype
names(metadata(spe)$color_vectors$col_celltype_ordered) <- recode(names(metadata(spe)$color_vectors$col_celltype), !!!codes)
```

## Subset dataset

We will first read in the previously generated `SpatialExperiment` object and transform it into a catalyst compatible object.

```{r read-data-pheno, message=FALSE}

keep_rows <- rowData(spe)$use_channel 
keep_cols <- spe$tissue=="BM" & spe$celltype != "other" & (spe$timepoint=="DE"|spe$timepoint=="") & spe$metacluster!="tumor" & spe$metacluster!="mesenchymal" 

sce <- SingleCellExperiment(assays=list(counts=assay(spe, "counts")[keep_rows, keep_cols], exprs=assay(spe, "counts")[keep_rows, keep_cols]))

sce$solidity_clipped <- spe[keep_rows, keep_cols]$solidity_clipped

sce$control <- factor(spe[keep_rows, keep_cols]$control)
sce$cluster_id <- factor(spe[keep_rows, keep_cols]$celltype_ordered)
sce$patient <- factor(spe[keep_rows, keep_cols]$study_id)
sce$tissue <- factor(spe[keep_rows, keep_cols]$tissue)
sce$metacluster <- factor(spe[keep_rows, keep_cols]$metacluster)
sce$tmp <- factor(spe[keep_rows, keep_cols]$tmp)

sce$mna <- factor(spe[keep_rows, keep_cols]$MYCN_amp)
mna <- as.vector(sce$mna)
mna <- replace(x = mna, list =  mna %in% c(1), values =  'MNA')
mna <- replace(x = mna, list =  mna %in% c(3), values =  'het')
mna <- replace(x = mna, list =  mna %in% c(0), values =  'nMNA')
mna <- replace(x = mna, list =  mna %in% c(9), values =  'unknown')
sce$mna <- as.factor(mna)

sce$progress <- factor(spe[keep_rows, keep_cols]$progression)
progress <- as.vector(sce$progress)
progress <- replace(x = progress, list =  progress==0, values =  'NoProg')
progress <- replace(x = progress, list =  progress==1, values =  'Prog')
sce$progress <- as.factor(progress)

sce$timepoint <- spe[keep_rows, keep_cols]$timepoint
sce$timepoint <- factor(sce$timepoint)
sce$tissue_crtl <- paste0(sce$tissue, "_", sce$control)
sce$condition <- sce$control

sce$sample_id <- factor(paste0(sce$progress, "_", sce$mna, "_", spe[keep_rows, keep_cols]$study_id, "_", spe[keep_rows, keep_cols]$fm_id, "_", sce$tissue))

# Add celltype information to metadata
metadata(sce)$cluster_codes <- data.frame(celltype = factor(spe[keep_rows, keep_cols]$celltype_ordered))

# Add metadata
metadata(sce)$experiment_info <- as.data.frame(colData(sce)) %>% group_by(sample_id,patient, mna, condition, progress) %>% 
    summarise(n_cells=n(),.groups = 'drop') %>%
    as.data.frame()

# Define cell_type_markers 
type_markers <- c("MPO_Y89_mean", "CD44_In115_mean", "CD11b_Nd142_mean", "HLA-DR_Nd143_mean", "PRPH_Nd144_mean", "HLA-ABC_Sm147_mean", "CD20_Nd148_mean", "LUM_Sm149_mean", "CD11c_Nd150_mean",
                  "CD24_Eu151_mean", "CD3_Sm152_mean", "CD45_Eu153_mean", "CD8a_Sm154_mean", "GD2_Gd155_mean", "CD34_Gd156_mean", "CD10_Gd158_mean", "CXCR4_Tb159_mean","SOX10_Dy162_mean",
                  "FOXP3_Dy163_mean", "CHGA_Dy164_mean", "GATA3_Er168_mean", "CD56_Er170_mean", "CD4_Yb171_mean", "ELAVL4_Yb174_mean", "CD14_Lu175_mean", "Vimentin_Pt196_mean", "CD15_Bi209_mean")

# Define cell_state_markers 
state_markers <- c("CD274_Gd160_mean", "S100B_Dy161_mean", "CD279_Ho165_mean", "Ki-67_Tm169_mean", "GZMB_Yb173_mean")

# Add to spe
rowData(sce)$marker_class <- ifelse(rownames(sce) %in% type_markers, "type",
                                    ifelse(rownames(sce) %in% state_markers, "state", 
                                    "other"))

#Add dim reduction

reducedDim(sce, "UMAP") <- reducedDim(spe, "UMAP")[keep_cols,]
```

## Plot overview of cohort

Differential analysis of cell population abundance compares the proportions of cell types across experimental conditions and aims to highlight populations that are present at different ratios. First, we calculate two tables: one that contains cell counts for each sample and population and one with the corresponding  proportions of cell types by sample.  The proportions are used only for plotting, since the statistical modeling takes the cell counts by cluster and sample as input.

For each sample, we plot its PBMC cell type composition represented with colored bars, where the size of a given stripe reflects the proportion of the corresponding cell type in a given sample (Figure \@ref(fig:diff-freqs-plot-props-barplot)).

```{r diff-freqs-plot-props-barplot, fig.cap = "Relative abundance of the 8 PBMC populations in each sample (x-axis), in the PBMC dataset, represented with a barplot. The 8 cell populations are a result of manual merging of the 20 FlowSOM metaclusters", fig.width=20, fig.height=10}

set.seed(20231103)
p <- plotAbundances(sce, k = "celltype", by = "sample", col_clust=T, k_pal=metadata(spe)$color_vectors$col_celltype_ordered, 
                    linkage="ward.D", distance="manhattan") 
#Best linkage: ward.D
#Best distance: manhattan

condition <- c("yes", "no")
clustered_samples <- list()
for (i in seq_along(condition)){
  clustered_samples <- c(clustered_samples, sapply(strsplit(levels(droplevels(p$data[p$data$condition==condition[i],]$sample_id)), "_"), function(x) paste(x[4], x[5], sep = "_")))
}
clustered_samples <- as.character(clustered_samples)

setEPS()
postscript(file.path(path, "results/Manuscript/4_Figure/R/barplot_BM_crtlvsDE_samples.eps"), height=12, width=20)
p
dev.off()
```

As a next step, we will compute the heterogeneity of samples based on the ME and tumor cell composition.
```{r heterogeneity, message=FALSE, fig.width=7, fig.height=7}
library(BioQC)
library(hrbrthemes)

proportions <- plotAbundances(sce, k = "celltype", by = "sample")
prop <- proportions$data %>%
  select(-condition) %>%
  pivot_wider(names_from = cluster_id, values_from = Freq)

prop <- data.frame(prop)
rownames(prop) <- prop$sample_id
prop <- prop %>% select(-sample_id)

prop <- t(prop)
rownames(prop) <- levels(sce$cluster_id)

rownames(prop) <- levels(sce$cluster_id)
colnames(prop) <- sapply(strsplit(colnames(prop), "_"), function(x) paste(x[4], x[5], sep = "_"))

prop <- prop[,clustered_samples]
het <- apply(prop,2, entropy)
```

Next, we will plot metadata on top of the clustered barplot. First, we will define the color code for the metadata. Find all built-in R color names [here](https://www.farb-tabelle.de/de/farbtabelle.htm). The [paletteer package](https://r-charts.com/color-palettes/) offers a wide range of color palettes.  

```{r define-color-code, message=FALSE}
library(ComplexHeatmap)
library(RColorBrewer)
library(paletteer)
library(bruceR)
library(circlize)
library(viridis)

meta_path <- file.path(path, '20231128_metadata_complete.csv')

qual_col_pals = brewer.pal.info[brewer.pal.info$category == 'qual',]
col_vector_74 = unlist(mapply(brewer.pal, qual_col_pals$maxcolors, rownames(qual_col_pals)))

col_vector_433 = colors()[grep('gr(a|e)y', grDevices::colors(), invert = T)]

meta_all = read.csv(meta_path)
meta_all <- meta_all[order(meta_all$Study_ID),] #order by study-id

exclude <- replace(x = meta_all$exclude, list =  !meta_all$exclude %in% c('no', 'maybe'), values =  'partially')

#DE_year <- sub("-.*", "", meta_all$DE_date)

col_sample = setNames(sample(col_vector_433, length(unique(meta_all$Sample))), unique(meta_all$Sample))
col_puncture_side = setNames(brewer.pal(length(unique(meta_all$Puncture_side)), 'Set3'), unique(meta_all$Puncture_side))
col_staining_batch = setNames(viridis(length(unique(meta_all$Staining_date))), sort(unique(meta_all$Staining_date)))
col_stainer = setNames(c('#a6611a', '#dfc27d', '#80cdc1', '#018571'), unique(meta_all$Stainer))
col_imc_batch = setNames(viridis(length(unique(meta_all$IMC_date))), sort(unique(meta_all$IMC_date)))
col_imc_score = colorRamp2(c(min(meta_all$IMC_score, na.rm = TRUE), max(meta_all$IMC_score, na.rm = TRUE)/2, max(meta_all$IMC_score, na.rm = TRUE)), viridis(3)) 
col_study_id = setNames(sample(col_vector_433, length(unique(meta_all$Study_ID))), unique(meta_all$Study_ID))
col_DE_year = setNames(viridis(length(unique(meta_all$DE_date))), sort(unique(meta_all$DE_date)))
col_timepoint = setNames(c('#7fc97f', '#beaed4', '#386cb0', '#ffff99', '#386cb0'), unique(meta_all$Timepoint)) #'#fdc086'
col_aberrations = setNames(c('#a6611a','#018571','#80cdc1','#f5f5f5'), c(0,1,2,9))
col_stages = setNames(c('#a6cee3', '#1f78b4', '#b2df8a', '#33a02c', '#fb9a99', '#e31a1c'), unique(meta_all$stage4))
col_efs_dur = colorRamp2(c(min(meta_all$efs_dur, na.rm = TRUE), max(meta_all$efs_dur, na.rm = TRUE)/2, max(meta_all$efs_dur, na.rm = TRUE)), viridis(3))
col_tissue = setNames(c('aquamarine4','brown'), unique(meta_all$Tissue))
col_control = setNames(c('gray77', 'mediumseagreen'), unique(meta_all$control))
col_exclude = setNames(c('#018571','#a6611a','#80cdc1'), unique(exclude))
col_progress = setNames(c('gray77', 'palegreen3', 'firebrick1'), c('n.a.', 0, 1))
#col_age_diag = colorRamp2(c(min(meta_all$age_at_diag, na.rm = TRUE), max(meta_all$age_at_diag, na.rm = TRUE)/2, max(meta_all$age_at_diag, na.rm = TRUE)), viridis(3))
col_age_diag = colorRamp2(c(min(meta_all$age_at_diag, na.rm = TRUE), (quantile(meta_all$age_at_diag, 0.95, na.rm = TRUE)-min(meta_all$age_at_diag, na.rm = TRUE))/2, quantile(meta_all$age_at_diag,0.95, na.rm = TRUE)), viridis(3))
col_therapy = setNames(c('#a6611a', '#018571'), c('RAPID COJEC', 'MOD. N7'))

col_het = colorRamp2(c(min(het, na.rm = TRUE), (max(het, na.rm = TRUE)-min(het, na.rm = TRUE))/2, max(het, na.rm = TRUE)), viridis(3))
```

Now, we will plot an overview of the cohort.

```{r overview-cohort, message=FALSE, fig.height=25}

rownames(meta_all) <- paste0(meta_all$Sample, "_", meta_all$Tissue)
meta_all <- meta_all[clustered_samples,]

#reading patient groups from Figure 5
patient_groups <- read.csv(file.path(path, "results/Manuscript/5_Figure/R/cellular_communities.csv"))
rownames(patient_groups) <- sapply(strsplit(patient_groups$X, "_"), function(x) paste(x[4], x[5], sep = "_"))
meta_all$patient_groups <- "na"
meta_all[rownames(patient_groups),]$patient_groups <- patient_groups$community_cluster
col_patient_group = setNames(c("slategray1", "slategray", "slategray3", "black"),c("1", "2", "3", "na"))

setEPS()
postscript(file.path(path, "results/Manuscript/4_Figure/R/annotationbar_BM_crtlvsDE_patientgroups.eps"), height=30, width=30)

columnAnnotation(control=meta_all$control, col=list(control=col_control)) %v%
columnAnnotation(sample=meta_all$Sample, col=list(sample=col_sample), annotation_legend_param = list(sample = list(grid_height=unit(0.05, "cm"), labels_gp = gpar(fontsize = 5)))) %v%
#columnAnnotation(exclusion=exclude, col=list(exclusion=col_exclude)) %v%
columnAnnotation(imc_score=meta_all$IMC_score, col=list(imc_score=col_imc_score))  %v%
columnAnnotation(staining_batch=meta_all$Staining_date, col=list(staining_batch=col_staining_batch)) %v%
columnAnnotation(stainer=meta_all$Stainer, col=list(stainer=col_stainer)) %v%
columnAnnotation(imc_batch=meta_all$IMC_date, col=list(imc_batch=col_imc_batch)) %v%
columnAnnotation(D_year=meta_all$DE_date, col=list(D_year=col_DE_year))  %v%
columnAnnotation(study_id=meta_all$Study_ID, col=list(study_id=col_study_id))  %v%
columnAnnotation(timepoint=meta_all$Timepoint, col=list(timepoint=col_timepoint))  %v%
columnAnnotation(tissue=meta_all$Tissue, col=list(tissue=col_tissue)) %v%
columnAnnotation(stage=meta_all$stage4, col=list(stage=col_stages)) %v%
columnAnnotation(progression=meta_all$efs, col=list(progression=col_progress))  %v%
columnAnnotation(efs_dur=meta_all$efs_dur, col=list(efs_dur=col_efs_dur)) %v%

columnAnnotation(X1p_loss=meta_all$X1p_loss, col=list(X1p_loss=col_aberrations), show_legend=FALSE) %v%
columnAnnotation(X1q_gain=meta_all$X1q_gain, col=list(X1q_gain=col_aberrations), show_legend=FALSE) %v%
columnAnnotation(X1q_loss=meta_all$X1q_loss, col=list(X1q_loss=col_aberrations), show_legend=FALSE) %v%
columnAnnotation(X2p_gain=meta_all$X2p_gain, col=list(X2p_gain=col_aberrations), show_legend=FALSE) %v%
columnAnnotation(X3p_loss=meta_all$X3p_loss, col=list(X3p_loss=col_aberrations), show_legend=FALSE) %v%
columnAnnotation(X3q_loss=meta_all$X3q_loss, col=list(X3q_loss=col_aberrations), show_legend=FALSE) %v%
columnAnnotation(X4p_loss=meta_all$X4p_loss, col=list(X4p_loss=col_aberrations), show_legend=FALSE) %v%
columnAnnotation(X6q_loss=meta_all$X6q_loss, col=list(X6q_loss=col_aberrations), show_legend=FALSE) %v%
columnAnnotation(X7q_gain=meta_all$X7q_gain, col=list(X7q_gain=col_aberrations), show_legend=FALSE) %v%
columnAnnotation(PTPRD_del=meta_all$PTPRD_del, col=list(PTPRD_del=col_aberrations), show_legend=FALSE) %v%
columnAnnotation(CDKN2AB_del=meta_all$CDKN2AB_del, col=list(CDKN2AB_del=col_aberrations), show_legend=FALSE) %v%
columnAnnotation(X11q_loss=meta_all$X11q_loss, col=list(X11q_loss=col_aberrations), show_legend=FALSE) %v%
columnAnnotation(X12q_gain=meta_all$X12q_gain, col=list(X12q_gain=col_aberrations), show_legend=FALSE) %v%
columnAnnotation(X14q_loss=meta_all$X14q_loss, col=list(X14q_loss=col_aberrations), show_legend=FALSE) %v%
columnAnnotation(X17p_loss=meta_all$X17p_loss, col=list(X17p_loss=col_aberrations), show_legend=FALSE) %v%
columnAnnotation(X17q_gain=meta_all$X17q_gain, col=list(X17q_gain=col_aberrations), show_legend=FALSE) %v%
columnAnnotation(X19p_loss=meta_all$X19p_loss, col=list(X19p_loss=col_aberrations), show_legend=FALSE) %v%
columnAnnotation(X19q_loss=meta_all$X19q_loss, col=list(X19q_loss=col_aberrations), show_legend=FALSE) %v%
columnAnnotation(X22q_loss=meta_all$X22q_loss, col=list(X22q_loss=col_aberrations), show_legend=FALSE) %v%
columnAnnotation(CDKN2AB_del=meta_all$CDKN2AB_del, col=list(CDKN2AB_del=col_aberrations), show_legend=FALSE) %v%
columnAnnotation(X11q_loss=meta_all$X11q_loss, col=list(X11q_loss=col_aberrations), show_legend=FALSE) %v%
columnAnnotation(X12q_gain=meta_all$X12q_gain, col=list(X12q_gain=col_aberrations), show_legend=FALSE) %v%
columnAnnotation(X14q_loss=meta_all$X14q_loss, col=list(X14q_loss=col_aberrations), show_legend=FALSE) %v%
columnAnnotation(MYCN_amp=meta_all$MYCN_amp, col=list(MYCN_amp=col_aberrations)) %v% #annotation_legend_param = list(MYCN_amp = list(title='aberration', 
                                                                                                              #labels = c('no','yes','het'))))  %v%
                                                                                                              #labels = c('no', 'yes', 'het', 'NA'))))  %v%
columnAnnotation(TERT_rear=meta_all$TERT_rear, col=list(TERT_rear=col_aberrations), show_legend=FALSE) %v%
columnAnnotation(ATRX_del=meta_all$ATRX_del, col=list(ATRX_del=col_aberrations), show_legend=FALSE) %v% 
columnAnnotation(ALK_amp=meta_all$ALK_amp, col=list(ALK_amp=col_aberrations), show_legend=FALSE) %v%
columnAnnotation(ALK_mut=meta_all$ALK_mut, col=list(ALK_mut=col_aberrations), show_legend=FALSE) %v%

columnAnnotation(age=meta_all$age_at_diag, col=list(age=col_age_diag)) %v%
columnAnnotation(therapy=meta_all$therapy, col=list(therapy=col_therapy)) %v%
columnAnnotation(het=het, col=list(het=col_het)) %v%
columnAnnotation(patient_group=meta_all$patient_groups, col=list(patient_group=col_patient_group), show_legend=FALSE)


dev.off()

#absolute cellcount
sce$sample_id <- factor(sapply(strsplit(as.character(sce$sample_id), "_"), function(x) paste(x[4], x[5], sep = "_")))
sce <- sce[, sce$sample_id%in%clustered_samples]
sce$sample_id <- droplevels(sce$sample_id)
sce$sample_id <- factor(sce$sample_id, levels=clustered_samples)

#absolute cellcount
setEPS()
postscript(file.path(path, paste0("results/Manuscript/4_Figure/R/BM_crtlvsDE_annotationbar_ncells.eps")), height=30, width=30)
draw(anno_barplot(as.vector(unname(table(sce$sample_id))), add_numbers=T, height = unit(1, "cm")))
dev.off() 

#tumor cell proportion
norm_minmax <- function(x){(x- min(x)) /(max(x)-min(x))}

robust_scalar<- function(x){(x- median(x)) /(quantile(x,probs = .75)-quantile(x,probs = .25))}

clipper <- function(data, lower, upper) {
  data[data > quantile(data, probs = c(upper))] <- quantile(data, probs = c(upper))
  data[data < quantile(data, probs = c(lower))] <- quantile(data, probs = c(lower))
  clipped_data <- norm_minmax(data)
  return(clipped_data)}
```

Subsequently, we will calculate the mean roundness of the granulocyte cluster per sample.
```{r granulocyte-roundness, message=FALSE}
library("ggpubr")
library(gtools)

sce$sample <- factor(sapply(strsplit(as.character(sce$sample_id), "_"), function(x) paste(x[4], x[5], sep = "_")))
df <- aggregate(sce[,sce$metacluster=="granulocyte"]$solidity_clipped, list(sce[,sce$metacluster=="granulocyte"]$sample), mean)
rownames(df) <- df[,1]
df <- df[-c(1)]
colnames (df) <- "y"

rownames(meta_all) <- paste0(meta_all$Sample, "_", meta_all$Tissue)
df$condition <- meta_all[rownames(df),]$control

stats_df <- compare_means(y ~ condition,  
                      #p.adjust.method = "BH", 
                      method='wilcox.test', 
                      data = df)
  
stats_df$p_adj_signif <- stars.pval(stats_df$p.adj)
stats_df$p_adj_signif[stats_df$p_adj_signif==" "] <- "ns"
stats_df$p_long <- paste0(stats_df$p.adj, " (", stats_df$p_adj_signif, ")")

p <- ggboxplot(df, x = "condition", y = "y",
                color = "condition", palette =c("slategray1", "slategray"),
                add.params = list(size = 3),
                add = "jitter")+ 
                stat_pvalue_manual(stats_df, label = "p_long", y.position = max(df$y), step.increase = 0.2,) +
                theme(aspect.ratio = 1.5, legend.position="none") 

setEPS()
postscript(file.path(path, paste0("results/Manuscript/4_Figure/R/BM_crtl_inf_roundness.eps")), height=30, width=5)
p
dev.off()


```

## Differential cell population abundance
Since the diffcyt wrapper does not work, we will test the test-functions from the diffcyt package. 

```{r test-diffcyt, message=FALSE}

#Create list of samples as input to diffcyt methods
df_list <- lapply(unique(sce$sample_id), function(x){as.data.frame(t(counts(sce[,sce$sample_id==x])))})
names(df_list) <- unique(sce$sample_id)

#Create experiment info
experiment_info <- metadata(sce)$experiment_info
experiment_info <- experiment_info[match(names(df_list),experiment_info$sample_id),]

#Create marker info
# Add rowData
channel_name <- sapply(strsplit(rownames(rowData(spe)[keep_rows,]), "_"), "[[", 2)
marker_name <- sapply(strsplit(rownames(rowData(spe)[keep_rows,]), "_"), "[[", 1)
marker_class <-ifelse(rownames(spe)[keep_rows] %in% type_markers, "type",
                                    ifelse(rownames(spe)[keep_rows] %in% state_markers, "state", 
                                    "other"))
row_data <- data.frame(cbind(channel_name, marker_name), row.names=marker_name)
row_data$marker_class <- as.factor(marker_class)

marker_info <-  row_data

#Create summarized experiment - not working properly for some reason - cannot be fed to function diffcyt
d_se <- prepareData(df_list, experiment_info, marker_info)
rowData(d_se)$cluster_id <- sce[,rownames(d_se)]$cluster_id 
metadata(d_se)$cluster_codes <- metadata(sce)$cluster_codes

# Transform data
d_se <- transformData(d_se)

# Calculate counts
d_counts <- calcCounts(d_se)
```

```{r test-diffcyt, message=FALSE}
# Create design matrix
#design <- createDesignMatrix(experiment_info, cols_design = c("condition", "patient")) #paired
design <- createDesignMatrix(experiment_info, cols_design = "condition")

# Create contrast matrix
contrast <- createContrast(c(0, 1)) # unpaired
#contrast <- createContrast(c(0, 1, rep(0, length(unique(experiment_info$patient))-1))) #paired
nrow(contrast) == ncol(design)

# Test for differential abundance (DA) of clusters
res_DA <- testDA_voom(d_counts,design, contrast, 
                       min_cells=3, 
                       #min_samples=round(dim(d_counts)[2]*0.05)#has to be used when working with ME 
                       min_samples=round(dim(d_counts)[2]*0.2)
                       )


FDR_cutoff = 0.1
rowData(res_DA)[rowData(res_DA)$p_adj<FDR_cutoff,]
res<- data.frame(rowData(res_DA))
```

Next, we will plot these significance values onto a boxplot according to [this](https://www.datanovia.com/en/blog/ggpubr-how-to-add-p-values-generated-elsewhere-to-a-ggplot/).
```{r, fig.width=22, fig.height=18}
library(tidyr)
library(ggpubr)
library(rstatix)
library(gridExtra)
library(gtools)

p <- plotAbundances(sce, k = "celltype", by = "cluster_id", shape_by = "sample_id")

comparisons <- c("PTvsBM")

plot_list_values <- list()
plot_list_stars <- list()
plot_list_fc <- list()
for(i in 1:length(unique(p$data$cluster_id))){
  cluster_name <- unique(p$data$cluster_id)[i]
  proportions <- data.frame(p$data[p$data$cluster_id == cluster_name, c("sample_id", "condition", "Freq")])
  
  p_adj <- c()
  group1 <- c()
  group2 <- c()
  logFC <- c()
  
  for(j in 1:1){
  stats <- res
  p_adj <- c(p_adj, round(stats[rownames(stats)==droplevels(cluster_name), "p_adj"],3))
  logFC <- c(logFC, -1*round(stats[rownames(stats)==droplevels(cluster_name), "logFC"],2))
  group1 <- c(group1, unlist(strsplit(comparisons[j], "vs"))[1])
  group2 <- c(group2, unlist(strsplit(comparisons[j], "vs"))[2])
  }

  stats_df <- data.frame(p_adj=p_adj, group1=group1, group2=group2, logFC=logFC)
  stats_df$p_adj_signif <- stars.pval(stats_df$p_adj)
  stats_df$p_adj_signif[stats_df$p_adj_signif==" "] <- "ns"
  stats_df$p_fc <- paste0(stats_df$logFC, " (", stats_df$p_adj, ")", " (", stats_df$p_adj_signif, ")")

  color <- unname(metadata(spe)$color_vectors$col_celltype_ordered[as.character(cluster_name)]) 
  
  plot_list_fc[[i]] <- ggboxplot(proportions, x = "condition", y = "Freq", order=c("PT", "BM"),
              color = "condition", palette =c(color, color), 
              #lwd=1.6, 
              add.params = list(size = 3),
              add = "jitter", 
              title=cluster_name, show.legend = F) +
      stat_pvalue_manual(
        stats_df,
        y.position = max(proportions$Freq),
        step.increase = 0.4,
        label = "p_fc"
        )  +
      theme(aspect.ratio = 1.5, legend.position="none") +
      scale_y_continuous(expand = expansion(mult = c(0.05, 0.15)))
}

setEPS()
postscript(file.path(path, "results/Manuscript/4_Figure/R/PTvsBM_tumor_boxplot_voom.eps"), height=26, width=20)
grid.arrange(grobs = plot_list_fc, ncol = 6)
   dev.off()
```

First, we want to show the ME composition in PT, crtl and DE BM samples. 

```{r}
p <- plotAbundances(sce, k = "celltype", by = "cluster_id", shape_by = "sample_id")

df_normalized <- p$data %>%
  group_by(cluster_id, condition) %>%
  summarise(normalized_count = mean(Freq, na.rm = TRUE)) %>%
  ungroup()


df_normalized$cluster_id <- factor(df_normalized$cluster_id)

setEPS()
postscript(file.path(path, "results/Manuscript/4_Figure/R/PT_crtl_BM_bar_metacluster.eps"), height=4, width=8)
ggplot(df_normalized, aes(fill=cluster_id, y=normalized_count, x=condition)) + 
  geom_bar(
    stat="identity",
    show.legend = T) + 
  scale_fill_manual(values = metadata(spe)$color_vectors$col_metacluster) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

dev.off()

```


