---
title: "Generation of spillover matrix"
output: html_document
date: '2023-05-19'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Read in the data {#read-data}

```{r, message=FALSE}
library(imcRtools)
library(cytomapper)
library(BiocParallel)
```


## Generate the spillover matrix

In the first step, we will generate a spillover matrix based on the single-metal
spots and save it for later use.

### Read in the data

Here, we will read in the individual .txt files into a `SingleCellExperiment`
object. This object can be used directly by the `CATALYST` package to estimate
the spillover.

For this to work, the .txt file names need to contain the spotted metal isotope
name. By default, the first occurrence of the isotope in the format `(mt)(mass)`
(e.g. `Sm152` for Samarium isotope with the atomic mass 152) will be used as
spot identifier. Alternatively, a named list of already read-in pixel intensities
can be provided. For more information, please refer to the man page `?readSCEfromTXT`.

For further downstream analysis, we will asinh-transform the data using a
cofactor of 5; a common transformation for CyTOF data [@Bendall2011].

```{r read-txts, message=FALSE}
library(imcRtools)

path_to_spillover_data <- "/mnt/Multimodal_Imaging_Daria/20220518_DL_SpilloverCorrection/Extra"

# Create SingleCellExperiment from .txt files
sce <- readSCEfromTXT(path_to_spillover_data) 
assay(sce, "exprs") <- asinh(counts(sce)/5)

sce1 <- sce
```


### Optional pixel binning {#pixel_binning}

In cases where median pixel intensities are low (< 200 counts), consecutive
pixels can be summed to increase the robustness of the spillover estimation.
The `imcRtools` package provides the `binAcrossPixels` function,
which performs aggregation for each channel across `bin_size` consecutive pixels 
per spotted metal.

```{r binning, message=FALSE}
# Define grouping
bin_size = 10

sce2 <- binAcrossPixels(sce, bin_size = bin_size)
assay(sce2, "exprs") <- asinh(counts(sce2)/5)
```

### Filtering incorrectly assigned pixels

The following step uses functions provided by the `CATALYST` package to
"debarcode" the pixels. Based on the intensity distribution of all channel,
pixels are assigned to their corresponding barcode; here this is the already
known metal spot. This procedure serves the purpose to identify pixels that
cannot be robustly assigned to the spotted metal. Pixels of such kind can be
regarded as "noisy", "background" or "artefacts" that should be removed prior to
spillover estimation.

We will also need to specify which channels were spotted (argument `bc_key`).
This information is directly contained in the `colData(sce)` slot.
To facilitate visualization, we will order the `bc_key` by mass.

The general workflow for pixel debarcoding is as follows:

1. assign a preliminary metal mass to each pixel
2. for each pixel, estimate a cutoff parameter for the distance between 
positive and negative pixel sets
3. apply the estimated cutoffs to identify truly positive pixels

```{r debarcoding, message=FALSE}
library(CATALYST)

bc_key <- as.numeric(unique(sce1$sample_mass))
bc_key <- bc_key[order(bc_key)]

sce1 <- assignPrelim(sce1, bc_key = bc_key)
sce1 <- estCutoffs(sce1)
sce1 <- applyCutoffs(sce1)

bc_key <- as.numeric(unique(sce2$sample_mass))
bc_key <- bc_key[order(bc_key)]

sce2 <- assignPrelim(sce2, bc_key = bc_key)
sce2 <- estCutoffs(sce2)
sce2 <- applyCutoffs(sce2)
```

The obtained `SingleCellExperiment` now contains the additional `bc_id` entry.
For each pixel, this vector indicates the assigned mass (e.g. `161`) or
`0`, meaning unassigned. 

However, in cases where incorrect assignment occurred or where few pixels were
measured for some spots, the `imcRtools` package exports a simple helper
function to exclude pixels based on these criteria:

```{r pixel-filtering}
sce1 <- filterPixels(sce1, minevents = 40, correct_pixels = TRUE)
sce2 <- filterPixels(sce2, minevents = 40, correct_pixels = TRUE)
```

### Compute spillover matrix

Based on the single-positive pixels, we use the `CATALYST::computeSpillmat()`
function to compute the spillover matrix and `CATALYST::plotSpillmat()` to
visualize it. The `plotSpillmat` function checks the spotted and acquired
metal isotopes against a pre-defined `CATALYST::isotope_list()`. 

```{r compute-spillover}

sce1_all <- computeSpillmat(sce1, interactions="all")
sce2_all <- computeSpillmat(sce2, interactions="all")

sce1 <- computeSpillmat(sce1, interactions="default")
sce2 <- computeSpillmat(sce2, interactions="default")

isotope_list <- CATALYST::isotope_list

```

The computeSpillMat excludes all interactions that are considered chemically and physically non-sensible, which is why we do not see the spillover from Pt196 to Lu175 anymore. Based on detailed examination, we now know that Pt196 spills into Lu175. We therefore decide to use the spillover matrix without binning and all interactions for the compensation. 

Here, we will save the spillover matrix as .csv.

### Save spillover matrix as csv

Next, we will write out the spillover matrix as .csv file for possible later use.

```{r write_sm_csv}
path <- "/mnt/Multimodal_Imaging_Daria/20220518_DL_SpilloverCorrection/Extra/spillover_matrix"
sm1 <- metadata(sce1)$spillover_matrix
sm2 <- metadata(sce2)$spillover_matrix
sm1_all <- metadata(sce1_all)$spillover_matrix
sm2_all <- metadata(sce2_all)$spillover_matrix
write.csv(sm1, file.path(path,"spillover_matrix_no-binning_default-interactions.csv"))
write.csv(sm2, file.path(path,"spillover_matrix_binning_default-interactions.csv"))
write.csv(sm1_all, file.path(path,"spillover_matrix_no-binning_all-interactions.csv"))
write.csv(sm2_all, file.path(path,"spillover_matrix_binning_all-interactions.csv"))

panel <- read.csv(file.path(path,"panel_sm.csv"))
adapted_sm <- adaptSpillmat(sm1_all, paste0(panel$channel[panel$keep == 1], "Di"), 
                            isotope_list = isotope_list)

write.csv(adapted_sm, file.path(path,"spillover_matrix_no-binning_all-interactions_final.csv"))
```


