---
title: "Differential abundance testing with limma-voom"
output: html_document
date: '2023-08-19'
---

```{r setup, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(echo = TRUE, out.width = "120%")
```

## Load data

First, we will find paired samples. 

```{r selection-paired-sample, message=FALSE}
library(SpatialExperiment)
library(CATALYST)
library(cowplot)
library(ggplot2)
library(diffcyt)
library(flowCore)
library(readxl)
library(scater)
library(dplyr)
library(tidyr)

path <- "/mnt/Multimodal_Imaging_Daria/PIPELINE/NEWEST_RESULTS"
spe <- readRDS(file.path(path,"results/Manuscript/spe_clustered.rds"))

clusternames_ordered <- read.csv(file.path(path, "results/Manuscript/celltypes_ordered_2.csv"), sep=";", colClasses = c("order" = "character"))
codes <- setNames(paste0(clusternames_ordered$order, "_", clusternames_ordered$celltype), clusternames_ordered$celltype)
spe$celltype_ordered <- recode(spe$celltype, !!!codes)

metadata(spe)$color_vectors$col_celltype_ordered <- metadata(spe)$color_vectors$col_celltype
names(metadata(spe)$color_vectors$col_celltype_ordered) <- recode(names(metadata(spe)$color_vectors$col_celltype), !!!codes)
```

## Subset dataset

We will first read in the previously generated `SpatialExperiment` object and transform it into a catalyst compatible object.

```{r read-data-pheno, message=FALSE}

keep_rows <- rowData(spe)$use_channel 
keep_cols <- spe$celltype != "other" & spe$tissue=="BM" & spe$metacluster!="tumor" & spe$metacluster!="mesenchymal" & spe$control=="no"

sce <- SingleCellExperiment(assays=list(counts=assay(spe, "counts")[keep_rows, keep_cols], exprs=assay(spe, "counts")[keep_rows, keep_cols]))

#for combining RE1 and RE2
sce$timepoint <- factor(spe[keep_rows, keep_cols]$timepoint)
sce[,sce$timepoint=="RE1"]$timepoint <- "RE2"
sce$condition <- factor(sce$timepoint)

sce$control <- factor(spe[keep_rows, keep_cols]$control)
sce$cluster_id <- factor(spe[keep_rows, keep_cols]$celltype_ordered)
sce$patient <- factor(spe[keep_rows, keep_cols]$study_id)
sce$tissue <- factor(spe[keep_rows, keep_cols]$tissue)
sce$metacluster <- factor(spe[keep_rows, keep_cols]$metacluster)
sce$tmp <- factor(spe[keep_rows, keep_cols]$tmp)

sce$mna <- factor(spe[keep_rows, keep_cols]$MYCN_amp)
mna <- as.vector(sce$mna)
mna <- replace(x = mna, list =  mna %in% c(1), values =  'MNA')
mna <- replace(x = mna, list =  mna %in% c(3), values =  'het')
mna <- replace(x = mna, list =  mna %in% c(0), values =  'nMNA')
mna <- replace(x = mna, list =  mna %in% c(9), values =  'unknown')
sce$mna <- as.factor(mna)

sce$progress <- factor(spe[keep_rows, keep_cols]$progression)
progress <- as.vector(sce$progress)
progress <- replace(x = progress, list =  progress==0, values =  'NoProg')
progress <- replace(x = progress, list =  progress==1, values =  'Prog')
sce$progress <- as.factor(progress)

sce$sample_id <- factor(paste0(sce$progress, "_", sce$mna, "_", spe[keep_rows, keep_cols]$study_id, "_", spe[keep_rows, keep_cols]$fm_id, "_", sce$tissue))

# Add celltype information to metadata
metadata(sce)$cluster_codes <- data.frame(celltype = factor(spe[keep_rows, keep_cols]$celltype_ordered))

# Add metadata
metadata(sce)$experiment_info <- as.data.frame(colData(sce)) %>% group_by(sample_id,patient, mna, condition, progress) %>% 
    summarise(n_cells=n(),.groups = 'drop') %>%
    as.data.frame()

# Define cell_type_markers 
type_markers <- c("MPO_Y89_mean", "CD44_In115_mean", "CD11b_Nd142_mean", "HLA-DR_Nd143_mean", "PRPH_Nd144_mean", "HLA-ABC_Sm147_mean", "CD20_Nd148_mean", "LUM_Sm149_mean", "CD11c_Nd150_mean",
                  "CD24_Eu151_mean", "CD3_Sm152_mean", "CD45_Eu153_mean", "CD8a_Sm154_mean", "GD2_Gd155_mean", "CD34_Gd156_mean", "CD10_Gd158_mean", "CXCR4_Tb159_mean","SOX10_Dy162_mean",
                  "FOXP3_Dy163_mean", "CHGA_Dy164_mean", "GATA3_Er168_mean", "CD56_Er170_mean", "CD4_Yb171_mean", "ELAVL4_Yb174_mean", "CD14_Lu175_mean", "Vimentin_Pt196_mean", "CD15_Bi209_mean")

# Define cell_state_markers 
state_markers <- c("CD274_Gd160_mean", "S100B_Dy161_mean", "CD279_Ho165_mean", "Ki-67_Tm169_mean", "GZMB_Yb173_mean")

# Add to spe
rowData(sce)$marker_class <- ifelse(rownames(sce) %in% type_markers, "type",
                                    ifelse(rownames(sce) %in% state_markers, "state", 
                                    "other"))

#Add dim reduction

reducedDim(sce, "UMAP") <- reducedDim(spe, "UMAP")[keep_cols,]
```

## Plot overview of cohort

Differential analysis of cell population abundance compares the proportions of cell types across experimental conditions and aims to highlight populations that are present at different ratios. First, we calculate two tables: one that contains cell counts for each sample and population and one with the corresponding  proportions of cell types by sample.  The proportions are used only for plotting, since the statistical modeling takes the cell counts by cluster and sample as input.

For each sample, we plot its PBMC cell type composition represented with colored bars, where the size of a given stripe reflects the proportion of the corresponding cell type in a given sample (Figure \@ref(fig:diff-freqs-plot-props-barplot)).

```{r diff-freqs-plot-props-barplot, fig.cap = "Relative abundance of the 8 PBMC populations in each sample (x-axis), in the PBMC dataset, represented with a barplot. The 8 cell populations are a result of manual merging of the 20 FlowSOM metaclusters", fig.width=20, fig.height=10}

set.seed(20231103)
p <- plotAbundances(sce, k = "celltype", by = "sample", col_clust=T, k_pal=metadata(spe)$color_vectors$col_celltype_ordered, 
                    linkage="ward.D", distance="manhattan") 
#Best linkage: ward.D
#Best distance: manhattan
```

## Differential cell population abundance
Since the diffcyt wrapper does not work, we will test the test-functions from the diffcyt package. 

```{r test-diffcyt, message=FALSE}

#Create list of samples as input to diffcyt methods
df_list <- lapply(unique(sce$sample_id), function(x){as.data.frame(t(counts(sce[,sce$sample_id==x])))})
names(df_list) <- unique(sce$sample_id)

#Create experiment info
experiment_info <- metadata(sce)$experiment_info
experiment_info <- experiment_info[match(names(df_list),experiment_info$sample_id),]

#Create marker info
# Add rowData
channel_name <- sapply(strsplit(rownames(rowData(spe)[keep_rows,]), "_"), "[[", 2)
marker_name <- sapply(strsplit(rownames(rowData(spe)[keep_rows,]), "_"), "[[", 1)
marker_class <-ifelse(rownames(spe)[keep_rows] %in% type_markers, "type",
                                    ifelse(rownames(spe)[keep_rows] %in% state_markers, "state", 
                                    "other"))
row_data <- data.frame(cbind(channel_name, marker_name), row.names=marker_name)
row_data$marker_class <- as.factor(marker_class)

marker_info <-  row_data

#Create summarized experiment - not working properly for some reason - cannot be fed to function diffcyt
d_se <- prepareData(df_list, experiment_info, marker_info)
rowData(d_se)$cluster_id <- sce[,rownames(d_se)]$cluster_id 
metadata(d_se)$cluster_codes <- metadata(sce)$cluster_codes

# Transform data
d_se <- transformData(d_se)

# Calculate counts
d_counts <- calcCounts(d_se)
```

```{r test-diffcyt-forloop, message=FALSE}
library(edgeR)
#Create design matrix
design <- model.matrix(~0+condition, experiment_info) 

colnames(design) <- c("DE", "RE2", "REL")
contrasts <- makeContrasts(DEvsRE2 = DE-RE2,
                           DEvsREL = DE-REL,
                           RE2vsREL = RE2-REL,
                           levels = design)

res <- list()

set.seed(230722)
for (i in 1:ncol(contrasts)){

    res_DA <- testDA_edgeR(d_counts,design, contrasts[,i], 
                       min_cells=1, 
                       min_samples=1
                       )
    
    res <- append(res, list(data.frame(rowData(res_DA))))
}


```

Next, we will plot these significance values onto a boxplot according to [this](https://www.datanovia.com/en/blog/ggpubr-how-to-add-p-values-generated-elsewhere-to-a-ggplot/).
```{r, fig.width=22, fig.height=18}
library(tidyr)
library(ggpubr)
library(rstatix)
library(gridExtra)
library(gtools)

p <- plotAbundances(sce, k = "celltype", by = "cluster_id", shape_by = "sample_id")

comparisons <-colnames(contrasts)

plot_list_values <- list()
plot_list_stars <- list()
plot_list_fc <- list()
for(i in 1:length(unique(p$data$cluster_id))){
  cluster_name <- unique(p$data$cluster_id)[i]
  proportions <- data.frame(p$data[p$data$cluster_id == cluster_name, c("sample_id", "condition", "Freq")])
  
  p_adj <- c()
  group1 <- c()
  group2 <- c()
  logFC <- c()
  
  for(j in 1:ncol(contrasts)){
  stats <- res[[j]]
  p_adj <- c(p_adj, round(stats[rownames(stats)==droplevels(cluster_name), "p_adj"],3))
  logFC <- c(logFC, -1*round(stats[rownames(stats)==droplevels(cluster_name), "logFC"],2))
  #logFC <- c(logFC, round(stats[rownames(stats)==droplevels(cluster_name), "logFC"],2))
  group1 <- c(group1, unlist(strsplit(comparisons[j], "vs"))[1])
  group2 <- c(group2, unlist(strsplit(comparisons[j], "vs"))[2])
  }

  stats_df <- data.frame(p_adj=p_adj, group1=group1, group2=group2, logFC=logFC)
  stats_df$p_adj_signif <- stars.pval(stats_df$p_adj)
  stats_df$p_adj_signif[stats_df$p_adj_signif==" "] <- "ns"
  stats_df$p_fc <- paste0(stats_df$logFC, " (", stats_df$p_adj, ")", " (", stats_df$p_adj_signif, ")")

  color <- unname(metadata(spe)$color_vectors$col_celltype_ordered[as.character(cluster_name)]) 
  
  plot_list_fc[[i]] <- ggboxplot(proportions, x = "condition", y = "Freq", order=c("DE", "RE2", "REL"),
              color = "condition", palette =c(color, color,color), 
              #lwd=1.6, 
              add.params = list(size = 3),
              add = "jitter", 
              title=cluster_name, show.legend = F) +
      stat_pvalue_manual(
        stats_df,
        y.position = max(proportions$Freq),
        step.increase = 0.1,
        label = "p_fc"
        )  +
      theme(aspect.ratio = 1.75, legend.position="none") +
      scale_y_continuous(expand = expansion(mult = c(0.05, 0.15)))
}

setEPS()
postscript(file.path(path, "results/Manuscript/5_Figure/R/BM_timepoint_ME_boxplot_edgeR.eps"), height=26, width=20)
grid.arrange(grobs = plot_list_fc, ncol = 6)
   dev.off()
```

Here we plot the abundances of celltypes however, we divide by the number of samples per group so that the result is not driven by outliers. 

```{r}
p <- plotAbundances(sce, k = "celltype", by = "cluster_id", shape_by = "sample_id")

df_normalized <- p$data %>%
  group_by(cluster_id, condition) %>%
  summarise(normalized_count = mean(Freq, na.rm = TRUE)) %>%
  ungroup()


df_normalized$cluster_id <- factor(df_normalized$cluster_id)

setEPS()
postscript(file.path(path, "results/Manuscript/5_Figure/R/BM_timepoint_tumor_dodge.eps"), height=4, width=8)


ggplot(df_normalized, aes(fill=cluster_id, y=normalized_count, x=condition)) + 
  geom_bar(
    stat="identity",
    position="dodge",
    show.legend = T) + 
  scale_fill_manual(values = metadata(spe)$color_vectors$col_celltype_ordered) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

dev.off()

#Piechart
setEPS()
postscript(file.path(path, "results/Manuscript/5_Figure/R/BM_timepoint_tumor_piechart.eps"), height=2, width=6.5)

ggplot(df_normalized, aes(x="", y=normalized_count, fill=cluster_id, group=cluster_id)) +
  geom_bar(stat="identity", width=1, color="white") +
  scale_fill_manual(values = metadata(spe)$color_vectors$col_celltype_ordered) +
  coord_polar("y", start=0)  +  
  facet_grid(.~ condition) +
  theme_void()
dev.off()
```


