---
title: "Differential abundance testing with limma-voom"
output: html_document
date: '2023-08-19'
---

```{r setup, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(echo = TRUE, out.width = "120%")
```

## Load data

First, we will find paired samples. 

```{r selection-paired-sample, message=FALSE}
library(SpatialExperiment)
library(CATALYST)
library(cowplot)
library(ggplot2)
library(diffcyt)
library(flowCore)
library(readxl)
library(scater)
library(dplyr)
library(tidyr)

path <- "/mnt/Multimodal_Imaging_Daria/PIPELINE/NEWEST_RESULTS"
spe <- readRDS(file.path(path,"results/Manuscript/spe_clustered.rds"))

clusternames_ordered <- read.csv(file.path(path, "results/Manuscript/celltypes_ordered_2.csv"), sep=";", colClasses = c("order" = "character"))
codes <- setNames(paste0(clusternames_ordered$order, "_", clusternames_ordered$celltype), clusternames_ordered$celltype)
spe$celltype_ordered <- recode(spe$celltype, !!!codes)

metadata(spe)$color_vectors$col_celltype_ordered <- metadata(spe)$color_vectors$col_celltype
names(metadata(spe)$color_vectors$col_celltype_ordered) <- recode(names(metadata(spe)$color_vectors$col_celltype), !!!codes)
```

## Subset dataset

We will first read in the previously generated `SpatialExperiment` object and transform it into a catalyst compatible object.

```{r read-data-pheno, message=FALSE}

samples_with_tumor_cells <- unique(spe[,spe$tissue=="BM"& spe$metacluster=="tumor"]$sample_id) #be careful here- use sample_id when looking for samples with and sample when looking for samples without tumor cells

keep_rows <- rowData(spe)$use_channel 
keep_cols <- spe$celltype != "other" & spe$tissue=="BM" & spe$sample_id%in%samples_with_tumor_cells & spe$control=="no"

sce <- SingleCellExperiment(assays=list(counts=assay(spe, "counts")[keep_rows, keep_cols], exprs=assay(spe, "counts")[keep_rows, keep_cols]))

#for combining RE1 and RE2
sce$timepoint <- factor(spe[keep_rows, keep_cols]$timepoint)
sce[,sce$timepoint=="RE1"]$timepoint <- "RE2"
sce$condition <- factor(sce$timepoint)

sce$control <- factor(spe[keep_rows, keep_cols]$control)
sce$cluster_id <- factor(spe[keep_rows, keep_cols]$celltype_ordered)
sce$patient <- factor(spe[keep_rows, keep_cols]$study_id)
sce$tissue <- factor(spe[keep_rows, keep_cols]$tissue)
sce$metacluster <- factor(spe[keep_rows, keep_cols]$metacluster)
sce$tmp <- factor(spe[keep_rows, keep_cols]$tmp)

sce$mna <- factor(spe[keep_rows, keep_cols]$MYCN_amp)
mna <- as.vector(sce$mna)
mna <- replace(x = mna, list =  mna %in% c(1), values =  'MNA')
mna <- replace(x = mna, list =  mna %in% c(3), values =  'het')
mna <- replace(x = mna, list =  mna %in% c(0), values =  'nMNA')
mna <- replace(x = mna, list =  mna %in% c(9), values =  'unknown')
sce$mna <- as.factor(mna)

sce$progress <- factor(spe[keep_rows, keep_cols]$progression)
progress <- as.vector(sce$progress)
progress <- replace(x = progress, list =  progress==0, values =  'NoProg')
progress <- replace(x = progress, list =  progress==1, values =  'Prog')
sce$progress <- as.factor(progress)

sce$sample_id <- factor(paste0(sce$progress, "_", sce$mna, "_", spe[keep_rows, keep_cols]$study_id, "_", spe[keep_rows, keep_cols]$fm_id, "_", sce$tissue))

# Add celltype information to metadata
metadata(sce)$cluster_codes <- data.frame(celltype = factor(spe[keep_rows, keep_cols]$celltype_ordered))

# Add metadata
metadata(sce)$experiment_info <- as.data.frame(colData(sce)) %>% group_by(sample_id,patient, mna, condition, progress) %>% 
    summarise(n_cells=n(),.groups = 'drop') %>%
    as.data.frame()

# Define cell_type_markers 
type_markers <- c("MPO_Y89_mean", "CD44_In115_mean", "CD11b_Nd142_mean", "HLA-DR_Nd143_mean", "PRPH_Nd144_mean", "HLA-ABC_Sm147_mean", "CD20_Nd148_mean", "LUM_Sm149_mean", "CD11c_Nd150_mean",
                  "CD24_Eu151_mean", "CD3_Sm152_mean", "CD45_Eu153_mean", "CD8a_Sm154_mean", "GD2_Gd155_mean", "CD34_Gd156_mean", "CD10_Gd158_mean", "CXCR4_Tb159_mean","SOX10_Dy162_mean",
                  "FOXP3_Dy163_mean", "CHGA_Dy164_mean", "GATA3_Er168_mean", "CD56_Er170_mean", "CD4_Yb171_mean", "ELAVL4_Yb174_mean", "CD14_Lu175_mean", "Vimentin_Pt196_mean", "CD15_Bi209_mean")

# Define cell_state_markers 
state_markers <- c("CD274_Gd160_mean", "S100B_Dy161_mean", "CD279_Ho165_mean", "Ki-67_Tm169_mean", "GZMB_Yb173_mean")

# Add to spe
rowData(sce)$marker_class <- ifelse(rownames(sce) %in% type_markers, "type",
                                    ifelse(rownames(sce) %in% state_markers, "state", 
                                    "other"))

#Add dim reduction

reducedDim(sce, "UMAP") <- reducedDim(spe, "UMAP")[keep_cols,]
```

## Plot overview of cohort

Differential analysis of cell population abundance compares the proportions of cell types across experimental conditions and aims to highlight populations that are present at different ratios. First, we calculate two tables: one that contains cell counts for each sample and population and one with the corresponding  proportions of cell types by sample.  The proportions are used only for plotting, since the statistical modeling takes the cell counts by cluster and sample as input.

For each sample, we plot its PBMC cell type composition represented with colored bars, where the size of a given stripe reflects the proportion of the corresponding cell type in a given sample (Figure \@ref(fig:diff-freqs-plot-props-barplot)).

```{r diff-freqs-plot-props-barplot, fig.cap = "Relative abundance of the 8 PBMC populations in each sample (x-axis), in the PBMC dataset, represented with a barplot. The 8 cell populations are a result of manual merging of the 20 FlowSOM metaclusters", fig.width=20, fig.height=10}
sce_tumor <- sce[,sce$metacluster=="tumor"]

set.seed(20231103)
p <- plotAbundances(sce_tumor, k = "celltype", by = "sample", col_clust=T, k_pal=metadata(spe)$color_vectors$col_celltype_ordered, 
                    linkage="ward.D", distance="manhattan") 
#Best linkage: ward.D
#Best distance: manhattan

condition <- c("DE", "RE2", "REL")
clustered_samples <- list()
for (i in seq_along(condition)){
  clustered_samples <- c(clustered_samples, sapply(strsplit(levels(droplevels(p$data[p$data$condition==condition[i],]$sample_id)), "_"), function(x) paste(x[4], x[5], sep = "_")))
}
clustered_samples <- as.character(clustered_samples)

#clustered_samples <- levels(p$data$sample_id) #when subsetting the dataset

setEPS()
postscript(file.path(path, "results/Manuscript/5_Figure/R/barplot_BM_timepointamples.eps"), height=12, width=20)
p
dev.off()
```

We now want to plot the tumor cells of the clustered microenvironments. 

```{r plot-tumor-compositions, message=FALSE, fig.width=20, fig.height=10}

sce$sample_id <- factor(paste0(spe[keep_rows, keep_cols]$fm_id, "_", spe[keep_rows, keep_cols]$tissue), levels=clustered_samples)
sce_ME <- sce[,sce$metacluster!="tumor" & sce$metacluster!="mesenchymal"]
sce_tumor <- sce[,sce$metacluster=="tumor"]

p_me <- plotAbundances(sce_ME, k = "celltype", by = "sample", col_clust=F, k_pal=metadata(spe)$color_vectors$col_celltype_ordered)

setEPS()
postscript(file.path(path, "results/Manuscript/5_Figure/R/barplot_BM_timepointamples_ME.eps"), height=12, width=20)
p_me
dev.off()
```

As a next step, we will compute the heterogeneity of samples based on the ME and tumor cell composition.
```{r heterogeneity, message=FALSE, fig.width=7, fig.height=7}
library(BioQC)
library(hrbrthemes)

proportions <- plotAbundances(sce, k = "celltype", by = "sample")
prop <- proportions$data %>%
  select(-condition) %>%
  pivot_wider(names_from = cluster_id, values_from = Freq)

prop <- data.frame(prop)
rownames(prop) <- prop$sample_id
prop <- prop %>% select(-sample_id)

prop <- t(prop)
rownames(prop) <- levels(sce$cluster_id)
me_clusters <- unique(sce[,sce$metacluster!="tumor" & sce$metacluster!="mesenchymal"]$cluster_id)
me_prop <- prop[rownames(prop)%in%me_clusters,]
me_het <- apply(me_prop,2, entropy)
  
tumor_clusters <- unique(sce[,sce$metacluster=="tumor"]$cluster_id)
tumor_prop <- prop[rownames(prop)%in%tumor_clusters,]
tumor_het <- apply(tumor_prop,2, entropy)
```


Next, we will plot metadata on top of the clustered barplot. First, we will define the color code for the metadata. Find all built-in R color names [here](https://www.farb-tabelle.de/de/farbtabelle.htm). The [paletteer package](https://r-charts.com/color-palettes/) offers a wide range of color palettes.  

```{r define-color-code, message=FALSE}
library(ComplexHeatmap)
library(RColorBrewer)
library(paletteer)
library(bruceR)
library(circlize)
library(viridis)

meta_path <- file.path(path, '20231128_metadata_complete.csv')

qual_col_pals = brewer.pal.info[brewer.pal.info$category == 'qual',]
col_vector_74 = unlist(mapply(brewer.pal, qual_col_pals$maxcolors, rownames(qual_col_pals)))

col_vector_433 = colors()[grep('gr(a|e)y', grDevices::colors(), invert = T)]

meta_all = read.csv(meta_path)
meta_all <- meta_all[order(meta_all$Study_ID),] #order by study-id

exclude <- replace(x = meta_all$exclude, list =  !meta_all$exclude %in% c('no', 'maybe'), values =  'partially')

#DE_year <- sub("-.*", "", meta_all$DE_date)

col_sample = setNames(sample(col_vector_433, length(unique(meta_all$Sample))), unique(meta_all$Sample))
col_puncture_side = setNames(brewer.pal(length(unique(meta_all$Puncture_side)), 'Set3'), unique(meta_all$Puncture_side))
col_staining_batch = setNames(viridis(length(unique(meta_all$Staining_date))), sort(unique(meta_all$Staining_date)))
col_stainer = setNames(c('#a6611a', '#dfc27d', '#80cdc1', '#018571'), unique(meta_all$Stainer))
col_imc_batch = setNames(viridis(length(unique(meta_all$IMC_date))), sort(unique(meta_all$IMC_date)))
col_imc_score = colorRamp2(c(min(meta_all$IMC_score, na.rm = TRUE), max(meta_all$IMC_score, na.rm = TRUE)/2, max(meta_all$IMC_score, na.rm = TRUE)), viridis(3)) 
col_study_id = setNames(sample(col_vector_433, length(unique(meta_all$Study_ID))), unique(meta_all$Study_ID))
col_DE_year = setNames(viridis(length(unique(meta_all$DE_date))), sort(unique(meta_all$DE_date)))
col_timepoint = setNames(c('#7fc97f', '#beaed4', '#386cb0', '#ffff99', '#386cb0'), unique(meta_all$Timepoint)) #'#fdc086'
col_aberrations = setNames(c('#a6611a','#018571','#80cdc1','#f5f5f5'), c(0,1,2,9))
col_stages = setNames(c('#a6cee3', '#1f78b4', '#b2df8a', '#33a02c', '#fb9a99', '#e31a1c'), unique(meta_all$stage4))
col_efs_dur = colorRamp2(c(min(meta_all$efs_dur, na.rm = TRUE), max(meta_all$efs_dur, na.rm = TRUE)/2, max(meta_all$efs_dur, na.rm = TRUE)), viridis(3))
col_tissue = setNames(c('aquamarine4','brown'), unique(meta_all$Tissue))
col_control = setNames(c('gray77', 'mediumseagreen'), unique(meta_all$control))
col_exclude = setNames(c('#018571','#a6611a','#80cdc1'), unique(exclude))
col_progress = setNames(c('gray77', 'palegreen3', 'firebrick1'), c('n.a.', 0, 1))
#col_age_diag = colorRamp2(c(min(meta_all$age_at_diag, na.rm = TRUE), max(meta_all$age_at_diag, na.rm = TRUE)/2, max(meta_all$age_at_diag, na.rm = TRUE)), viridis(3))
col_age_diag = colorRamp2(c(min(meta_all$age_at_diag, na.rm = TRUE), (quantile(meta_all$age_at_diag, 0.95, na.rm = TRUE)-min(meta_all$age_at_diag, na.rm = TRUE))/2, quantile(meta_all$age_at_diag,0.95, na.rm = TRUE)), viridis(3))
col_therapy = setNames(c('#a6611a', '#018571'), c('RAPID COJEC', 'MOD. N7'))

col_me_het = colorRamp2(c(min(me_het, na.rm = TRUE), (max(me_het, na.rm = TRUE)-min(me_het, na.rm = TRUE))/2, max(me_het, na.rm = TRUE)), viridis(3))
#col_tumor_het = colorRamp2(c(sort(unique(tumor_het))[2], (max(tumor_het, na.rm = TRUE)-sort(unique(tumor_het))[2])/2, max(tumor_het, na.rm = TRUE)), viridis(3))
col_tumor_het = colorRamp2(c(min(tumor_het, na.rm = TRUE), (max(tumor_het, na.rm = TRUE)-min(tumor_het, na.rm = TRUE))/2, max(tumor_het, na.rm = TRUE)), viridis(3))
```

Now, we will plot an overview of the cohort.

```{r overview-cohort, message=FALSE, fig.height=25}

rownames(meta_all) <- paste0(meta_all$Sample, "_", meta_all$Tissue)
meta_all <- meta_all[clustered_samples,]

setEPS()
postscript(file.path(path, "results/Manuscript/5_Figure/R/annoationbar_BM_timepoint.eps"), height=30, width=30)

columnAnnotation(control=meta_all$control, col=list(control=col_control)) %v%
columnAnnotation(sample=meta_all$Sample, col=list(sample=col_sample), annotation_legend_param = list(sample = list(grid_height=unit(0.05, "cm"), labels_gp = gpar(fontsize = 5)))) %v%
#columnAnnotation(exclusion=exclude, col=list(exclusion=col_exclude)) %v%
columnAnnotation(imc_score=meta_all$IMC_score, col=list(imc_score=col_imc_score))  %v%
columnAnnotation(staining_batch=meta_all$Staining_date, col=list(staining_batch=col_staining_batch)) %v%
columnAnnotation(stainer=meta_all$Stainer, col=list(stainer=col_stainer)) %v%
columnAnnotation(imc_batch=meta_all$IMC_date, col=list(imc_batch=col_imc_batch)) %v%
columnAnnotation(D_year=meta_all$DE_date, col=list(D_year=col_DE_year))  %v%
columnAnnotation(study_id=meta_all$Study_ID, col=list(study_id=col_study_id))  %v%
columnAnnotation(timepoint=meta_all$Timepoint, col=list(timepoint=col_timepoint))  %v%
columnAnnotation(tissue=meta_all$Tissue, col=list(tissue=col_tissue)) %v%
columnAnnotation(stage=meta_all$stage4, col=list(stage=col_stages)) %v%
columnAnnotation(progression=meta_all$efs, col=list(progression=col_progress))  %v%
columnAnnotation(efs_dur=meta_all$efs_dur, col=list(efs_dur=col_efs_dur)) %v%

columnAnnotation(X1p_loss=meta_all$X1p_loss, col=list(X1p_loss=col_aberrations), show_legend=FALSE) %v%
columnAnnotation(X1q_gain=meta_all$X1q_gain, col=list(X1q_gain=col_aberrations), show_legend=FALSE) %v%
columnAnnotation(X1q_loss=meta_all$X1q_loss, col=list(X1q_loss=col_aberrations), show_legend=FALSE) %v%
columnAnnotation(X2p_gain=meta_all$X2p_gain, col=list(X2p_gain=col_aberrations), show_legend=FALSE) %v%
columnAnnotation(X3p_loss=meta_all$X3p_loss, col=list(X3p_loss=col_aberrations), show_legend=FALSE) %v%
columnAnnotation(X3q_loss=meta_all$X3q_loss, col=list(X3q_loss=col_aberrations), show_legend=FALSE) %v%
columnAnnotation(X4p_loss=meta_all$X4p_loss, col=list(X4p_loss=col_aberrations), show_legend=FALSE) %v%
columnAnnotation(X6q_loss=meta_all$X6q_loss, col=list(X6q_loss=col_aberrations), show_legend=FALSE) %v%
columnAnnotation(X7q_gain=meta_all$X7q_gain, col=list(X7q_gain=col_aberrations), show_legend=FALSE) %v%
columnAnnotation(PTPRD_del=meta_all$PTPRD_del, col=list(PTPRD_del=col_aberrations), show_legend=FALSE) %v%
columnAnnotation(CDKN2AB_del=meta_all$CDKN2AB_del, col=list(CDKN2AB_del=col_aberrations), show_legend=FALSE) %v%
columnAnnotation(X11q_loss=meta_all$X11q_loss, col=list(X11q_loss=col_aberrations), show_legend=FALSE) %v%
columnAnnotation(X12q_gain=meta_all$X12q_gain, col=list(X12q_gain=col_aberrations), show_legend=FALSE) %v%
columnAnnotation(X14q_loss=meta_all$X14q_loss, col=list(X14q_loss=col_aberrations), show_legend=FALSE) %v%
columnAnnotation(X17p_loss=meta_all$X17p_loss, col=list(X17p_loss=col_aberrations), show_legend=FALSE) %v%
columnAnnotation(X17q_gain=meta_all$X17q_gain, col=list(X17q_gain=col_aberrations), show_legend=FALSE) %v%
columnAnnotation(X19p_loss=meta_all$X19p_loss, col=list(X19p_loss=col_aberrations), show_legend=FALSE) %v%
columnAnnotation(X19q_loss=meta_all$X19q_loss, col=list(X19q_loss=col_aberrations), show_legend=FALSE) %v%
columnAnnotation(X22q_loss=meta_all$X22q_loss, col=list(X22q_loss=col_aberrations), show_legend=FALSE) %v%
columnAnnotation(CDKN2AB_del=meta_all$CDKN2AB_del, col=list(CDKN2AB_del=col_aberrations), show_legend=FALSE) %v%
columnAnnotation(X11q_loss=meta_all$X11q_loss, col=list(X11q_loss=col_aberrations), show_legend=FALSE) %v%
columnAnnotation(X12q_gain=meta_all$X12q_gain, col=list(X12q_gain=col_aberrations), show_legend=FALSE) %v%
columnAnnotation(X14q_loss=meta_all$X14q_loss, col=list(X14q_loss=col_aberrations), show_legend=FALSE) %v%
columnAnnotation(MYCN_amp=meta_all$MYCN_amp, col=list(MYCN_amp=col_aberrations)) %v% #annotation_legend_param = list(MYCN_amp = list(title='aberration', 
                                                                                                              #labels = c('no','yes','het'))))  %v%
                                                                                                              #labels = c('no', 'yes', 'het', 'NA'))))  %v%
columnAnnotation(TERT_rear=meta_all$TERT_rear, col=list(TERT_rear=col_aberrations), show_legend=FALSE) %v%
columnAnnotation(ATRX_del=meta_all$ATRX_del, col=list(ATRX_del=col_aberrations), show_legend=FALSE) %v% 
columnAnnotation(ALK_amp=meta_all$ALK_amp, col=list(ALK_amp=col_aberrations), show_legend=FALSE) %v%
columnAnnotation(ALK_mut=meta_all$ALK_mut, col=list(ALK_mut=col_aberrations), show_legend=FALSE) %v%

columnAnnotation(age=meta_all$age_at_diag, col=list(age=col_age_diag)) %v%
columnAnnotation(therapy=meta_all$therapy, col=list(therapy=col_therapy)) %v%
columnAnnotation(ME_het=me_het, col=list(ME_het=col_me_het)) %v%
columnAnnotation(TUM_het=tumor_het, col=list(TUM_het=col_tumor_het))


dev.off()

#absolute cellcount
setEPS()
postscript(file.path(path, paste0("results/Manuscript/5_Figure/R/BM_timepoint_annotationbar_ncells.eps")), height=30, width=30)
draw(anno_barplot(as.vector(unname(table(sce$sample_id))), add_numbers=T, height = unit(1, "cm")))
dev.off() 

setEPS()
postscript(file.path(path, paste0("results/Manuscript/5_Figure/R/BM_timepoint_annotationbar_ntumorcells.eps")), height=30, width=30)
draw(anno_barplot(as.vector(unname(table(sce_tumor$sample_id))), add_numbers=T, height = unit(1, "cm")))
dev.off() 

#tumor cell proportion
norm_minmax <- function(x){(x- min(x)) /(max(x)-min(x))}

robust_scalar<- function(x){(x- median(x)) /(quantile(x,probs = .75)-quantile(x,probs = .25))}

clipper <- function(data, lower, upper) {
  data[data > quantile(data, probs = c(upper))] <- quantile(data, probs = c(upper))
  data[data < quantile(data, probs = c(lower))] <- quantile(data, probs = c(lower))
  clipped_data <- norm_minmax(data)
  return(clipped_data)}

tumor_infiltration <- round((unname(table(sce_tumor$sample_id))/unname(table(sce$sample_id)))*100,3)
tumor_inf_norm <- round(clipper(tumor_infiltration, 0, 0.65),3)
setEPS()
postscript(file.path(path, paste0("results/Manuscript/5_Figure/R/BM_timepoint_annotationbar_tumor_inf.eps")), height=30, width=30)
draw(anno_barplot(as.vector(tumor_infiltration), add_numbers=T, height = unit(1, "cm")))
dev.off() 

setEPS()
postscript(file.path(path, paste0("results/Manuscript/5_Figure/R/BM_timepoint_annotationbar_tumor_inf_norm.eps")), height=30, width=30)
draw(anno_barplot(as.vector(tumor_inf_norm), add_numbers=T, height = unit(1, "cm")))
dev.off() 
```

As a last step, we will plot summary plots (boxplots, piecharts) of continuous and discrete metavariables per derived community cluster.
```{r summary_plots-dis, message=FALSE, fig.height=40}
library(gridExtra)

#when testing between other groups than cellular communities
tmp <- meta_all$Timepoint
tmp[tmp=="RE1"] <- "RE2"
condition <- tmp

meta_all$condition <- condition

discrete_metadata <- c("control", "Tissue", "Timepoint", "X1p_loss", "X1q_gain", "X1q_loss", "X2p_gain", "X3p_loss", "X3q_loss", "X4p_loss",
                       "X6q_loss", "X7q_gain", "PTPRD_del", "CDKN2AB_del", "X11q_loss", "X12q_gain", "X14q_loss", "X17p_loss", "X17q_gain",
                       "X19p_loss", "X19q_loss", "X22q_loss", "MYCN_amp", "TERT_rear", "ATRX_del", "ALK_amp", "ALK_mut", "efs", "therapy")

col_timepoint = setNames(c('#beaed4', '#386cb0', '#386cb0', '#ffff99'), c("DE", "RE1", "RE2", "REL")) 
#col_aberrations = setNames(c('#a6611a','#018571','#80cdc1','#f5f5f5'), c(0,1,3,9))

discrete_metadata_colors <- list(col_control, col_tissue, col_timepoint, col_aberrations, col_aberrations, col_aberrations, col_aberrations, col_aberrations, col_aberrations, col_aberrations, col_aberrations, col_aberrations, col_aberrations, col_aberrations, col_aberrations, col_aberrations, col_aberrations, col_aberrations, col_aberrations, col_aberrations, col_aberrations, col_aberrations, col_aberrations, col_aberrations, col_aberrations, col_aberrations, col_aberrations, col_progress, col_therapy)

plot_list <- list()
for (d in seq_along(discrete_metadata)) {
  meta <- na.omit(meta_all[,c(discrete_metadata[d], "condition")]) #filter out rows with NA values
  meta <- meta[rowSums(meta == "") == 0, ] # filter out rows with "", e.g. for timepoint in controls
  meta <- meta[rowSums(meta == 9) == 0, ] # filter out rows with "", e.g. for timepoint in controls
  colnames(meta) <- c("dis_var", "condition")
  
  result <- meta %>%
    group_by(condition, dis_var) %>%
    summarise(Freq = n(), .groups = 'drop')
  
  result[,"dis_var"]<- as.factor(unlist(result[,"dis_var"]))
  
plot_list[[d]] <- ggplot(result, aes(x="", y=Freq, fill=dis_var, group=dis_var)) +
    geom_col(position = "fill") +
    scale_fill_manual(values = discrete_metadata_colors[[d]], name=discrete_metadata[d]) +
    coord_polar("y", start=0)  +  
    facet_grid(.~ condition) +
    theme_void()
}

setEPS()
postscript(file.path(path, paste0("results/Manuscript/5_Figure/R/BM_timepoint_discrete.eps")), height=30, width=5)
do.call("grid.arrange", c(plot_list, ncol=1))
dev.off() 

```
We will then plot continuous variables in boxplots. 

```{r summary_plots-con, message=FALSE, fig.height=25, fig.width=5}
library("ggpubr")
library(gtools)

condition <- as.factor(condition)
n_overall <- as.vector(unname(table(sce$sample_id)[clustered_samples]))
n_tumor <- as.vector(unname(table(sce_tumor$sample_id)[clustered_samples]))
perc_tumor <- round((as.vector(unname(table(sce_tumor$sample_id)[clustered_samples]))/as.vector(unname(table(sce$sample_id)[clustered_samples])))*100,3)
perc_tumor_norm <- round(clipper(perc_tumor, 0, 0.65),3)
het_m <- as.vector(unname(me_het[clustered_samples]))
het_t <- as.vector(unname(tumor_het[clustered_samples]))

age <- meta_all$age_at_diag
age_clipped <- meta_all$age_at_diag
age_clipped[age_clipped > quantile(age_clipped, 0.95, na.rm=T) & !is.na(age_clipped)] <- quantile(age_clipped, 0.95, na.rm=T)

cont_df <- data.frame(cbind(n_overall, n_tumor, perc_tumor, perc_tumor_norm, het_m, het_t, age, age_clipped, condition),row.names=clustered_samples)

plot_list <- list()

for (d in colnames(cont_df)[1:ncol(cont_df)-1]) {

  df <- cont_df[,c(d, "condition")] 
  
  colnames(df) <- c("y", "condition")
  stats_df <- compare_means(y ~ condition,  
                        p.adjust.method = "BH", 
                        method='wilcox.test', 
                        data = df)
  
  stats_df$p_adj_signif <- stars.pval(stats_df$p.adj)
  stats_df$p_adj_signif[stats_df$p_adj_signif==" "] <- "ns"
  stats_df$p_long <- paste0(stats_df$p.adj, " (", stats_df$p_adj_signif, ")")

#logarithmic scale for perc_tumor
  if (d == "perc_tumor"){
    plot_list[[d]] <- ggboxplot(df, x = "condition", y = d,
            color = "condition", palette =c("slategray1", "slategray", "slategray3"), 
            add.params = list(size = 3),
            add = "jitter")+ 
            stat_pvalue_manual(stats_df, label = "p_long", y.position = log10(max(df$y)), step.increase = 0.2,) +
            theme(aspect.ratio = 1.5, legend.position="none") +
            scale_y_log10(
                breaks = scales::trans_breaks("log10", function(x) 10^x),
                labels = scales::trans_format("log10", scales::math_format(10^.x))
            ) #+
            #theme_bw() +
            #annotation_logticks(sides = 'lr') +
            #theme(panel.grid.minor = element_blank())
  }else{
    plot_list[[d]] <- ggboxplot(df, x = "condition", y = d,
                color = "condition", palette =c("slategray1", "slategray", "slategray3"), 
                add.params = list(size = 3),
                add = "jitter")+ 
                stat_pvalue_manual(stats_df, label = "p_long", y.position = max(df$y), step.increase = 0.2,) +
                theme(aspect.ratio = 1.5, legend.position="none") 
  }
}

setEPS()
postscript(file.path(path, paste0("results/Manuscript/5_Figure/R/BM_timepoint_continuous.eps")), height=30, width=5)
do.call("grid.arrange", c(plot_list, ncol=1))
dev.off()


```

Finally we will plot the metaclusters. 

```{r plot-metaclusters, message=FALSE, fig.width=20, fig.height=10}
sce$cluster_id <- sce$metacluster
metadata(sce)$cluster_codes <- data.frame(celltype = factor(sce$metacluster))
p_all <- plotAbundances(sce, k = "celltype", by = "sample", col_clust=F, k_pal=metadata(spe)$color_vectors$col_metacluster)

setEPS()
postscript(file.path(path, "results/Manuscript/5_Figure/R/barplot_BM_timepoint_metaclusters_samples.eps"), height=12, width=20)
p_all
dev.off()
```
```