---
title: "Differential abundance testing with limma-voom"
output: html_document
date: '2023-08-19'
---

```{r setup, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(echo = TRUE, out.width = "120%")
```

## Load data

First, we will find paired samples. 

```{r selection-paired-sample, message=FALSE}
library(SpatialExperiment)
library(CATALYST)
library(cowplot)
library(ggplot2)
library(diffcyt)
library(flowCore)
library(readxl)
library(scater)
library(dplyr)
library(tidyr)

path <- "/mnt/Multimodal_Imaging_Daria/PIPELINE/NEWEST_RESULTS"
spe <- readRDS(file.path(path,"results/in_house/spe_clustered_5.rds"))
```

We will first read in the previously generated `SpatialExperiment` object and transform it into a catalyst compatible object.

```{r read-data-pheno, message=FALSE}
BM_patients_1 <- unique(spe[,spe$tissue=="BM" & spe$control=="no" & spe$timepoint == "DE"]$study_id) #spe$metacluster!="tumor")
BM_patients_2 <- unique(spe[,spe$tissue=="BM" & spe$control=="no" & spe$timepoint == "RE2"]$study_id) #spe$metacluster!="tumor")
#PT_patients <- unique(spe[,spe$tissue=="PT"& spe$metacluster=="tumor"]$study_id)
paired_patients <- intersect(BM_patients_1, BM_patients_2)

keep_rows <- rowData(spe)$use_channel
keep_cols <- spe$tissue=="BM" & spe$control=="no" & spe$metacluster=="tumor" & spe$celltype!="other" & spe$celltype!="ELAVL4+"  #& spe$study_id%in%paired_patients & (spe$timepoint=="DE" | spe$timepoint=="RE2") #& spe$study_id%in%paired_patients & (spe$MYCN_amp==1 | spe$MYCN_amp==3) 
#spe$metacluster!="LIN- prol." & spe$metacluster!="other" #spe$study_id%in%paired_patients & (spe$timepoint=="DE" | spe$timepoint=="RE1") 
# keep_cols <- spe$tissue=="BM" & spe$control=="no" & spe$celltype!="other" & spe$metacluster!="tumor" & spe$timepoint=="DE"

sce <- SingleCellExperiment(assays=list(counts=assay(spe, "counts")[keep_rows, keep_cols], exprs=assay(spe, "counts")[keep_rows, keep_cols]))

sce$condition <- factor(spe[keep_rows, keep_cols]$timepoint)
#sce$condition <- factor(spe[keep_rows, keep_cols]$control)
sce$cluster_id <- factor(spe[keep_rows, keep_cols]$celltype)
#sce$cluster_id <- factor(spe[keep_rows, keep_cols]$predicted_clusters)
sce$patient <- factor(spe[keep_rows, keep_cols]$study_id)
sce$tissue <- factor(spe[keep_rows, keep_cols]$tissue)

sce$mna <- factor(spe[keep_rows, keep_cols]$MYCN_amp)
mna <- as.vector(sce$mna)
mna <- replace(x = mna, list =  mna %in% c(1, 3), values =  'MNA')
mna <- replace(x = mna, list =  mna %in% c(0), values =  'nMNA')
mna <- replace(x = mna, list =  mna %in% c(9), values =  'unknown')
sce$mna <- as.factor(mna)

sce$progress <- factor(spe[keep_rows, keep_cols]$progression)
progress <- as.vector(sce$progress)
progress <- replace(x = progress, list =  progress==0, values =  'NoProg')
progress <- replace(x = progress, list =  progress==1, values =  'Prog')
sce$progress <- as.factor(progress)

#sce$condition <- sce$progress

sce$sample_id <- factor(paste0(sce$mna, "_", sce$progress, "_", spe[keep_rows, keep_cols]$study_id, "_", spe[keep_rows, keep_cols]$fm_id, "_", sce$tissue))
#sce$sample_id <- factor(sample)
#sce$sample_id <- factor(paste0(spe[keep_rows, keep_cols]$study_id, "_", spe[keep_rows, keep_cols]$fm_id))

# Add celltype information to metadata
metadata(sce)$cluster_codes <- data.frame(celltype = factor(spe[keep_rows, keep_cols]$celltype))
#metadata(sce)$cluster_codes <- data.frame(celltype = factor(spe[keep_rows, keep_cols]$predicted_clusters))

# Add metadata
metadata(sce)$experiment_info <- as.data.frame(colData(sce)) %>% group_by(sample_id,patient, mna, condition, progress) %>% 
    summarise(n_cells=n(),.groups = 'drop') %>%
    as.data.frame()

# Define cell_type_markers 
type_markers <- c("MPO_Y89_mean", "CD44_In115_mean", "CD11b_Nd142_mean", "HLA-DR_Nd143_mean", "PRPH_Nd144_mean", "HLA-ABC_Sm147_mean", "CD20_Nd148_mean", "LUM_Sm149_mean", "CD11c_Nd150_mean",
                  "CD24_Eu151_mean", "CD3_Sm152_mean", "CD45_Eu153_mean", "CD8a_Sm154_mean", "GD2_Gd155_mean", "CD34_Gd156_mean", "CD10_Gd158_mean", "CXCR4_Tb159_mean","SOX10_Dy162_mean",
                  "FOXP3_Dy163_mean", "CHGA_Dy164_mean", "GATA3_Er168_mean", "CD56_Er170_mean", "CD4_Yb171_mean", "ELAVL4_Yb174_mean", "CD14_Lu175_mean", "Vimentin_Pt196_mean", "CD15_Bi209_mean")

# Define cell_state_markers 
state_markers <- c("CD274_Gd160_mean", "S100B_Dy161_mean", "CD279_Ho165_mean", "Ki-67_Tm169_mean", "GZMB_Yb173_mean")

# Add to spe
rowData(sce)$marker_class <- ifelse(rownames(sce) %in% type_markers, "type",
                                    ifelse(rownames(sce) %in% state_markers, "state", 
                                    "other"))

#Add dim reduction

reducedDim(sce, "UMAP") <- reducedDim(spe, "UMAP")[keep_cols,]


# keep <- as.vector(spe_cat[,spe_cat$tissue=="BM" & spe_cat$control=="no" & spe_cat$metacluster=="tumor"]$sample_id)
# sce <- filterSCE(spe_cat, sample_id %in% keep) 
```

## Differential cell population abundance

Differential analysis of cell population abundance compares the proportions of cell types across experimental conditions and aims to highlight populations that are present at different ratios. First, we calculate two tables: one that contains cell counts for each sample and population and one with the corresponding  proportions of cell types by sample.  The proportions are used only for plotting, since the statistical modeling takes the cell counts by cluster and sample as input.

For each sample, we plot its PBMC cell type composition represented with colored bars, where the size of a given stripe reflects the proportion of the corresponding cell type in a given sample (Figure \@ref(fig:diff-freqs-plot-props-barplot)).

```{r diff-freqs-plot-props-barplot, fig.cap = "Relative abundance of the 8 PBMC populations in each sample (x-axis), in the PBMC dataset, represented with a barplot. The 8 cell populations are a result of manual merging of the 20 FlowSOM metaclusters", fig.height=9, fig.width=12}
#
tiff(filename = "/mnt/Multimodal_Imaging_Daria/PIPELINE/NEWEST_RESULTS/results/poster/barplot_tumor_anniversary.tiff",
     width = 219, height = 100, units = "mm",
     res=600)
plotAbundances(sce, k = "celltype", by = "sample", col_clust=F, k_pal=metadata(spe)$color_vectors$col_celltype)
dev.off()
```

#test-functions Diffcyt
Since the diffcyt wrapper does not work, we will test the test-functions from the diffcyt package. 

```{r test-diffcyt, message=FALSE}

#Create list of samples as input to diffcyt methods
df_list <- lapply(unique(sce$sample_id), function(x){as.data.frame(t(counts(sce[,sce$sample_id==x])))})
names(df_list) <- unique(sce$sample_id)

#Create experiment info
experiment_info <- metadata(sce)$experiment_info
experiment_info <- experiment_info[match(names(df_list),experiment_info$sample_id),]

#Create marker info
# Add rowData
channel_name <- sapply(strsplit(rownames(rowData(spe)[keep_rows,]), "_"), "[[", 2)
marker_name <- sapply(strsplit(rownames(rowData(spe)[keep_rows,]), "_"), "[[", 1)
marker_class <-ifelse(rownames(spe)[keep_rows] %in% type_markers, "type",
                                    ifelse(rownames(spe)[keep_rows] %in% state_markers, "state", 
                                    "other"))
row_data <- data.frame(cbind(channel_name, marker_name), row.names=marker_name)
row_data$marker_class <- as.factor(marker_class)

marker_info <-  row_data

#Create summarized experiment - not working properly for some reason - cannot be fed to function diffcyt
d_se <- prepareData(df_list, experiment_info, marker_info)
rowData(d_se)$cluster_id <- sce[,rownames(d_se)]$cluster_id 
metadata(d_se)$cluster_codes <- metadata(sce)$cluster_codes

# Transform data
d_se <- transformData(d_se)

# Calculate counts
d_counts <- calcCounts(d_se)
```

```{r test-diffcyt, message=FALSE}
# Create design matrix
#design <- createDesignMatrix(experiment_info, cols_design = c("condition", "patient")) #paired
design <- createDesignMatrix(experiment_info, cols_design = "condition")

# Create contrast matrix
contrast <- createContrast(c(0, 1)) # unpaired
#contrast <- createContrast(c(0, 1, rep(0, length(unique(experiment_info$patient))-1))) #paired
nrow(contrast) == ncol(design)

# Test for differential abundance (DA) of clusters
res_DA <- testDA_voom(d_counts,design, contrast, 
                       min_cells=1, 
                       min_samples=1
                       )


FDR_cutoff = 0.1
rowData(res_DA)[rowData(res_DA)$p_adj<FDR_cutoff,]
res<- data.frame(rowData(res_DA))
```


```{r test-diffcyt-forloop, message=FALSE}
library(edgeR)
#Create design matrix
design <- model.matrix(~0+condition, experiment_info) 

colnames(design) <- c("DE", "RE1", "RE2", "REL")
contrasts <- makeContrasts(DEvsRE1 = DE-RE1,
                           DEvsRE2 = DE-RE2,
                           DEvsREL = DE-REL,
                           RE1vsREL = RE1-REL,
                           RE2vsREL = RE2-REL,
                           levels = design)

res <- list()

set.seed(230722)
for (i in 1:ncol(contrasts)){

    res_DA <- testDA_edgeR(d_counts,design, contrasts[,i], 
                       min_cells=1, 
                       min_samples=1
                       )
    
    res <- append(res, list(data.frame(rowData(res_DA))))
}


```

Next, we will plot these significance values onto a boxplot according to [this](https://www.datanovia.com/en/blog/ggpubr-how-to-add-p-values-generated-elsewhere-to-a-ggplot/).
```{r, fig.width=14, fig.height=9}
library(tidyr)
library(ggpubr)
library(rstatix)
library(gridExtra)
library(gtools)

p <- plotAbundances(sce, k = "celltype", by = "cluster_id", shape_by = "sample_id")


comparisons <-colnames(contrasts)
#only for one comparison
#comparisons <- c("NoProgvsProg")

plot_list_values <- list()
plot_list_stars <- list()
plot_list_fc <- list()
for(i in 1:length(unique(p$data$cluster_id))){
  cluster_name <- unique(p$data$cluster_id)[i]
  proportions <- data.frame(p$data[p$data$cluster_id == cluster_name, c("sample_id", "condition", "Freq")])
  
  p_adj <- c()
  group1 <- c()
  group2 <- c()
  logFC <- c()
  
  for(j in 1:ncol(contrasts)){
  #for(j in 1:1){
    stats <- res[[j]]
    #stats <- res
    p_adj <- c(p_adj, round(stats[rownames(stats)==droplevels(cluster_name), "p_adj"],3))
    logFC <- c(logFC, -1*round(stats[rownames(stats)==droplevels(cluster_name), "logFC"],2))
    #logFC <- c(logFC, round(stats[rownames(stats)==droplevels(cluster_name), "logFC"],2))
    group1 <- c(group1, unlist(strsplit(comparisons[j], "vs"))[1])
    group2 <- c(group2, unlist(strsplit(comparisons[j], "vs"))[2])
  }
  
  stats_df <- data.frame(p_adj=p_adj, group1=group1, group2=group2, logFC=logFC)
  stats_df$p_adj_signif <- stars.pval(stats_df$p_adj)
  stats_df$p_adj_signif[stats_df$p_adj_signif==" "] <- "ns"
  stats_df$p_fc <- paste0(stats_df$logFC, " (", stats_df$p_adj, ")", " (", stats_df$p_adj_signif, ")")
  
  plot_list_fc[[i]] <- ggboxplot(proportions, x = "condition", y = "Freq",
              color = "condition", palette =c("#e41a1c", "#377eb8", "#4daf4a", "#984ea3"),
              add = "jitter", title=cluster_name, show.legend = F) +
      stat_pvalue_manual(
        stats_df, 
        y.position = max(proportions$Freq), 
        step.increase = 0.4,
        label = "p_fc"
        )  + 
      theme(aspect.ratio = 0.8, legend.position="none") +
      scale_y_continuous(expand = expansion(mult = c(0.05, 0.15)))
}
pdf(file = "/mnt/Multimodal_Imaging_Daria/PIPELINE/NEWEST_RESULTS/results/poster/DA_paired.pdf",   
    width = 15, # The width of the plot in inches
    height = 10)
grid.arrange(grobs = plot_list_fc, ncol = 4)
dev.off()
```