---
title: "Identifying cellular communities in diagnostic primary tumors"
author: "LazDaria"
date: "`r Sys.Date()`"
output: 
  workflowr::wflow_html
editor_options:
  chunk_output_type: inline
params: 
  input: "/mnt/Multimodal_Imaging_Daria/Publication/20240811_Zenodo-Upload"
  output: "/mnt/Multimodal_Imaging_Daria/Publication/20240811_R_output"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      error=FALSE,
                      messages=FALSE,
                      warning=FALSE)
```

# Identifying cellular communities in diagnostic primary tumors {#pt-dx}
Following the path of tumor progression, we start with an analysis of the primary site at diagnosis. Therefore, we perform hierarchical clustering of tumor cell composition profiles in primary tumors.

## Load libraries
<details>
   <summary>Load libraries</summary>
   
```{r load-libraries}
library(diffcyt)
library(SingleCellExperiment)
library(CATALYST)
library(dplyr)
library(tidyr)
library(BioQC)
library(ComplexHeatmap)
library(viridis)
library(circlize)
library(ggplot2)
library(gridExtra)
library(ggpubr)
```
</details>

## Read data
First, we load the processed and phenotyped single-cell data from [chapter 4](#phenotyping). 

```{r read-data}
spe <- readRDS(file.path(params$output,"spe_final.rds"))
```

We will then transform the SPE object into a catalyst-compatible object. 

```{r catalyst-comp-object}
keep_rows <- rowData(spe)$use_channel 
keep_cols <- spe$celltype != "other" & spe$tissue=="PT" & (spe$timepoint=="DE")

sce <- SingleCellExperiment(assays=list(counts=assay(spe, "counts")[keep_rows, keep_cols], exprs=assay(spe, "counts")[keep_rows, keep_cols]))

#recode metadata on progression
sce$progress <- factor(spe[keep_rows, keep_cols]$progression)
progress <- recode(sce$progress, `0` = 'NoProg', `1` = 'Prog')
sce$progress <- as.factor(progress)

#recode metadata on MYCN amplification
sce$mna <- factor(spe[keep_rows, keep_cols]$MYCN_amp)
mna <- recode(sce$mna, `1` = 'MNA', `2` = 'het', `0` = 'nMNA', `9` = 'unknown')
sce$mna <- as.factor(mna)

sce$patient <- factor(spe[keep_rows, keep_cols]$patient_id)
sce$fm_id <- spe[keep_rows, keep_cols]$fm_id
sce$tissue <- factor(spe[keep_rows, keep_cols]$tissue)

sce$sample_id <- factor(paste0(sce$progress, "_", sce$mna, "_", sce$patient, "_", sce$fm_id, "_", sce$tissue))

#recode metadata on 1p loss
sce$X1p_loss <- factor(spe[keep_rows, keep_cols]$X1p_loss)
X1p_loss <- recode(sce$X1p_loss, `1` = 'yes', `2` = 'het', `0` = 'no', `9` = 'unknown')
sce$X1p_loss <- as.factor(X1p_loss)

sce$age_diag <- spe[keep_rows, keep_cols]$age_diag

sce$cluster_id <- factor(spe[keep_rows, keep_cols]$celltype_ordered)
sce$metacluster <- factor(spe[keep_rows, keep_cols]$metacluster)

sce$condition <- 'na' 

# Add celltype information to metadata
metadata(sce)$cluster_codes <- data.frame(celltype = factor(spe[keep_rows, keep_cols]$celltype_ordered))
```

## Cluster cellular communities

Next, we cluster proportions of tumor cell types across all diagnostic PT samples (instead of conditions) and plot the result in barplots:

```{r cohort-overview}
set.seed(20231103)
sce_tumor <- sce[,sce$metacluster=="tumor"]

p <- plotAbundances(sce_tumor, k = "celltype", by = "sample", col_clust=T, k_pal=metadata(spe)$color_vectors$col_celltype_ordered, 
                    linkage="ward.D", distance="manhattan") 

p
```

along with the dendogram:

```{r dendogram, fig.height=8, fig.width=7}
#Dendogram
set.seed(20231103)
wide_df <- p$data %>%
  select(-condition) %>%
  pivot_wider(names_from = cluster_id, values_from = Freq)

wide_df <- data.frame(wide_df)
rownames(wide_df) <- wide_df$sample_id
wide_df <- wide_df %>% select(-sample_id)

dist_matrix <- dist(wide_df, method = "manhattan")
hc <- hclust(dist_matrix, method = "ward.D")

plot(hc)
```
We then save the order of clustered samples along with cluster labels.

```{r}
clustered_samples <- sapply(strsplit(levels(p$data$sample_id), "_"), function(x) paste(x[5], x[6], sep = "_"))

#Extract clusters of cellular communities
cellular_communities <- as.data.frame(cutree(hc, k = 2)[levels(p$data$sample_id)])
colnames(cellular_communities) <- c("community_cluster")
cluster_names <- setNames(order(unique(cellular_communities$community_cluster)), c(1:max(cellular_communities$community_cluster)))
cellular_communities$community_cluster <- recode(cellular_communities$community_cluster, !!!cluster_names)
```

## Plot TME cellular profiles 

We will then plot the cellular profiles of the TME for these samples. 

```{r plot-TME, fig.width=12, fig.height=5}
#order samples based on above clustering
sce$sample_id <- factor(paste0(spe[keep_rows, keep_cols]$fm_id, "_", spe[keep_rows, keep_cols]$tissue), levels=clustered_samples)

sce_ME <- sce[,sce$metacluster!="tumor" & sce$metacluster!="mesenchymal"]

p_ME <- plotAbundances(sce_ME, k = "celltype", by = "sample", col_clust=F, k_pal=metadata(spe)$color_vectors$col_celltype_ordered)

p_ME
```

## Compute tumor and TME heterogeneity

We compute tumor and TME intra-sample heterogeneity.

```{r heterogeneity}
proportions <- plotAbundances(sce, k = "celltype", by = "sample")
prop <- proportions$data %>%
  select(-condition) %>%
  pivot_wider(names_from = cluster_id, values_from = Freq)

prop <- data.frame(prop)
rownames(prop) <- prop$sample_id
prop <- prop %>% select(-sample_id)
prop <- t(prop)
rownames(prop) <- levels(sce$cluster_id)

me_clusters <- unique(sce[,sce$metacluster!="tumor" & sce$metacluster!="mesenchymal"]$cluster_id)
me_prop <- prop[rownames(prop)%in%me_clusters,]
me_het <- apply(me_prop,2, entropy)
  
tumor_clusters <- unique(sce[,sce$metacluster=="tumor"]$cluster_id)
tumor_prop <- prop[rownames(prop)%in%tumor_clusters,]
tumor_het <- apply(tumor_prop,2, entropy)
```

## Plot metadata 

Then we plot intra-sample heterogeneity together with sample-specific metadata for the clustered sammples.

```{r plot-metadata}
#access sample-level data
sample_summary <- as.data.frame(colData(sce)) %>%
  group_by(sample_id) %>%
  summarize(mna = first(mna), 
            X1p_loss = first(X1p_loss),
            progress=first(progress),
            age_diag = first(age_diag)
  )

sample_summary <- as.data.frame(sample_summary)
rownames(sample_summary) <- sample_summary$sample_id
sample_summary <- sample_summary %>% select(-sample_id)


#assign color codes
col_code <- c("MNA"=rgb(202,0,32, maxColorValue = 255),
              "het"=rgb(146,197,222, maxColorValue = 255),
              "nMNA"=rgb(5,113,176, maxColorValue = 255),
              "unknown"="white",
              "yes"=rgb(202,0,32, maxColorValue = 255),
              "no"=rgb(5,113,176, maxColorValue = 255),
              "Prog"=rgb(202,0,32, maxColorValue = 255),
              "NoProg"=rgb(5,113,176, maxColorValue = 255)
              )

col_age_diag = colorRamp2(c(min(sample_summary$age_diag, na.rm = TRUE), (quantile(sample_summary$age_diag, 0.95, na.rm = TRUE)-min(sample_summary$age_diag, na.rm = TRUE))/2, quantile(sample_summary$age_diag,0.95, na.rm = TRUE)), viridis(3))
col_me_het = colorRamp2(c(min(me_het, na.rm = TRUE), (max(me_het, na.rm = TRUE)-min(me_het, na.rm = TRUE))/2, max(me_het, na.rm = TRUE)), viridis(3))
col_tumor_het = colorRamp2(c(min(tumor_het, na.rm = TRUE), (max(tumor_het, na.rm = TRUE)-min(tumor_het, na.rm = TRUE))/2, max(tumor_het, na.rm = TRUE)), viridis(3))

# Plot metadata
columnAnnotation(MYCN_amp=sample_summary$mna, col=list(MYCN_amp=col_code)) %v%
  columnAnnotation(X1p_loss=sample_summary$X1p_loss, col=list(X1p_loss=col_code)) %v%
  columnAnnotation(progression=sample_summary$progress, col=list(progression=col_code)) %v%
  columnAnnotation(age=sample_summary$age_diag, col=list(age=col_age_diag)) %v%
  columnAnnotation(TUM_het=tumor_het, col=list(TUM_het=col_tumor_het)) %v%
  columnAnnotation(ME_het=me_het, col=list(ME_het=col_me_het))
  
```

## Plot continous and discrete summary statistics

We then plot summary statistics of continuous and discrete metadata in boxplots and piecharts, respectively.

```{r piecharts}
rownames(cellular_communities) <- sapply(strsplit(as.vector(rownames(cellular_communities)), "_"), function(x) paste(x[length(x) - 1], x[length(x)], sep = "_"))
sample_summary[rownames(cellular_communities),"community_clusters"] <- cellular_communities$community_cluster

discrete_metadata <- c("mna", "X1p_loss", "progress")

discrete_metadata_colors <- list(col_code, col_code, col_code)

plot_list <- list()
for (d in seq_along(discrete_metadata)) {
  sample_meta <- sample_summary[,c(discrete_metadata[d], "community_clusters")]
  sample_meta <- sample_meta[rowSums(sample_meta == "unknown") == 0, ] 
  colnames(sample_meta) <- c("dis_var", "condition")
  
  result <- sample_meta %>%
    group_by(condition, dis_var) %>%
    summarise(Freq = n(), .groups = 'drop')
  
  result[,"dis_var"]<- as.factor(unlist(result[,"dis_var"]))
  
plot_list[[d]] <- ggplot(result, aes(x="", y=Freq, fill=dis_var, group=dis_var)) +
    geom_col(position = "fill") +
    scale_fill_manual(values = discrete_metadata_colors[[d]], name=discrete_metadata[d]) +
    coord_polar("y", start=0)  +  
    facet_grid(.~ condition) +
    theme_void()
}

do.call("grid.arrange", c(plot_list, ncol=1))
```
```{r boxplots}
het_m <- as.vector(unname(me_het[clustered_samples]))
het_t <- as.vector(unname(tumor_het[clustered_samples]))

age <- sample_summary$age_diag

cont_df <- data.frame(cbind(het_m, het_t, age, cellular_communities),row.names=clustered_samples)

plot_list <- list()

for (d in colnames(cont_df)[1:ncol(cont_df)-1]) {

  df <- cont_df[,c(d, "community_cluster")] 

plot_list[[d]] <- ggboxplot(df, x = "community_cluster", y = d,
                color = "community_cluster", palette=c("slategray3", "slategray"),
                add = "jitter")+ 
                stat_compare_means() +
                theme(aspect.ratio = 1.5, legend.position="none") 
}

do.call("grid.arrange", c(plot_list, ncol=3))
```

## Plot metaclusters

Finally, we plot the proportions of metaclusters in the clustered samples.

```{r plot-metaclusters}
sce$cluster_id <- sce$metacluster
metadata(sce)$cluster_codes <- data.frame(celltype = factor(sce$metacluster))
p_all <- plotAbundances(sce, k = "celltype", by = "sample", col_clust=F, k_pal=metadata(spe)$color_vectors$col_metacluster)

p_all
```

## Session Info

<details>
   <summary>SessionInfo</summary>
   
```{r, echo = FALSE}
sessionInfo()
```
</details>
