---
title: "Differential abundance analysis - BM (crtl vs inf.)"
author: "LazDaria"
date: "`r Sys.Date()`"
output: 
  workflowr::wflow_html
editor_options:
  chunk_output_type: inline
params: 
  input: "/mnt/Multimodal_Imaging_Daria/Publication/20240811_Zenodo-Upload"
  output: "/mnt/Multimodal_Imaging_Daria/Publication/20240811_R_output"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      error=FALSE,
                      messages=FALSE,
                      warning=FALSE)
```

# Sample overview - before and after metastasis {#crtl-inf}

Next, to understand how metastasis affects the microenvironment at the metastatic site, we investigate the TME composition in control (non-infiltrated) versus tumor-infiltrated BM samples. 

## Load libraries
<details>
   <summary>Load libraries</summary>
   
```{r load-libraries}
library(diffcyt)
library(SingleCellExperiment)
library(CATALYST)
library(dplyr)
library(tidyr)
library(BioQC)
library(ComplexHeatmap)
library(viridis)
library(circlize)
library(ggplot2)
library(gridExtra)
library(ggpubr)
library(rstatix)
library(gridExtra)
library(gtools)
```
</details>

## Read data

First, we load the processed and phenotyped single-cell data from [chapter 4](#phenotyping). 

```{r read-data}
spe <- readRDS(file.path(params$output,"spe_final.rds"))
```

We will then transform the SPE object into a catalyst-compatible object. 

```{r catalyst-comp-object}
keep_rows <- rowData(spe)$use_channel 
keep_cols <- spe$tissue=="BM" & spe$celltype != "other" & (spe$timepoint=="DE"|spe$timepoint=="NA") & spe$metacluster!="tumor" & spe$metacluster!="mesenchymal" 

sce <- SingleCellExperiment(assays=list(counts=assay(spe, "counts")[keep_rows, keep_cols], exprs=assay(spe, "counts")[keep_rows, keep_cols]))

#recode metadata on progression
sce$progress <- factor(spe[keep_rows, keep_cols]$progression)
progress <- recode(sce$progress, `0` = 'NoProg', `1` = 'Prog')
sce$progress <- as.factor(progress)

#recode metadata on crtl/inf
sce$control <- factor(spe[keep_rows, keep_cols]$control)
control <- recode(sce$control, `yes` = 'crtl', `no` = 'inf.')
sce$control <- as.factor(control)

#recode metadata on MYCN amplification
sce$mna <- factor(spe[keep_rows, keep_cols]$MYCN_amp)
mna <- recode(sce$mna, `1` = 'MNA', `2` = 'het', `0` = 'nMNA', `9` = 'unknown')
sce$mna <- as.factor(mna)

sce$patient <- factor(spe[keep_rows, keep_cols]$patient_id)
sce$fm_id <- spe[keep_rows, keep_cols]$fm_id
sce$tissue <- factor(spe[keep_rows, keep_cols]$tissue)

sce$sample_id <- factor(paste0(sce$progress, "_", sce$mna, "_", sce$patient, "_", sce$fm_id, "_", sce$tissue))

sce$age_diag <- spe[keep_rows, keep_cols]$age_diag
sce$solidity_clipped <- spe[keep_rows, keep_cols]$solidity_clipped

sce$condition <- sce$control

sce$cluster_id <- factor(spe[keep_rows, keep_cols]$celltype_ordered)
sce$metacluster <- factor(spe[keep_rows, keep_cols]$metacluster)

# Add celltype information to metadata
metadata(sce)$cluster_codes <- data.frame(celltype = factor(spe[keep_rows, keep_cols]$celltype_ordered))

# Add metadata
metadata(sce)$experiment_info <- as.data.frame(colData(sce)) %>% group_by(sample_id,patient, mna, condition, progress) %>% 
    summarise(n_cells=n(),.groups = 'drop') %>%
    as.data.frame()

# Define cell_type_markers 
type_markers <- c("MPO_Y89_mean", "CD44_In115_mean", "CD11b_Nd142_mean", "HLA-DR_Nd143_mean", "PRPH_Nd144_mean", "HLA-ABC_Sm147_mean", "CD20_Nd148_mean", "LUM_Sm149_mean", "CD11c_Nd150_mean",
                  "CD24_Eu151_mean", "CD3_Sm152_mean", "CD45_Eu153_mean", "CD8a_Sm154_mean", "GD2_Gd155_mean", "CD34_Gd156_mean", "CD10_Gd158_mean", "CXCR4_Tb159_mean","SOX10_Dy162_mean",
                  "FOXP3_Dy163_mean", "CHGA_Dy164_mean", "GATA3_Er168_mean", "CD56_Er170_mean", "CD4_Yb171_mean", "ELAVL4_Yb174_mean", "CD14_Lu175_mean", "Vimentin_Pt196_mean", "CD15_Bi209_mean")

# Define cell_state_markers 
state_markers <- c("CD274_Gd160_mean", "S100B_Dy161_mean", "CD279_Ho165_mean", "Ki-67_Tm169_mean", "GZMB_Yb173_mean")

# Add to sce
rowData(sce)$marker_class <- ifelse(rownames(sce) %in% type_markers, "type",
                                    ifelse(rownames(sce) %in% state_markers, "state", 
                                    "other"))
```

## Overview celltype abundances

Next, we plot an overview of TME celltype abundances in crtl and infl. BM samples.

```{r cohort-overview, fig.width=16, fig.height=6}
set.seed(20231103)

p <- plotAbundances(sce, k = "celltype", by = "sample", col_clust=T, k_pal=metadata(spe)$color_vectors$col_celltype_ordered, 
                    linkage="ward.D", distance="manhattan") 
p
```

We store the order of the clustered samples to be able to plot corresponding metadata.

```{r store-order}
condition <- c("crtl", "inf.")
clustered_samples <- list()
for (i in seq_along(condition)){
  clustered_samples <- c(clustered_samples, sub(".*_([^_]+_[^_]+$)", "\\1", levels(droplevels(p$data[p$data$condition==condition[i],]$sample_id))))
}
clustered_samples <- as.character(clustered_samples)
```

## Compute TME heterogeneity

We compute TME intra-sample heterogeneity.

```{r heterogeneity}
prop <- p$data %>%
  select(-condition) %>%
  pivot_wider(names_from = cluster_id, values_from = Freq)

prop <- data.frame(prop)
rownames(prop) <- prop$sample_id
prop <- prop %>% select(-sample_id)
prop <- t(prop)
rownames(prop) <- levels(sce$cluster_id)
colnames(prop) <- sub(".*_([^_]+_[^_]+$)", "\\1", colnames(prop))
prop <- prop[,clustered_samples]
het <- apply(prop,2, entropy)
```

## Plot metadata 
Then we plot intra-sample heterogeneity together with sample-specific metadata for the clustered samples.

```{r plot-metadata}
#access sample-level data
sample_summary <- as.data.frame(colData(sce)) %>%
  group_by(sample_id) %>%
  summarize(mna = first(mna), 
            progress=first(progress),
            control = first(control),
            age_diag = first(age_diag)
  )

sample_summary <- as.data.frame(sample_summary)
rownames(sample_summary) <- sub(".*_([^_]+_[^_]+$)", "\\1", sample_summary$sample_id)  
sample_summary <- sample_summary %>% select(-sample_id)
sample_summary <- sample_summary[clustered_samples,]

#assign color codes
col_code <- c("MNA"=rgb(202,0,32, maxColorValue = 255),
              "het"=rgb(146,197,222, maxColorValue = 255),
              "nMNA"=rgb(5,113,176, maxColorValue = 255),
              "Prog"=rgb(202,0,32, maxColorValue = 255),
              "NoProg"=rgb(5,113,176, maxColorValue = 255)
              )

col_age_diag = colorRamp2(c(min(sample_summary$age_diag, na.rm = TRUE), (quantile(sample_summary$age_diag, 0.95, na.rm = TRUE)-min(sample_summary$age_diag, na.rm = TRUE))/2, quantile(sample_summary$age_diag,0.95, na.rm = TRUE)), viridis(3))
col_het = colorRamp2(c(min(het, na.rm = TRUE), (max(het, na.rm = TRUE)-min(het, na.rm = TRUE))/2, max(het, na.rm = TRUE)), viridis(3))

# Plot metadata
columnAnnotation(MYCN_amp=sample_summary$mna, col=list(MYCN_amp=col_code)) %v%
  columnAnnotation(progression=sample_summary$progress, col=list(progression=col_code)) %v%
  columnAnnotation(age=sample_summary$age_diag, col=list(age=col_age_diag)) %v%
  columnAnnotation(het=het, col=list(het=col_het))
  
```

## Plot mean granulocyte roundness per sample

To investigate the shift in granulocyte maturity upon metastasis, we plot the mean granulocyte roundness per sample. 

```{r granulocyte-roundness, fig.width=4, fig.height=5}
sce$sample <- sub(".*_([^_]+_[^_]+$)", "\\1", sce$sample_id)  

df <- aggregate(sce[,sce$metacluster=="granulocyte"]$solidity_clipped, list(sce[,sce$metacluster=="granulocyte"]$sample), mean)
rownames(df) <- df[,1]
df <- df[-c(1)]
colnames (df) <- "y"

df$condition <- sample_summary[rownames(df),]$control

stats_df <- compare_means(y ~ condition,  
                      method='wilcox.test', 
                      data = df)
  
stats_df$p_adj_signif <- stars.pval(stats_df$p.adj)
stats_df$p_adj_signif[stats_df$p_adj_signif==" "] <- "ns"
stats_df$p_long <- paste0(stats_df$p.adj, " (", stats_df$p_adj_signif, ")")

p <- ggboxplot(df, x = "condition", y = "y",
                color = "condition", palette =c("slategray1", "slategray"),
                add.params = list(size = 1.5),
                add = "jitter")+ 
                stat_pvalue_manual(stats_df, label = "p_long", y.position = max(df$y), step.increase = 0.2,) +
                theme(aspect.ratio = 1.5, legend.position="none") 

p
```

# Differential abundance analysis - CRTL vs INF

## Prepare data

We first prepare data for DA analysis using the `diffcyt` package. 

```{r prepare-data}
#list of samples as input to diffcyt methods
df_list <- lapply(unique(sce$sample_id), function(x){as.data.frame(t(counts(sce[,sce$sample_id==x])))})
names(df_list) <- unique(sce$sample_id)

#experiment info
experiment_info <- metadata(sce)$experiment_info
experiment_info <- experiment_info[match(names(df_list),experiment_info$sample_id),]

#marker info
channel_name <- sapply(strsplit(rownames(rowData(spe)[keep_rows,]), "_"), "[[", 2)
marker_name <- sapply(strsplit(rownames(rowData(spe)[keep_rows,]), "_"), "[[", 1)
marker_class <-ifelse(rownames(spe)[keep_rows] %in% type_markers, "type",
                                    ifelse(rownames(spe)[keep_rows] %in% state_markers, "state", 
                                    "other"))
row_data <- data.frame(cbind(channel_name, marker_name), row.names=marker_name)
row_data$marker_class <- as.factor(marker_class)

marker_info <-  row_data

#prepare data into format for diffcyt pipeline 
d_se <- prepareData(df_list, experiment_info, marker_info)
rowData(d_se)$cluster_id <- sce[,rownames(d_se)]$cluster_id 
metadata(d_se)$cluster_codes <- metadata(sce)$cluster_codes

# Transform data
d_se <- transformData(d_se)

# Calculate counts
d_counts <- calcCounts(d_se)
```

Subsequently, we perform the DA analysis comparing celltype proportions between crtl and inf. BM samples. 

```{r da-analysis}
design <- createDesignMatrix(experiment_info, cols_design = "condition")
contrast <- createContrast(c(0, 1)) 


nrow(contrast) == ncol(design)

# Test for differential abundance (DA) of clusters
res_DA <- testDA_edgeR(d_counts,design, contrast, 
                       min_cells=3, 
                       min_samples=round(dim(d_counts)[2]*0.05)
                       )

FDR_cutoff = 0.1
rowData(res_DA)[rowData(res_DA)$p_adj<FDR_cutoff,]
res<- data.frame(rowData(res_DA))
```
## Visualization of DA results

Finally, we visualize DA results in boxplots.

```{r visualize-da, fig.width=12, fig.height=20}
p <- plotAbundances(sce, k = "celltype", by = "cluster_id", shape_by = "sample_id")

comparisons <- c("crtlvsinf.")
ord <- c("crtl", "inf.")

plot_list_fc <- list()

for(i in 1:length(unique(p$data$cluster_id))){
  cluster_name <- unique(p$data$cluster_id)[i]
  proportions <- data.frame(p$data[p$data$cluster_id == cluster_name, c("sample_id", "condition", "Freq")])
  
  p_adj <- c()
  group1 <- c()
  group2 <- c()
  logFC <- c()
  
  for(j in 1:1){
  stats <- res
  p_adj <- c(p_adj, round(stats[rownames(stats)==droplevels(cluster_name), "p_adj"],3))
  logFC <- c(logFC, -1*round(stats[rownames(stats)==droplevels(cluster_name), "logFC"],2))
  group1 <- c(group1, unlist(strsplit(comparisons[j], "vs"))[1])
  group2 <- c(group2, unlist(strsplit(comparisons[j], "vs"))[2])
  }

  stats_df <- data.frame(p_adj=p_adj, group1=group1, group2=group2, logFC=logFC)
  stats_df$p_adj_signif <- stars.pval(stats_df$p_adj)
  stats_df$p_adj_signif[stats_df$p_adj_signif==" "] <- "ns"
  stats_df$p_fc <- paste0(stats_df$logFC, " (", stats_df$p_adj, ")", " (", stats_df$p_adj_signif, ")")
  
  color <- unname(metadata(spe)$color_vectors$col_celltype_ordered[as.character(cluster_name)])
  
  plot_list_fc[[i]] <- ggboxplot(proportions, x = "condition", y = "Freq", order=ord,
              color = "condition", palette =c(color, color), 
              add.params = list(size = 1.5),
              add = "jitter", 
              title=cluster_name, show.legend = F) +
      stat_pvalue_manual(
        stats_df,
        y.position = max(proportions$Freq),
        step.increase = 0.4,
        label = "p_fc"
        )  +
      theme(aspect.ratio = 1.5, legend.position="none") +
      scale_y_continuous(expand = expansion(mult = c(0.05, 0.15)))
}

grid.arrange(grobs = plot_list_fc, ncol = 4)
```
Finally we plot the mean celltype abundance over all samples per condition.

```{r visualize-barplot}
df_normalized <- p$data %>%
  group_by(cluster_id, condition) %>%
  summarise(normalized_count = mean(Freq, na.rm = TRUE)) %>%
  ungroup()

df_normalized$cluster_id <- factor(df_normalized$cluster_id)

ggplot(df_normalized, aes(fill=cluster_id, y=normalized_count, x=condition)) + 
  geom_bar(
    stat="identity",
    show.legend = T) + 
  scale_fill_manual(values = metadata(spe)$color_vectors$col_celltype_ordered) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```
## Session Info

<details>
   <summary>SessionInfo</summary>
   
```{r, echo = FALSE}
sessionInfo()
```
</details>
