---
title: "Differential abundance analysis - PT vs BM"
author: "LazDaria"
date: "`r Sys.Date()`"
output: 
  workflowr::wflow_html
editor_options:
  chunk_output_type: inline
params: 
  input: "/mnt/Multimodal_Imaging_Daria/Publication/processed_data"
  output: "/mnt/Multimodal_Imaging_Daria/Publication/processed_data/data"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      error=FALSE,
                      messages=FALSE,
                      warning=FALSE)
```

# Sample overview - primary versus metastatic site {#pt-bm}
Next, to understand how metastasis affects tumor cell diversity, we investigate the tumor cellular composition in PT samples as compared to BM samples. 

## Load libraries
<details>
   <summary>Load libraries</summary>
   
```{r load-libraries}
library(diffcyt)
library(SingleCellExperiment)
library(CATALYST)
library(dplyr)
library(tidyr)
library(BioQC)
library(ComplexHeatmap)
library(viridis)
library(circlize)
library(ggplot2)
library(gridExtra)
library(ggpubr)
library(rstatix)
library(gridExtra)
library(gtools)
```
</details>

## Read data
First, we load the processed and phenotyped single-cell data from [chapter 4](#phenotyping). 

```{r read-data}
spe <- readRDS(file.path(params$output,"spe_final.rds"))
```

We will then transform the SPE object into a catalyst-compatible object. We decide, whether a `paired` or `unpaired` analysis should be performed. 

```{r catalyst-comp-object}
analysis <- "unpaired"

if(analysis=="paired"){
  BM_patients <- unique(spe[,spe$tissue=="BM" & spe$control=="no" & spe$metacluster=="tumor" & spe$timepoint == "DE"]$study_id )
  PT_patients <- unique(spe[,spe$tissue=="PT"& spe$metacluster=="tumor"]$study_id)
  paired_patients <- intersect(BM_patients, PT_patients)
  keep_cols <- spe$celltype != "other" & spe$timepoint=="DE" & spe$metacluster=="tumor" & spe$study_id%in%paired_patients
}else{
  keep_cols <- spe$celltype != "other" & spe$timepoint=="DE" & spe$metacluster=="tumor"
}

keep_rows <- rowData(spe)$use_channel 

sce <- SingleCellExperiment(assays=list(counts=assay(spe, "counts")[keep_rows, keep_cols], exprs=assay(spe, "counts")[keep_rows, keep_cols]))

#recode metadata on progression
sce$progress <- factor(spe[keep_rows, keep_cols]$progression)
progress <- recode(sce$progress, `0` = 'NoProg', `1` = 'Prog')
sce$progress <- as.factor(progress)

#recode metadata on MYCN amplification
sce$mna <- factor(spe[keep_rows, keep_cols]$MYCN_amp)
mna <- recode(sce$mna, `1` = 'MNA', `2` = 'het', `0` = 'nMNA', `9` = 'unknown')
sce$mna <- as.factor(mna)

sce$patient <- factor(spe[keep_rows, keep_cols]$study_id)
sce$fm_id <- spe[keep_rows, keep_cols]$fm_id
sce$tissue <- factor(spe[keep_rows, keep_cols]$tissue)

sce$sample_id <- factor(paste0(sce$progress, "_", sce$mna, "_", sce$patient, "_", sce$fm_id, "_", sce$tissue))

sce$age_diag <- spe[keep_rows, keep_cols]$age_diag

sce$tissue_mna <- paste0(sce$tissue, "_", sce$mna)
sce$condition <- sce$tissue_mna

sce$cluster_id <- factor(spe[keep_rows, keep_cols]$celltype_ordered)
sce$metacluster <- factor(spe[keep_rows, keep_cols]$metacluster)

# Add celltype information to metadata
metadata(sce)$cluster_codes <- data.frame(celltype = factor(spe[keep_rows, keep_cols]$celltype_ordered))

# Add metadata
tmp <- as.data.frame(colData(sce)) %>% group_by(sample_id,patient, mna, tissue, progress) %>% 
    summarise(n_cells=n(),.groups = 'drop') %>%
    as.data.frame()
colnames(tmp) <- c("sample_id", "patient", "mna", "condition", "progress", "n_cells")
metadata(sce)$experiment_info <- tmp

# Define cell_type_markers 
type_markers <- c("MPO_Y89_mean", "CD44_In115_mean", "CD11b_Nd142_mean", "HLA-DR_Nd143_mean", "PRPH_Nd144_mean", "HLA-ABC_Sm147_mean", "CD20_Nd148_mean", "LUM_Sm149_mean", "CD11c_Nd150_mean",
                  "CD24_Eu151_mean", "CD3_Sm152_mean", "CD45_Eu153_mean", "CD8a_Sm154_mean", "GD2_Gd155_mean", "CD34_Gd156_mean", "CD10_Gd158_mean", "CXCR4_Tb159_mean","SOX10_Dy162_mean",
                  "FOXP3_Dy163_mean", "CHGA_Dy164_mean", "GATA3_Er168_mean", "CD56_Er170_mean", "CD4_Yb171_mean", "ELAVL4_Yb174_mean", "CD14_Lu175_mean", "Vimentin_Pt196_mean", "CD15_Bi209_mean")

# Define cell_state_markers 
state_markers <- c("CD274_Gd160_mean", "S100B_Dy161_mean", "CD279_Ho165_mean", "Ki-67_Tm169_mean", "GZMB_Yb173_mean")

# Add to sce
rowData(sce)$marker_class <- ifelse(rownames(sce) %in% type_markers, "type",
                                    ifelse(rownames(sce) %in% state_markers, "state", 
                                    "other"))
```

## Overview celltype abundances

Next, we plot an overview of celltype abundances in PT and BM samples.

```{r cohort-overview, fig.width=14, fig.height=10}
set.seed(20231103)

p <- plotAbundances(sce, k = "celltype", by = "sample", col_clust=T, k_pal=metadata(spe)$color_vectors$col_celltype_ordered, 
                    linkage="ward.D", distance="manhattan") 
p
```

We save the order of the clustered samples to be able to plot corresponding metadata.

```{r store-order}
condition <- c("PT_nMNA", "PT_MNA", "BM_nMNA", "BM_MNA")
clustered_samples <- list()
clustered_samples_long <- list()
for (i in seq_along(condition)){
  clustered_samples <- c(clustered_samples, sapply(strsplit(levels(droplevels(p$data[p$data$condition==condition[i],]$sample_id)), "_"), function(x) paste(x[5], x[6], sep = "_")))
  clustered_samples_long <- c(clustered_samples_long, levels(droplevels(p$data[p$data$condition==condition[i],]$sample_id)))
}
clustered_samples <- as.character(clustered_samples)
clustered_samples_long <- as.character(clustered_samples_long)
```

## Compute tumor heterogeneity

We compute tumor intra-sample heterogeneity.
```{r heterogeneity}
prop <- p$data %>%
  select(-condition) %>%
  pivot_wider(names_from = cluster_id, values_from = Freq)

prop <- data.frame(prop)
rownames(prop) <- prop$sample_id
prop <- prop %>% select(-sample_id)
prop <- t(prop)
rownames(prop) <- levels(sce$cluster_id)
prop <- prop[,clustered_samples_long]
  
tumor_clusters <- unique(sce[,sce$metacluster=="tumor"]$cluster_id)
tumor_prop <- prop[rownames(prop)%in%tumor_clusters,]
tumor_het <- apply(tumor_prop,2, entropy)
```

## Plot metadata 
Then we plot intra-sample heterogeneity together with sample-specific metadata for the clustered sammples.
```{r plot-metadata}
#access sample-level data
sample_summary <- as.data.frame(colData(sce)) %>%
  group_by(sample_id) %>%
  summarize(mna = first(mna), 
            progress=first(progress),
            age_diag = first(age_diag)
  )

sample_summary <- as.data.frame(sample_summary)
rownames(sample_summary) <- sample_summary$sample_id
sample_summary <- sample_summary %>% select(-sample_id)
sample_summary <- sample_summary[clustered_samples_long,]

#assign color codes
col_code <- c("MNA"=rgb(202,0,32, maxColorValue = 255),
              "het"=rgb(146,197,222, maxColorValue = 255),
              "nMNA"=rgb(5,113,176, maxColorValue = 255),
              "Prog"=rgb(202,0,32, maxColorValue = 255),
              "NoProg"=rgb(5,113,176, maxColorValue = 255)
              )

col_age_diag = colorRamp2(c(min(sample_summary$age_diag, na.rm = TRUE), (quantile(sample_summary$age_diag, 0.95, na.rm = TRUE)-min(sample_summary$age_diag, na.rm = TRUE))/2, quantile(sample_summary$age_diag,0.95, na.rm = TRUE)), viridis(3))
col_tumor_het = colorRamp2(c(min(tumor_het, na.rm = TRUE), (max(tumor_het, na.rm = TRUE)-min(tumor_het, na.rm = TRUE))/2, max(tumor_het, na.rm = TRUE)), viridis(3))

# Plot metadata
columnAnnotation(MYCN_amp=sample_summary$mna, col=list(MYCN_amp=col_code)) %v%
  columnAnnotation(progression=sample_summary$progress, col=list(progression=col_code)) %v%
  columnAnnotation(age=sample_summary$age_diag, col=list(age=col_age_diag)) %v%
  columnAnnotation(TUM_het=tumor_het, col=list(TUM_het=col_tumor_het))
  
```

# Differential abundance analysis - PT vs BM 

## Prepare data
We first prepare data for DA analysis using the `diffcyt` package. 

```{r prepare-data}
sce$condition <- sce$tissue

#list of samples as input to diffcyt methods
df_list <- lapply(unique(sce$sample_id), function(x){as.data.frame(t(counts(sce[,sce$sample_id==x])))})
names(df_list) <- unique(sce$sample_id)

#experiment info
experiment_info <- metadata(sce)$experiment_info
experiment_info <- experiment_info[match(names(df_list),experiment_info$sample_id),]

#marker info
channel_name <- sapply(strsplit(rownames(rowData(spe)[keep_rows,]), "_"), "[[", 2)
marker_name <- sapply(strsplit(rownames(rowData(spe)[keep_rows,]), "_"), "[[", 1)
marker_class <-ifelse(rownames(spe)[keep_rows] %in% type_markers, "type",
                                    ifelse(rownames(spe)[keep_rows] %in% state_markers, "state", 
                                    "other"))
row_data <- data.frame(cbind(channel_name, marker_name), row.names=marker_name)
row_data$marker_class <- as.factor(marker_class)

marker_info <-  row_data

#prepare data into format for diffcyt pipeline 
d_se <- prepareData(df_list, experiment_info, marker_info)
rowData(d_se)$cluster_id <- sce[,rownames(d_se)]$cluster_id 
metadata(d_se)$cluster_codes <- metadata(sce)$cluster_codes

# Transform data
d_se <- transformData(d_se)

# Calculate counts
d_counts <- calcCounts(d_se)
```

Subsequently, we perform the DA analysis comparing celltype proportions between PT and BM samples.

```{r da-analysis}
if (analysis=="unpaired"){
  # Create design matrix
  design <- createDesignMatrix(experiment_info, cols_design = "condition")
  # Create contrast matrix
  contrast <- createContrast(c(0, 1)) 
}else{
  design <- createDesignMatrix(experiment_info, cols_design = c("condition", "patient"))
  contrast <- createContrast(c(0, 1, rep(0, length(unique(experiment_info$patient))-1)))
}

nrow(contrast) == ncol(design)

# Test for differential abundance (DA) of clusters
res_DA <- testDA_edgeR(d_counts,design, contrast, 
                       min_cells=3, 
                       min_samples=round(dim(d_counts)[2]*0.2)
                       )

FDR_cutoff = 0.1
rowData(res_DA)[rowData(res_DA)$p_adj<FDR_cutoff,]
res<- data.frame(rowData(res_DA))
```
## Visualization of DA results
Finally, we visualize DA results in boxplots.

```{r visualize-da, fig.width=12, fig.height=8}
p <- plotAbundances(sce, k = "celltype", by = "cluster_id", shape_by = "sample_id")

comparisons <- c("PTvsBM")
ord <- c("PT", "BM")

plot_list_fc <- list()

for(i in 1:length(unique(p$data$cluster_id))){
  cluster_name <- unique(p$data$cluster_id)[i]
  proportions <- data.frame(p$data[p$data$cluster_id == cluster_name, c("sample_id", "condition", "Freq")])
  
  p_adj <- c()
  group1 <- c()
  group2 <- c()
  logFC <- c()
  
  for(j in 1:1){
  stats <- res
  p_adj <- c(p_adj, round(stats[rownames(stats)==droplevels(cluster_name), "p_adj"],3))
  logFC <- c(logFC, -1*round(stats[rownames(stats)==droplevels(cluster_name), "logFC"],2))
  group1 <- c(group1, unlist(strsplit(comparisons[j], "vs"))[1])
  group2 <- c(group2, unlist(strsplit(comparisons[j], "vs"))[2])
  }

  stats_df <- data.frame(p_adj=p_adj, group1=group1, group2=group2, logFC=logFC)
  stats_df$p_adj_signif <- stars.pval(stats_df$p_adj)
  stats_df$p_adj_signif[stats_df$p_adj_signif==" "] <- "ns"
  stats_df$p_fc <- paste0(stats_df$logFC, " (", stats_df$p_adj, ")", " (", stats_df$p_adj_signif, ")")
  
  color <- unname(metadata(spe)$color_vectors$col_celltype_ordered[as.character(cluster_name)])
  
  plot_list_fc[[i]] <- ggboxplot(proportions, x = "condition", y = "Freq", order=ord,
              color = "condition", palette =c(color, color), 
              add.params = list(size = 1.5),
              add = "jitter", 
              title=cluster_name, show.legend = F) +
      stat_pvalue_manual(
        stats_df,
        y.position = max(proportions$Freq),
        step.increase = 0.4,
        label = "p_fc"
        )  +
      theme(aspect.ratio = 1.5, legend.position="none") +
      scale_y_continuous(expand = expansion(mult = c(0.05, 0.15)))
}

grid.arrange(grobs = plot_list_fc, ncol = 5)
```
Finally we plot the mean celltype abundance over all samples per condition in a piechart.

```{r visualize-piechart}
df_normalized <- p$data %>%
  group_by(cluster_id, condition) %>%
  summarise(normalized_count = mean(Freq, na.rm = TRUE)) %>%
  ungroup()

df_normalized$cluster_id <- factor(df_normalized$cluster_id)

ggplot(df_normalized, aes(x="", y=normalized_count, fill=cluster_id, group=cluster_id)) +
  geom_bar(stat="identity", width=1, color="white") +
  scale_fill_manual(values = metadata(spe)$color_vectors$col_celltype_ordered) +
  coord_polar("y", start=0)  +
  facet_grid(.~ condition) +
  theme_void()
```
## Session Info

<details>
   <summary>SessionInfo</summary>
   
```{r, echo = FALSE}
sessionInfo()
```
</details>
