---
title: "Differential abundance across time in BM"
author: "LazDaria"
date: "`r Sys.Date()`"
output: 
  workflowr::wflow_html
editor_options:
  chunk_output_type: inline
params: 
  input: "/mnt/Multimodal_Imaging_Daria/Publication/processed_data"
  output: "/mnt/Multimodal_Imaging_Daria/Publication/processed_data/data"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      error=FALSE,
                      messages=FALSE,
                      warning=FALSE)
```

# Differential abundance analysis between therapeutic timepoints {#bm-th-da}
Next, we compare celltype abundances between therapeutic timepoints quantitatively.

## Load libraries
<details>
   <summary>Load libraries</summary>
   
```{r load-libraries}
library(diffcyt)
library(SingleCellExperiment)
library(CATALYST)
library(dplyr)
library(tidyr)
library(BioQC)
library(ComplexHeatmap)
library(viridis)
library(circlize)
library(ggplot2)
library(ggpubr)
library(rstatix)
library(gridExtra)
library(gtools)
library(edgeR)
```
</details>

## Read data
First, we load the processed and phenotyped single-cell data from [chapter 4](#phenotyping). 

```{r read-data}
spe <- readRDS(file.path(params$output,"spe_final.rds"))
```

We will then transform the SPE object into a catalyst-compatible object. Decide whether analysis should be performed in `nMNA` or `MNA`samples only with setting the `reduce_to` parameter, and whether tumor (`tumor`) or TME (`TME`) celltypes should be compared. We combine samples from RE1 and RE2 to increase statistical power.

```{r catalyst-comp-object}
reduce_to <- "none" #set to nMNA, MNA or none
compartment <- "TME"

keep_rows <- rowData(spe)$use_channel 

if(compartment=="tumor"){
  keep_cols <- spe$celltype != "other" & spe$tissue=="BM" & spe$metacluster=="tumor" & spe$control=="no"
}else{
  keep_cols <- spe$celltype != "other" & spe$tissue=="BM" & spe$metacluster!="tumor" & spe$metacluster!="mesenchymal" & spe$control=="no"
}

if (reduce_to=="nMNA"){keep_cols <- keep_cols & spe$MYCN_amp == 0}
if (reduce_to=="MNA"){keep_cols <- keep_cols & spe$MYCN_amp == 1}

sce <- SingleCellExperiment(assays=list(counts=assay(spe, "counts")[keep_rows, keep_cols], exprs=assay(spe, "counts")[keep_rows, keep_cols]))

sce$fm_id <- spe[keep_rows, keep_cols]$fm_id
sce$tissue <- factor(spe[keep_rows, keep_cols]$tissue)
sce$sample_id <- factor(paste0(sce$fm_id, "_", sce$tissue))

#combining RE1 and RE2
sce$timepoint <- factor(spe[keep_rows, keep_cols]$timepoint)
sce[,sce$timepoint=="RE1"]$timepoint <- "RE2"
sce$condition <- factor(sce$timepoint)

#solidity
sce$solidity_clipped <- spe[keep_rows, keep_cols]$solidity_clipped

sce$cluster_id <- factor(spe[keep_rows, keep_cols]$celltype_ordered)
sce$metacluster <- factor(spe[keep_rows, keep_cols]$metacluster)
metadata(sce)$cluster_codes <- data.frame(celltype = factor(spe[keep_rows, keep_cols]$celltype_ordered))

# Add metadata
metadata(sce)$experiment_info <- as.data.frame(colData(sce)) %>% group_by(sample_id, condition) %>% 
    summarise(n_cells=n(),.groups = 'drop') %>%
    as.data.frame()

# Define cell_type_markers 
type_markers <- c("MPO_Y89_mean", "CD44_In115_mean", "CD11b_Nd142_mean", "HLA-DR_Nd143_mean", "PRPH_Nd144_mean", "HLA-ABC_Sm147_mean", "CD20_Nd148_mean", "LUM_Sm149_mean", "CD11c_Nd150_mean",
                  "CD24_Eu151_mean", "CD3_Sm152_mean", "CD45_Eu153_mean", "CD8a_Sm154_mean", "GD2_Gd155_mean", "CD34_Gd156_mean", "CD10_Gd158_mean", "CXCR4_Tb159_mean","SOX10_Dy162_mean",
                  "FOXP3_Dy163_mean", "CHGA_Dy164_mean", "GATA3_Er168_mean", "CD56_Er170_mean", "CD4_Yb171_mean", "ELAVL4_Yb174_mean", "CD14_Lu175_mean", "Vimentin_Pt196_mean", "CD15_Bi209_mean")

# Define cell_state_markers 
state_markers <- c("CD274_Gd160_mean", "S100B_Dy161_mean", "CD279_Ho165_mean", "Ki-67_Tm169_mean", "GZMB_Yb173_mean")

# Add to sce
rowData(sce)$marker_class <- ifelse(rownames(sce) %in% type_markers, "type",
                                    ifelse(rownames(sce) %in% state_markers, "state", 
                                    "other"))
```

# Differential abundance analysis - BM during therapy

## Prepare data
We first prepare data for DA analysis using the `diffcyt` package. 

```{r prepare-data}
#list of samples as input to diffcyt methods
df_list <- lapply(unique(sce$sample_id), function(x){as.data.frame(t(counts(sce[,sce$sample_id==x])))})
names(df_list) <- unique(sce$sample_id)

#experiment info
experiment_info <- metadata(sce)$experiment_info
experiment_info <- experiment_info[match(names(df_list),experiment_info$sample_id),]

#marker info
channel_name <- sapply(strsplit(rownames(rowData(spe)[keep_rows,]), "_"), "[[", 2)
marker_name <- sapply(strsplit(rownames(rowData(spe)[keep_rows,]), "_"), "[[", 1)
marker_class <-ifelse(rownames(spe)[keep_rows] %in% type_markers, "type",
                                    ifelse(rownames(spe)[keep_rows] %in% state_markers, "state", 
                                    "other"))
row_data <- data.frame(cbind(channel_name, marker_name), row.names=marker_name)
row_data$marker_class <- as.factor(marker_class)

marker_info <-  row_data

#prepare data into format for diffcyt pipeline 
d_se <- prepareData(df_list, experiment_info, marker_info)
rowData(d_se)$cluster_id <- sce[,rownames(d_se)]$cluster_id 
metadata(d_se)$cluster_codes <- metadata(sce)$cluster_codes

# Transform data
d_se <- transformData(d_se)

# Calculate counts
d_counts <- calcCounts(d_se)
```

Subsequently, we perform the DA analysis comparing celltype proportions between different therapeutical timepoints.

```{r da-analysis}
design <- model.matrix(~0+condition, experiment_info) 
colnames(design) <- c("DE", "RE2", "REL")

contrasts <- makeContrasts(DEvsRE2 = DE-RE2,
                           DEvsREL = DE-REL,
                           RE2vsREL = RE2-REL,
                           levels = design)

res <- list()

for (i in 1:ncol(contrasts)){

    res_DA <- testDA_edgeR(d_counts,design, contrasts[,i], 
                       min_cells=1, 
                       min_samples=1
                       )
    
    res <- append(res, list(data.frame(rowData(res_DA))))
}
```

## Visualization of DA results
Finally, we visualize DA results in boxplots.

```{r visualize-da, fig.width=12, fig.height=20}
p <- plotAbundances(sce, k = "celltype", by = "cluster_id", shape_by = "sample_id")
comparisons <-colnames(contrasts)

plot_list_fc <- list()
for(i in 1:length(unique(p$data$cluster_id))){
  cluster_name <- unique(p$data$cluster_id)[i]
  proportions <- data.frame(p$data[p$data$cluster_id == cluster_name, c("sample_id", "condition", "Freq")])
  
  p_adj <- c()
  group1 <- c()
  group2 <- c()
  logFC <- c()
  
  for(j in 1:ncol(contrasts)){
  stats <- res[[j]]
  p_adj <- c(p_adj, round(stats[rownames(stats)==droplevels(cluster_name), "p_adj"],3))
  logFC <- c(logFC, -1*round(stats[rownames(stats)==droplevels(cluster_name), "logFC"],2))
  group1 <- c(group1, unlist(strsplit(comparisons[j], "vs"))[1])
  group2 <- c(group2, unlist(strsplit(comparisons[j], "vs"))[2])
  }

  stats_df <- data.frame(p_adj=p_adj, group1=group1, group2=group2, logFC=logFC)
  stats_df$p_adj_signif <- stars.pval(stats_df$p_adj)
  stats_df$p_adj_signif[stats_df$p_adj_signif==" "] <- "ns"
  stats_df$p_fc <- paste0(stats_df$logFC, " (", stats_df$p_adj, ")", " (", stats_df$p_adj_signif, ")")

  color <- unname(metadata(spe)$color_vectors$col_celltype_ordered[as.character(cluster_name)]) 
  
  plot_list_fc[[i]] <- ggboxplot(proportions, x = "condition", y = "Freq", order=c("DE", "RE2", "REL"),
              color = "condition", palette =c(color, color,color), 
              add.params = list(size = 1.5),
              add = "jitter", 
              title=cluster_name, show.legend = F) +
      stat_pvalue_manual(
        stats_df,
        y.position = max(proportions$Freq),
        step.increase = 0.2,
        label = "p_fc"
        )  +
      theme(aspect.ratio = 1.75, legend.position="none") +
      scale_y_continuous(expand = expansion(mult = c(0.05, 0.15)))
}

grid.arrange(grobs = plot_list_fc, ncol = 5)
```

Then, we plot the mean abundances of celltypes over all samples per condition.

```{r mean-abundance, fig.width=10}
df_normalized <- p$data %>%
  group_by(cluster_id, condition) %>%
  summarise(normalized_count = mean(Freq, na.rm = TRUE), std=sd(Freq, na.rm = TRUE)) %>%
  ungroup()

df_normalized$cluster_id <- factor(df_normalized$cluster_id)

ggplot(df_normalized, aes(fill=cluster_id, y=normalized_count, x=condition)) + 
  geom_bar(
    stat="identity",
    position="dodge",
    show.legend = T) + 
  scale_fill_manual(values = metadata(spe)$color_vectors$col_celltype_ordered) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

## Plot mean granulocyte roundness per sample
To investigate the shift in granulocyte maturity during therapy, we plot the mean granulocyte roundness per sample and timepoint. 

```{r granulocyte-roundness, fig.width=4, fig.height=5}
if (compartment == "TME"){
  sample_summary <- as.data.frame(colData(sce)) %>%
  group_by(sample_id) %>%
  summarize(condition=first(condition)
  )
  
  sample_summary <- as.data.frame(sample_summary)
  rownames(sample_summary) <- sample_summary$sample_id
  sample_summary <- sample_summary %>% select(-sample_id)
  
  sce$sample <- sce$sample_id
  
  df <- aggregate(sce[,sce$metacluster=="granulocyte"]$solidity_clipped, list(sce[,sce$metacluster=="granulocyte"]$sample), mean)
  rownames(df) <- df[,1]
  df <- df[-c(1)]
  colnames (df) <- "y"
  
  df$condition <- sample_summary[rownames(df),"condition"]
  
  stats_df <- compare_means(y ~ condition,  
                        p.adjust.method = "BH", 
                        method='wilcox.test', 
                        data = df)
    
  stats_df$p_adj_signif <- stars.pval(stats_df$p.adj)
  stats_df$p_adj_signif[stats_df$p_adj_signif==" "] <- "ns"
  stats_df$p_long <- paste0(stats_df$p.adj, " (", stats_df$p_adj_signif, ")")
  
  p <- ggboxplot(df, x = "condition", y = "y",
                  color = "condition", palette =c("skyblue2", "slategray3", "slategray"),
                  add.params = list(size = 1.5),
                  add = "jitter")+ 
                  stat_pvalue_manual(stats_df, label = "p_long", y.position = max(df$y), step.increase = 0.2,) +
                  theme(aspect.ratio = 1.5, legend.position="none") 
  p
}
```

## Session Info

<details>
   <summary>SessionInfo</summary>
   
```{r, echo = FALSE}
sessionInfo()
```
</details>
