---
title: "Differential abundance across time in paired BM samples"
author: "LazDaria"
date: "`r Sys.Date()`"
output: 
  workflowr::wflow_html
editor_options:
  chunk_output_type: inline
params: 
  input: "/mnt/Multimodal_Imaging_Daria/Publication/20240811_Zenodo-Upload"
  output: "/mnt/Multimodal_Imaging_Daria/Publication/20240811_R_output"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      error=FALSE,
                      messages=FALSE,
                      warning=FALSE)
```

# Differential abundance analysis between therapeutic timepoints in paired BM {#bm-th-da-paired}

Next, we compare celltype abundances between therapeutic timepoints in paired BM samples, quantitatively.

## Load libraries
<details>
   <summary>Load libraries</summary>
   
```{r load-libraries}
library(diffcyt)
library(SingleCellExperiment)
library(CATALYST)
library(dplyr)
library(tidyr)
library(BioQC)
library(ComplexHeatmap)
library(viridis)
library(circlize)
library(ggplot2)
library(ggpubr)
library(rstatix)
library(gridExtra)
library(gtools)
library(edgeR)
```
</details>

## Read data

First, we load the processed and phenotyped single-cell data from [chapter 4](#phenotyping). 

```{r read-data}
spe <- readRDS(file.path(params$output,"spe_final.rds"))
```

We will then transform the SPE object into a catalyst-compatible object. Decide whether tumor (`tumor`) or TME (`TME`) celltypes should be analysed and which timepoint (`RE1`, `RE2`) should be compared to paired diagnostic samples. 

```{r catalyst-comp-object}
timepoint <- "RE2"
compartment <- "TME"

if (compartment=="tumor"){
  BM_patients_1 <- unique(spe[,spe$tissue=="BM" & spe$control=="no" & spe$metacluster == "tumor" & spe$timepoint == "DE"]$patient_id )
  BM_patients_2 <- unique(spe[,spe$tissue=="BM" & spe$control=="no" & spe$metacluster == "tumor" & spe$timepoint == timepoint]$patient_id )
  paired_patients <- intersect(BM_patients_1, BM_patients_2)
  keep_cols <- spe$celltype != "other" & spe$tissue=="BM" & spe$metacluster=="tumor" & spe$control=="no" & (spe$timepoint=="DE" | spe$timepoint==timepoint) & spe$patient_id%in%paired_patients 
}else{
  BM_patients_1 <- unique(spe[,spe$tissue=="BM" & spe$control=="no" & spe$timepoint == "DE"]$patient_id )
  BM_patients_2 <- unique(spe[,spe$tissue=="BM" & spe$control=="no" & spe$timepoint == timepoint]$patient_id )
  paired_patients <- intersect(BM_patients_1, BM_patients_2)
  keep_cols <- spe$celltype != "other" & spe$tissue=="BM" & spe$metacluster!="tumor" & spe$metacluster!="mesenchymal" & spe$control=="no" & (spe$timepoint=="DE" | spe$timepoint==timepoint) & spe$patient_id%in%paired_patients 
}

keep_rows <- rowData(spe)$use_channel 

sce <- SingleCellExperiment(assays=list(counts=assay(spe, "counts")[keep_rows, keep_cols], exprs=assay(spe, "counts")[keep_rows, keep_cols]))

sce$fm_id <- spe[keep_rows, keep_cols]$fm_id
sce$tissue <- factor(spe[keep_rows, keep_cols]$tissue)
sce$sample_id <- factor(paste0(sce$fm_id, "_", sce$tissue))
sce$patient <- factor(spe[keep_rows, keep_cols]$patient_id)
sce$timepoint <- factor(spe[keep_rows, keep_cols]$timepoint)
sce$condition <- factor(sce$timepoint)

sce$cluster_id <- factor(spe[keep_rows, keep_cols]$celltype_ordered)
sce$metacluster <- factor(spe[keep_rows, keep_cols]$metacluster)
metadata(sce)$cluster_codes <- data.frame(celltype = factor(spe[keep_rows, keep_cols]$celltype_ordered))

# Add metadata
metadata(sce)$experiment_info <- as.data.frame(colData(sce)) %>% group_by(sample_id, condition, patient) %>% 
    summarise(n_cells=n(),.groups = 'drop') %>%
    as.data.frame()

# Define cell_type_markers 
type_markers <- c("MPO_Y89_mean", "CD44_In115_mean", "CD11b_Nd142_mean", "HLA-DR_Nd143_mean", "PRPH_Nd144_mean", "HLA-ABC_Sm147_mean", "CD20_Nd148_mean", "LUM_Sm149_mean", "CD11c_Nd150_mean",
                  "CD24_Eu151_mean", "CD3_Sm152_mean", "CD45_Eu153_mean", "CD8a_Sm154_mean", "GD2_Gd155_mean", "CD34_Gd156_mean", "CD10_Gd158_mean", "CXCR4_Tb159_mean","SOX10_Dy162_mean",
                  "FOXP3_Dy163_mean", "CHGA_Dy164_mean", "GATA3_Er168_mean", "CD56_Er170_mean", "CD4_Yb171_mean", "ELAVL4_Yb174_mean", "CD14_Lu175_mean", "Vimentin_Pt196_mean", "CD15_Bi209_mean")

# Define cell_state_markers 
state_markers <- c("CD274_Gd160_mean", "S100B_Dy161_mean", "CD279_Ho165_mean", "Ki-67_Tm169_mean", "GZMB_Yb173_mean")

# Add to sce
rowData(sce)$marker_class <- ifelse(rownames(sce) %in% type_markers, "type",
                                    ifelse(rownames(sce) %in% state_markers, "state", 
                                    "other"))
```

# Differential abundance analysis - paired sequential BM samples

## Prepare data

We first prepare data for DA analysis using the `diffcyt` package. 

```{r prepare-data}
#list of samples as input to diffcyt methods
df_list <- lapply(unique(sce$sample_id), function(x){as.data.frame(t(counts(sce[,sce$sample_id==x])))})
names(df_list) <- unique(sce$sample_id)

#experiment info
experiment_info <- metadata(sce)$experiment_info
experiment_info <- experiment_info[match(names(df_list),experiment_info$sample_id),]

#marker info
channel_name <- sapply(strsplit(rownames(rowData(spe)[keep_rows,]), "_"), "[[", 2)
marker_name <- sapply(strsplit(rownames(rowData(spe)[keep_rows,]), "_"), "[[", 1)
marker_class <-ifelse(rownames(spe)[keep_rows] %in% type_markers, "type",
                                    ifelse(rownames(spe)[keep_rows] %in% state_markers, "state", 
                                    "other"))
row_data <- data.frame(cbind(channel_name, marker_name), row.names=marker_name)
row_data$marker_class <- as.factor(marker_class)

marker_info <-  row_data

#prepare data into format for diffcyt pipeline 
d_se <- prepareData(df_list, experiment_info, marker_info)
rowData(d_se)$cluster_id <- sce[,rownames(d_se)]$cluster_id 
metadata(d_se)$cluster_codes <- metadata(sce)$cluster_codes

# Transform data
d_se <- transformData(d_se)

# Calculate counts
d_counts <- calcCounts(d_se)
```

Subsequently, we perform the DA analysis comparing celltype proportions between paired BM samples from diagnosis and `r timepoint`.

```{r da-analysis}
design <- createDesignMatrix(experiment_info, cols_design = c("condition", "patient")) 
contrast <- createContrast(c(0, 1, rep(0, length(unique(experiment_info$patient))-1))) #paired
nrow(contrast) == ncol(design)

if (compartment=="tumor"){
  min_sam = 2
}else{
  min_sam = 1
}

# Test for differential abundance (DA) of clusters
res_DA <- testDA_edgeR(d_counts,design, contrast, 
                       min_cells=3, 
                       min_samples=min_sam
                       )


FDR_cutoff = 0.1
rowData(res_DA)[rowData(res_DA)$p_adj<FDR_cutoff,]
res<- data.frame(rowData(res_DA))
```

## Visualization of DA results

Finally, we visualize DA results in boxplots.

```{r visualize-da, fig.width=12, fig.height=20}
p <- plotAbundances(sce, k = "celltype", by = "cluster_id", shape_by = "sample_id")

comparisons <- c(paste0("DEvs", timepoint))
ord <- c("DE", timepoint)

plot_list_values <- list()
plot_list_stars <- list()
plot_list_fc <- list()
for(i in 1:length(unique(p$data$cluster_id))){
  cluster_name <- unique(p$data$cluster_id)[i]
  proportions <- data.frame(p$data[p$data$cluster_id == cluster_name, c("sample_id", "condition", "Freq")])
  
  p_adj <- c()
  group1 <- c()
  group2 <- c()
  logFC <- c()
  
  for(j in 1:1){
  stats <- res
  p_adj <- c(p_adj, round(stats[rownames(stats)==droplevels(cluster_name), "p_adj"],3))
  logFC <- c(logFC, -1*round(stats[rownames(stats)==droplevels(cluster_name), "logFC"],2))
  group1 <- c(group1, unlist(strsplit(comparisons[j], "vs"))[1])
  group2 <- c(group2, unlist(strsplit(comparisons[j], "vs"))[2])
  }

  stats_df <- data.frame(p_adj=p_adj, group1=group1, group2=group2, logFC=logFC)
  stats_df$p_adj_signif <- stars.pval(stats_df$p_adj)
  stats_df$p_adj_signif[stats_df$p_adj_signif==" "] <- "ns"
  stats_df$p_fc <- paste0(stats_df$logFC, " (", stats_df$p_adj, ")", " (", stats_df$p_adj_signif, ")")

  color <- unname(metadata(spe)$color_vectors$col_celltype_ordered[as.character(cluster_name)]) 
  
  plot_list_fc[[i]] <- ggboxplot(proportions, x = "condition", y = "Freq", order=ord,
              color = "condition", palette =c(color, color), 
              add.params = list(size = 1.5),
              add = "jitter", 
              title=cluster_name, show.legend = F) +
      stat_pvalue_manual(
        stats_df,
        y.position = max(proportions$Freq),
        step.increase = 0.4,
        label = "p_fc"
        )  +
      theme(aspect.ratio = 1.5, legend.position="none") +
      scale_y_continuous(expand = expansion(mult = c(0.05, 0.15)))
}

grid.arrange(grobs = plot_list_fc, ncol = 6)
```

We will plot the logFC for the comparisons between conditions in a barplot.

```{r barplot-logFC, fig.width=7, fig.height=8}
df <- res[, c("logFC", "p_adj")]
df$p_adj_signif <- stars.pval(df$p_adj)
df$p_adj_signif[df$p_adj_signif==" "] <- "ns"
df$cluster <- rownames(df)

p <- ggplot(df, aes(fill=cluster, y=logFC, x=cluster)) + 
  ggtitle(colnames(contrasts)[i]) +
  geom_bar(
    stat="identity",
    position="dodge",
    show.legend = F) + 
  scale_fill_manual(values = metadata(spe)$color_vectors$col_celltype_ordered) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  geom_text(aes(label = p_adj_signif), vjust = 0)

p
```

## Session Info

<details>
   <summary>SessionInfo</summary>
   
```{r, echo = FALSE}
sessionInfo()
```
</details>
