---
title: "Investigate therapeutic dynamics in metastatic bone marrow"
author: "LazDaria"
date: "`r Sys.Date()`"
output: 
  workflowr::wflow_html
editor_options:
  chunk_output_type: inline
params: 
  input: "/mnt/Multimodal_Imaging_Daria/Publication/20240811_Zenodo-Upload"
  output: "/mnt/Multimodal_Imaging_Daria/Publication/20240811_R_output"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      error=FALSE,
                      messages=FALSE,
                      warning=FALSE)
```

# Investigation of cell composition dynamics in metastatic bone marrow {#bm-th}

Here, we investigate phenotypic changes in response to therapy in the metastatic bone marrow.

## Load libraries
<details>
   <summary>Load libraries</summary>
   
```{r load-libraries}
library(diffcyt)
library(SingleCellExperiment)
library(CATALYST)
library(dplyr)
library(tidyr)
library(BioQC)
library(ComplexHeatmap)
library(viridis)
library(circlize)
library(ggplot2)
library(gridExtra)
library(ggpubr)
library(gtools)
```
</details>

## Read data

First, we load the processed and phenotyped single-cell data from [chapter 4](#phenotyping). 

```{r read-data}
spe <- readRDS(file.path(params$output,"spe_final.rds"))
```
## Hierarchical clustering of each timepoint 

We first perform hierarchcial clustering for each timepoint, separately. 

```{r}
samples_with_tumor_cells <- unique(spe[,spe$tissue=="BM"& spe$metacluster=="tumor"]$sample_id)
keep_rows <- rowData(spe)$use_channel 
```

Therefore, we define a function for converting the SPE objects into catalayst compatible objects, for hierarhichal clustering and plotting of barplots and dendograms. 
```{r functions}
catalyst_convert <- function(timepoint){
  keep_cols <- spe$celltype != "other" & spe$tissue=="BM" & spe$sample_id%in%samples_with_tumor_cells & spe$control=="no" & spe$timepoint == timepoint
  sce_object <- SingleCellExperiment(assays=list(counts=assay(spe, "counts")[keep_rows, keep_cols], exprs=assay(spe, "counts")[keep_rows, keep_cols]))
  
  sce_object$fm_id <- spe[keep_rows, keep_cols]$fm_id
  sce_object$tissue <- factor(spe[keep_rows, keep_cols]$tissue)
  sce_object$sample_id <- factor(paste0(sce_object$fm_id, "_", sce_object$tissue))
  
  sce_object$condition <- "na" 
  
  sce_object$cluster_id <- factor(spe[keep_rows, keep_cols]$celltype_ordered)
  sce_object$metacluster <- factor(spe[keep_rows, keep_cols]$metacluster)
  metadata(sce_object)$cluster_codes <- data.frame(celltype = factor(spe[keep_rows, keep_cols]$celltype_ordered))
  
  return(sce_object)
}

cluster_barplot <- function(sce_obj){
  set.seed(20231103)
  sce_tumor <- sce_obj[,sce_obj$metacluster=="tumor"]
  
  p <- plotAbundances(sce_tumor, k = "celltype", by = "sample", col_clust=T, k_pal=metadata(spe)$color_vectors$col_celltype_ordered, 
                      linkage="ward.D", distance="manhattan") 
  
  clustered_samples <- levels(p$data$sample_id)
  return(list(plot = p, clustered_sample_list = clustered_samples))
}

plot_dend <- function(plot_object){
  set.seed(20231103)
  wide_df <- plot_object$data %>%
    select(-condition) %>%
    pivot_wider(names_from = cluster_id, values_from = Freq)
  
  wide_df <- data.frame(wide_df)
  rownames(wide_df) <- wide_df$sample_id
  wide_df <- wide_df %>% select(-sample_id)
  
  dist_matrix <- dist(wide_df, method = "manhattan")
  hc <- hclust(dist_matrix, method = "ward.D")
  
  return(hc)
}
```


### Diagnostic BM samples

We transform the SPE object into a catalyst-compatible object, cluster proportions of tumor cell types across all diagnostic BM samples and plot the result in barplots:

```{r catalyst-comp-cluster-DE}
p_DE <- cluster_barplot(catalyst_convert("DE")) 
clustered_samples_DE <- p_DE$clustered_sample_list
p_DE$plot
```

along with the dendogram:

```{r dendogram-de, fig.height=6, fig.width=10}
dend <- plot_dend(p_DE$plot)
plot(dend)
```

### Mid-induction chemotherapy (RE1) BM samples

We repeat the same for the other therapeutic timepoints.

```{r catalyst-comp-cluster-RE1}
p_RE1 <- cluster_barplot(catalyst_convert("RE1")) 
clustered_samples_RE1 <- p_RE1$clustered_sample_list
p_RE1$plot
```


```{r dendogram-re1, fig.height=6, fig.width=5}
dend <- plot_dend(p_RE1$plot)
plot(dend)
```

### End of induction chemotherapy (RE2) BM samples

```{r catalyst-comp-cluster-RE2}
p_RE2 <- cluster_barplot(catalyst_convert("RE2")) 
clustered_samples_RE2 <- p_RE2$clustered_sample_list
p_RE2$plot
```

```{r dendogram-re2, fig.height=6, fig.width=5}
dend <- plot_dend(p_RE2$plot)
plot(dend)
```

### Relapse (REL) BM samples

```{r catalyst-comp-cluster-REL}
p_REL <- cluster_barplot(catalyst_convert("REL")) 
clustered_samples_REL <- p_REL$clustered_sample_list
p_REL$plot
```


```{r dendogram-rel, fig.height=6, fig.width=5}
dend <- plot_dend(p_REL$plot)
plot(dend)
```

### Merging all inf. BM samples 

Next, we merge the cluster order of all BM samples to be able to plot the corresponding metadata. 
Let's first convert the SPE object of all inf. BM samples into a CATALYST-compatible object.

```{r catalyst-comp-object-all}
keep_cols <- spe$celltype != "other" & spe$tissue=="BM" & spe$sample_id%in%samples_with_tumor_cells & spe$control=="no"

sce <- SingleCellExperiment(assays=list(counts=assay(spe, "counts")[keep_rows, keep_cols], exprs=assay(spe, "counts")[keep_rows, keep_cols]))

#recode metadata on progression
sce$progress <- factor(spe[keep_rows, keep_cols]$progression)
progress <- recode(sce$progress, `0` = 'NoProg', `1` = 'Prog')
sce$progress <- as.factor(progress)

#recode metadata on MYCN amplification
sce$mna <- factor(spe[keep_rows, keep_cols]$MYCN_amp)
mna <- recode(sce$mna, `1` = 'MNA', `2` = 'het', `0` = 'nMNA', `9` = 'unknown')
sce$mna <- as.factor(mna)

sce$patient <- factor(spe[keep_rows, keep_cols]$patient_id)
sce$fm_id <- spe[keep_rows, keep_cols]$fm_id
sce$tissue <- factor(spe[keep_rows, keep_cols]$tissue)

sce$sample_id <- factor(paste0(sce$fm_id, "_", sce$tissue))

sce$timepoint <- factor(spe[keep_rows, keep_cols]$timepoint)
sce$age_diag <- spe[keep_rows, keep_cols]$age_diag
sce$condition <- sce$timepoint 

sce$cluster_id <- factor(spe[keep_rows, keep_cols]$celltype_ordered)
sce$metacluster <- factor(spe[keep_rows, keep_cols]$metacluster)
metadata(sce)$cluster_codes <- data.frame(celltype = factor(spe[keep_rows, keep_cols]$celltype_ordered))

# Add metadata
metadata(sce)$experiment_info <- as.data.frame(colData(sce)) %>% group_by(sample_id,patient, mna, condition, progress) %>% 
    summarise(n_cells=n(),.groups = 'drop') %>%
    as.data.frame()

# Define cell_type_markers 
type_markers <- c("MPO_Y89_mean", "CD44_In115_mean", "CD11b_Nd142_mean", "HLA-DR_Nd143_mean", "PRPH_Nd144_mean", "HLA-ABC_Sm147_mean", "CD20_Nd148_mean", "LUM_Sm149_mean", "CD11c_Nd150_mean",
                  "CD24_Eu151_mean", "CD3_Sm152_mean", "CD45_Eu153_mean", "CD8a_Sm154_mean", "GD2_Gd155_mean", "CD34_Gd156_mean", "CD10_Gd158_mean", "CXCR4_Tb159_mean","SOX10_Dy162_mean",
                  "FOXP3_Dy163_mean", "CHGA_Dy164_mean", "GATA3_Er168_mean", "CD56_Er170_mean", "CD4_Yb171_mean", "ELAVL4_Yb174_mean", "CD14_Lu175_mean", "Vimentin_Pt196_mean", "CD15_Bi209_mean")

# Define cell_state_markers 
state_markers <- c("CD274_Gd160_mean", "S100B_Dy161_mean", "CD279_Ho165_mean", "Ki-67_Tm169_mean", "GZMB_Yb173_mean")

# Add to spe
rowData(sce)$marker_class <- ifelse(rownames(sce) %in% type_markers, "type",
                                    ifelse(rownames(sce) %in% state_markers, "state", 
                                    "other"))
```

Then we merge cluster labels for all timepoints.

```{r merge-order}
clustered_samples <- c(clustered_samples_DE, clustered_samples_RE1, clustered_samples_RE2, clustered_samples_REL)
sce$sample_id <- factor(sce$sample_id, levels=clustered_samples)
sce_tumor <- sce[,sce$metacluster=="tumor"]
sce_ME <- sce[,sce$metacluster!="tumor" & sce$metacluster!="mesenchymal"]
```

We compute tumor and TME intra-sample heterogeneity.

```{r heterogeneity}
proportions <- plotAbundances(sce, k = "celltype", by = "sample")
prop <- proportions$data %>%
  select(-condition) %>%
  pivot_wider(names_from = cluster_id, values_from = Freq)

prop <- data.frame(prop)
rownames(prop) <- prop$sample_id
prop <- prop %>% select(-sample_id)
prop <- t(prop)
rownames(prop) <- levels(sce$cluster_id)

me_clusters <- unique(sce[,sce$metacluster!="tumor" & sce$metacluster!="mesenchymal"]$cluster_id)
me_prop <- prop[rownames(prop)%in%me_clusters,]
me_het <- apply(me_prop,2, entropy)
  
tumor_clusters <- unique(sce[,sce$metacluster=="tumor"]$cluster_id)
tumor_prop <- prop[rownames(prop)%in%tumor_clusters,]
tumor_het <- apply(tumor_prop,2, entropy)
```

## Plot metadata 

Then we plot intra-sample heterogeneity together with sample-specific metadata for the clustered sammples.

```{r plot-metadata}
#access sample-level data
sample_summary <- as.data.frame(colData(sce)) %>%
  group_by(sample_id) %>%
  summarize(mna = first(mna), 
            progress=first(progress),
            age_diag = first(age_diag),
            condition = first(condition)
  )

sample_summary <- as.data.frame(sample_summary)
rownames(sample_summary) <- sample_summary$sample_id
sample_summary <- sample_summary %>% select(-sample_id)


#assign color codes
col_code <- c("MNA"=rgb(202,0,32, maxColorValue = 255),
              "het"=rgb(146,197,222, maxColorValue = 255),
              "nMNA"=rgb(5,113,176, maxColorValue = 255),
              "Prog"=rgb(202,0,32, maxColorValue = 255),
              "NoProg"=rgb(5,113,176, maxColorValue = 255)
              )

col_age_diag = colorRamp2(c(min(sample_summary$age_diag, na.rm = TRUE), (quantile(sample_summary$age_diag, 0.95, na.rm = TRUE)-min(sample_summary$age_diag, na.rm = TRUE))/2, quantile(sample_summary$age_diag,0.95, na.rm = TRUE)), viridis(3))
col_me_het = colorRamp2(c(min(me_het, na.rm = TRUE), (max(me_het, na.rm = TRUE)-min(me_het, na.rm = TRUE))/2, max(me_het, na.rm = TRUE)), viridis(3))
col_tumor_het = colorRamp2(c(min(tumor_het, na.rm = TRUE), (max(tumor_het, na.rm = TRUE)-min(tumor_het, na.rm = TRUE))/2, max(tumor_het, na.rm = TRUE)), viridis(3))

# Plot metadata
columnAnnotation(MYCN_amp=sample_summary$mna, col=list(MYCN_amp=col_code)) %v%
  columnAnnotation(progression=sample_summary$progress, col=list(progression=col_code)) %v%
  columnAnnotation(age=sample_summary$age_diag, col=list(age=col_age_diag)) %v%
  columnAnnotation(TUM_het=tumor_het, col=list(TUM_het=col_tumor_het)) %v%
  columnAnnotation(ME_het=me_het, col=list(ME_het=col_me_het))
  
```

## Plot continuous and discrete summary statistics

We then plot summary statistics of continuous and discrete metadata in boxplots and piecharts, respectively.

```{r piecharts, fig.height=4, fig.width=5}
discrete_metadata <- c("mna", "progress")

discrete_metadata_colors <- list(col_code, col_code)

plot_list <- list()
for (d in seq_along(discrete_metadata)) {
  sample_meta <- sample_summary[,c(discrete_metadata[d], "condition")]
  sample_meta <- sample_meta[rowSums(sample_meta == "unknown") == 0, ] 
  colnames(sample_meta) <- c("dis_var", "condition")
  
  result <- sample_meta %>%
    group_by(condition, dis_var) %>%
    summarise(Freq = n(), .groups = 'drop')
  
  result[,"dis_var"]<- as.factor(unlist(result[,"dis_var"]))
  
plot_list[[d]] <- ggplot(result, aes(x="", y=Freq, fill=dis_var, group=dis_var)) +
    geom_col(position = "fill") +
    scale_fill_manual(values = discrete_metadata_colors[[d]], name=discrete_metadata[d]) +
    coord_polar("y", start=0)  +  
    facet_grid(.~ condition) +
    theme_void()
}

do.call("grid.arrange", c(plot_list, ncol=1))
```
```{r boxplots, fig.width=8, fig.height=6}
het_t <- as.vector(unname(tumor_het[clustered_samples]))
het_me <- as.vector(unname(me_het[clustered_samples]))
perc_tumor <- round((as.vector(unname(table(sce_tumor$sample_id)[clustered_samples]))/as.vector(unname(table(sce$sample_id)[clustered_samples])))*100,3)
condition <- sample_summary$condition

cont_df <- data.frame(cbind(het_t, perc_tumor, me_het, condition),row.names=clustered_samples)

plot_list <- list()

for (d in colnames(cont_df)[1:ncol(cont_df)-1]) {

  df <- cont_df[,c(d, "condition")] 
  
  colnames(df) <- c("y", "condition")
  stats_df <- compare_means(y ~ condition,  
                        p.adjust.method = "BH", 
                        method='wilcox.test', 
                        data = df)
  
  stats_df$p_adj_signif <- stars.pval(stats_df$p.adj)
  stats_df$p_adj_signif[stats_df$p_adj_signif==" "] <- "ns"
  stats_df$p_long <- paste0(stats_df$p.adj, " (", stats_df$p_adj_signif, ")")

  if (d == "perc_tumor"){
    plot_list[[d]] <- ggboxplot(df, x = "condition", y = d,
            color = "condition", palette =c("skyblue2", "slategray1", "slategray3", "slategray"), 
            add.params = list(size = 1.5),
            add = "jitter")+ 
            stat_pvalue_manual(stats_df, label = "p_long", y.position = log10(max(df$y)), step.increase = 0.2,) +
            theme(aspect.ratio = 2, legend.position="none") +
            scale_y_log10(
                breaks = scales::trans_breaks("log10", function(x) 10^x),
                labels = scales::trans_format("log10", scales::math_format(10^.x))
            )
  }else{
    plot_list[[d]] <- ggboxplot(df, x = "condition", y = d,
                color = "condition", palette =c("skyblue2", "slategray1", "slategray3", "slategray"), 
                add.params = list(size = 1.5),
                add = "jitter")+ 
                stat_pvalue_manual(stats_df, label = "p_long", y.position = max(df$y), step.increase = 0.2,) +
                theme(aspect.ratio = 2, legend.position="none") 
  }
}

do.call("grid.arrange", c(plot_list, ncol=3))
```

## Plot metaclusters

Finally, we plot the proportions of metaclusters in the clustered samples.

```{r plot-metaclusters, fig.width=14, figh.height=5}
sce_tmp <- sce
sce_tmp$cluster_id <- sce_tmp$metacluster
metadata(sce_tmp)$cluster_codes <- data.frame(celltype = factor(sce_tmp$metacluster))
p_all <- plotAbundances(sce_tmp, k = "celltype", by = "sample", col_clust=F, k_pal=metadata(spe)$color_vectors$col_metacluster)
rm(sce_tmp)
p_all
```

## Session Info

<details>
   <summary>SessionInfo</summary>
   
```{r, echo = FALSE}
sessionInfo()
```
</details>
