---
title: "Heatmap - Cytokine Assays"
author: "LazDaria"
date: "`r Sys.Date()`"
output: 
  workflowr::wflow_html
editor_options:
  chunk_output_type: inline
params: 
  path: "/mnt/Multimodal_Imaging_Daria/PIPELINE/NEWEST_RESULTS/results/Manuscript/cytokines"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      error=FALSE,
                      messages=FALSE,
                      warning=FALSE)
```


# Load libraries
<details>
   <summary>Load libraries</summary>
   
```{r load-libraries}
library(readxl)
library(dplyr)
library(stringr)
library(tidyr)
library(ComplexHeatmap)
library(viridis)
library(openxlsx)
library(bruceR)
```
</details>

## Read data
First, we load normalized data from three cytokine assays.

```{r read-data}
data_inf <- read_excel(file.path(params$path, 'array_all.xlsx'), sheet = "INF3")
data_che <- read_excel(file.path(params$path, 'array_all.xlsx'), sheet = "CHE1")
data_rec <- read_excel(file.path(params$path, 'array_all.xlsx'), sheet = "REC1")

#Remove positive controls
data_inf <- data_inf[!grepl("POS", data_inf$Normalization),]
data_che <- data_che[!grepl("POS", data_che$Normalization),]
data_rec <- data_rec[!grepl("POS", data_rec$Normalization),]

head(data_che)
```
## Rename column names

```{r rename-col}
# Rename columns: remove dots and numbers, replace spaces with underscores
clean_colnames <- function(colname) {
  colname %>%
    str_replace_all("\\.\\d*$", "") %>%  # Remove trailing numbers after a dot
    str_replace_all("\\.\\.\\.+\\d*", "") %>%  # Remove dots and trailing numbers
    str_replace_all(" ", "_")                  # Replace spaces with underscores
}

# Apply the renaming function
colnames(data_inf) <- sapply(colnames(data_inf), clean_colnames)
colnames(data_che) <- sapply(colnames(data_che), clean_colnames)
colnames(data_rec) <- sapply(colnames(data_rec), clean_colnames)
```


# Heatmap for cytokine assay {#cyto}
In this chapter, we will generate a heatmap for cytokine assays performed using RayBioTech. 

## Compute mean between duplicate rows

```{r mean-duplicate-rows}
# Combine duplicate columns by calculating the mean
combine_duplicate_columns <- function(df) {
  df %>%
     pivot_longer(-Normalization, names_to = "colname", values_to = "value") %>%
     group_by(Normalization, colname) %>%
     summarise(mean_value = mean(value, na.rm = TRUE), .groups = 'drop') %>%
     pivot_wider(names_from = colname, values_from = mean_value)
}

# Apply the function to combine duplicate columns
data_combined_inf <- combine_duplicate_columns(data_inf)
data_combined_che <- combine_duplicate_columns(data_che)
data_combined_rec <- combine_duplicate_columns(data_rec)

# Remove dots from the column names
colnames(data_combined_inf) <- str_replace_all(colnames(data_combined_inf), "\\.", "")
colnames(data_combined_che) <- str_replace_all(colnames(data_combined_che), "\\.", "")
colnames(data_combined_rec) <- str_replace_all(colnames(data_combined_rec), "\\.", "")

#Set first column as rownames
data_combined_inf <- as.data.frame(data_combined_inf)
rownames(data_combined_inf) <- data_combined_inf$Normalization
data_combined_inf <- data_combined_inf[-1]

data_combined_che <- as.data.frame(data_combined_che)
rownames(data_combined_che) <- data_combined_che$Normalization
data_combined_che <- data_combined_che[-1]

data_combined_rec <- as.data.frame(data_combined_rec)
rownames(data_combined_rec) <- data_combined_rec$Normalization
data_combined_rec <- data_combined_rec[-1]
```

## Filter out columns and rows

```{r filter}
col_filter <- c("I", "PBMC_I", "CHLA", "CHLA_D", "R", "PBMC_R", "CLB", "CLB_D")
rows_data <- read_excel(file.path(params$path, 'array_all.xlsx'), sheet = "filter")
rows_inf <- rows_data[rows_data$Array=="INF-3","Secreted factor"]$`Secreted factor`
rows_che <- rows_data[rows_data$Array=="CHE-1","Secreted factor"]$`Secreted factor`
rows_rec <- rows_data[rows_data$Array=="REC-1","Secreted factor"]$`Secreted factor`
setdiff(rows_che, rownames(data_combined_che))
```
## Create heatmap 

### Plot for one array and one cell line only

```{r heatmap-inf-filtered, fig.width=5, fig.height=3}
df <- as.data.frame(data_combined_inf)

cols_filter <- c("I", "PBMC_I", "CHLA", "CHLA_D")

Heatmap(scale(t(df[rows_inf, cols_filter])), 
        column_title = "INF-3-filtered",
        row_order = cols_filter,
        col=viridis(100)
        )
```

```{r heatmap-inf, fig.width=8, fig.height=3}
df <- as.data.frame(data_combined_inf)

Heatmap(scale(t(df[,cols_filter])), 
        column_title = "INF-3_all",
        row_order = cols_filter,
        col=viridis(100)
        )
```
### Plot for all arrays and all cell lines combined

```{r heatmap-combined-all-sepscale, fig.width=19, fig.height=3}
data_all <- rbind(data_combined_inf, data_combined_che, data_combined_rec)
df <- as.data.frame(data_all)

#Annotation
annot <- c(rep("INF-3", nrow(data_combined_inf)), rep("CHE-1", nrow(data_combined_che)), rep("REC-1", nrow(data_combined_rec)))
col_vector <- c("INF-3"="#fc8d59", "CHE-1"="#ffffbf", "REC-1"="#91bfdb")
column_ha <- HeatmapAnnotation(assay = annot, col=list(assay=col_vector))

#Separate scaling per cel line
col_filter_chla <- c("I", "PBMC_I", "CHLA", "CHLA_D")
col_filter_clb <- c("R", "PBMC_R", "CLB", "CLB_D")
data_sep_scale <- rbind(scale(t(df[,col_filter_chla])), scale(t(df[,col_filter_clb])))
# chla_scaled <- apply(df[,col_filter_chla], 1, RESCALE, to=0:1)
# clb_scaled <- apply(df[,col_filter_clb], 1, RESCALE, to=0:1)
# data_sep_scale <- rbind(chla_scaled, clb_scaled )

Heatmap(data_sep_scale, 
        column_title = "all assays-all- CHLA90 and CLB-MA",
        row_order = c("I", "R", "PBMC_I", "PBMC_R", "CHLA", "CLB", "CHLA_D", "CLB_D"),
        cluster_columns =F,
        top_annotation = column_ha,
        col=viridis(100)
        )
```

```{r heatmap-combined-clustered-all, fig.width=19, fig.height=5}
Heatmap(data_sep_scale, 
        column_title = "all assays-all- CHLA90 and CLB-MA",
        row_order = c("I", "R", "PBMC_I", "PBMC_R", "CHLA", "CLB", "CHLA_D", "CLB_D"),
        cluster_columns =T,
        top_annotation = column_ha,
        col=viridis(100)
        )
```
```{r write-csv}
#write unscaled data
df_unscaled <- cbind(df[,col_filter_chla], df[,col_filter_clb])
write.csv(df_unscaled[,c("I", "R", "PBMC_I", "PBMC_R", "CHLA", "CLB", "CHLA_D", "CLB_D")],file.path(params$path, 'heatmap.csv'))
write.xlsx(df_unscaled[,c("I", "R", "PBMC_I", "PBMC_R", "CHLA", "CLB", "CHLA_D", "CLB_D")],file.path(params$path, 'heatmap.xlsx'), , rowNames=TRUE)
```

## Session Info

<details>
   <summary>SessionInfo</summary>
   
```{r, echo = FALSE}
sessionInfo()
```
</details>
