---
title: "Heatmap - Cytokine Assays"
author: "LazDaria"
date: "`r Sys.Date()`"
output: 
  workflowr::wflow_html
editor_options:
  chunk_output_type: inline
params: 
  path: "/mnt/Multimodal_Imaging_Daria/PIPELINE/NEWEST_RESULTS/results/Manuscript/cytokines"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      error=FALSE,
                      messages=FALSE,
                      warning=FALSE)
```


# Load libraries
<details>
   <summary>Load libraries</summary>
   
```{r load-libraries}
library(readxl)
library(dplyr)
library(stringr)
library(tidyr)
library(ComplexHeatmap)
library(viridis)
library(openxlsx)
library(bruceR)
```
</details>

## Read data
First, we load normalized data from three cytokine assays.

```{r read-data}
data_inf <- read_excel(file.path(params$path, 'array_all.xlsx'), sheet = "INF3")
data_che <- read_excel(file.path(params$path, 'array_all.xlsx'), sheet = "CHE1")
data_rec <- read_excel(file.path(params$path, 'array_all.xlsx'), sheet = "REC1")

#Remove positive controls
data_inf <- data_inf[!grepl("POS", data_inf$Normalization),]
data_che <- data_che[!grepl("POS", data_che$Normalization),]
data_rec <- data_rec[!grepl("POS", data_rec$Normalization),]

head(data_che)
```
## Rename column names

```{r rename-col}
# Rename columns: remove dots and numbers, replace spaces with underscores
clean_colnames <- function(colname) {
  colname %>%
    str_replace_all("\\.\\d*$", "") %>%  # Remove trailing numbers after a dot
    str_replace_all("\\.\\.\\.+\\d*", "") %>%  # Remove dots and trailing numbers
    str_replace_all(" ", "_")                  # Replace spaces with underscores
}

# Apply the renaming function
colnames(data_inf) <- sapply(colnames(data_inf), clean_colnames)
colnames(data_che) <- sapply(colnames(data_che), clean_colnames)
colnames(data_rec) <- sapply(colnames(data_rec), clean_colnames)
```


# Heatmap for cytokine assay {#cyto}
In this chapter, we will generate a heatmap for cytokine assays performed using RayBioTech. 

## Compute mean between duplicate rows

```{r mean-duplicate-rows}
# Combine duplicate columns by calculating the mean
combine_duplicate_columns <- function(df) {
  df %>%
     pivot_longer(-Normalization, names_to = "colname", values_to = "value") %>%
     group_by(Normalization, colname) %>%
     summarise(mean_value = mean(value, na.rm = TRUE), .groups = 'drop') %>%
     pivot_wider(names_from = colname, values_from = mean_value)
}

# Apply the function to combine duplicate columns
data_combined_inf <- combine_duplicate_columns(data_inf)
data_combined_che <- combine_duplicate_columns(data_che)
data_combined_rec <- combine_duplicate_columns(data_rec)

# Remove dots from the column names
colnames(data_combined_inf) <- str_replace_all(colnames(data_combined_inf), "\\.", "")
colnames(data_combined_che) <- str_replace_all(colnames(data_combined_che), "\\.", "")
colnames(data_combined_rec) <- str_replace_all(colnames(data_combined_rec), "\\.", "")

#Set first column as rownames
data_combined_inf <- as.data.frame(data_combined_inf)
rownames(data_combined_inf) <- data_combined_inf$Normalization
data_combined_inf <- data_combined_inf[-1]

data_combined_che <- as.data.frame(data_combined_che)
rownames(data_combined_che) <- data_combined_che$Normalization
data_combined_che <- data_combined_che[-1]

data_combined_rec <- as.data.frame(data_combined_rec)
rownames(data_combined_rec) <- data_combined_rec$Normalization
data_combined_rec <- data_combined_rec[-1]
```

## Filter out columns and rows

```{r filter}
col_filter <- c("I", "PBMC_I", "CHLA", "CHLA_D", "R", "PBMC_R", "CLB", "CLB_D")
rows_data <- read_excel(file.path(params$path, 'array_all.xlsx'), sheet = "filter")
rows_inf <- rows_data[rows_data$Array=="INF-3","Secreted factor"]$`Secreted factor`
rows_che <- rows_data[rows_data$Array=="CHE-1","Secreted factor"]$`Secreted factor`
rows_rec <- rows_data[rows_data$Array=="REC-1","Secreted factor"]$`Secreted factor`
setdiff(rows_che, rownames(data_combined_che))
```
## Create heatmap 

### Plot for one array and one cell line only

```{r heatmap-inf-filtered, fig.width=5, fig.height=3}
df <- as.data.frame(data_combined_inf)

cols_filter <- c("I", "PBMC_I", "CHLA", "CHLA_D")

Heatmap(scale(t(df[rows_inf, cols_filter])), 
        column_title = "INF-3-filtered",
        row_order = cols_filter,
        col=viridis(100)
        )
```

```{r heatmap-inf, fig.width=8, fig.height=3}
df <- as.data.frame(data_combined_inf)

Heatmap(scale(t(df[,cols_filter])), 
        column_title = "INF-3_all",
        row_order = cols_filter,
        col=viridis(100)
        )
```
### Plot for all arrays and all cell lines combined

```{r heatmap-combined-all-sepscale, fig.width=19, fig.height=3}
data_all <- rbind(data_combined_inf, data_combined_che, data_combined_rec)
df <- as.data.frame(data_all)

#Annotation
annot <- c(rep("INF-3", nrow(data_combined_inf)), rep("CHE-1", nrow(data_combined_che)), rep("REC-1", nrow(data_combined_rec)))
col_vector <- c("INF-3"="#fc8d59", "CHE-1"="#ffffbf", "REC-1"="#91bfdb")
column_ha <- HeatmapAnnotation(assay = annot, col=list(assay=col_vector))

#Separate scaling per cel line
col_filter_chla <- c("I", "PBMC_I", "CHLA", "CHLA_D")
col_filter_clb <- c("R", "PBMC_R", "CLB", "CLB_D")
data_sep_scale <- rbind(scale(t(df[,col_filter_chla])), scale(t(df[,col_filter_clb])))
# chla_scaled <- apply(df[,col_filter_chla], 1, RESCALE, to=0:1)
# clb_scaled <- apply(df[,col_filter_clb], 1, RESCALE, to=0:1)
# data_sep_scale <- rbind(chla_scaled, clb_scaled )

Heatmap(data_sep_scale, 
        column_title = "all assays-all- CHLA90 and CLB-MA",
        row_order = c("I", "R", "PBMC_I", "PBMC_R", "CHLA", "CLB", "CHLA_D", "CLB_D"),
        cluster_columns =F,
        top_annotation = column_ha,
        col=viridis(100)
        )
```

```{r heatmap-combined-clustered-all, fig.width=19, fig.height=5}
Heatmap(data_sep_scale, 
        column_title = "all assays-all- CHLA90 and CLB-MA",
        row_order = c("I", "R", "PBMC_I", "PBMC_R", "CHLA", "CLB", "CHLA_D", "CLB_D"),
        cluster_columns =T,
        top_annotation = column_ha,
        col=viridis(100)
        )
```
We will then calculate the log2FC between mono- and co-cultures.
```{r log2FC}
df_unscaled <- cbind(df[,col_filter_chla], df[,col_filter_clb])

# Create the new columns with the specified calculations
df_unscaled <- df_unscaled %>%
  mutate(
    log2FC_CHLA_D_PBMC_I = log2(CHLA_D / PBMC_I),
    log2FC_CLB_D_PBMC_R = log2(CLB_D / PBMC_R),
    log2FC_CHLA_D_CHLA = log2(CHLA_D / CHLA),
    log2FC_CLB_D_CLB = log2(CLB_D / CLB)
  )

# Add the new column based on the condition
df_unscaled <- df_unscaled %>%
  mutate(threshold = if_else(
    log2FC_CHLA_D_PBMC_I > 0.5 & 
    log2FC_CLB_D_PBMC_R > 0.5 & 
    log2FC_CHLA_D_CHLA > 0.5 & 
    log2FC_CLB_D_CLB > 0.5, 
    1, 
    0
  ))
```

Let's plot the heatmap adding this annotation.
```{r heatmap-threshold, fig.width=19, fig.height=5}
#Annotation
column_ha <- HeatmapAnnotation(assay = annot,
                               threshold = as.factor(df_unscaled$threshold),
                               col=list(assay=col_vector))

#Separate scaling per cel line
col_filter_chla <- c("I", "PBMC_I", "CHLA", "CHLA_D")
col_filter_clb <- c("R", "PBMC_R", "CLB", "CLB_D")
data_sep_scale <- rbind(scale(t(df_unscaled[,col_filter_chla])), scale(t(df_unscaled[,col_filter_clb])))

h <- Heatmap(data_sep_scale, 
        column_title = "all assays-all- CHLA90 and CLB-MA",
        row_order = c("I", "R", "PBMC_I", "PBMC_R", "CHLA", "CLB", "CHLA_D", "CLB_D"),
        cluster_columns =T,
        top_annotation = column_ha,
        col=viridis(100)
        )

h
```

The data can then be written to memory for annotation by experts. 

```{r write-csv}
#write unscaled data
write.csv(df_unscaled[,c("I", "R", "PBMC_I", "PBMC_R", "CHLA", "CLB", "CHLA_D", "CLB_D","log2FC_CHLA_D_PBMC_I", "log2FC_CLB_D_PBMC_R", "log2FC_CHLA_D_CHLA", "log2FC_CLB_D_CLB", "threshold")],file.path(params$path, 'heatmap_th.csv'))
write.xlsx(df_unscaled[,c("I", "R", "PBMC_I", "PBMC_R", "CHLA", "CLB", "CHLA_D", "CLB_D","log2FC_CHLA_D_PBMC_I", "log2FC_CLB_D_PBMC_R", "log2FC_CHLA_D_CHLA", "log2FC_CLB_D_CLB", "threshold")],file.path(params$path, 'heatmap_th.xlsx') , rowNames=TRUE)
```

The excel sheet is then read again after annotation.

```{r}
data_annotated <- read.xlsx(file.path(params$path, 'heatmap_annotated.xlsx'), rowNames=TRUE, sheet="PBMC and Direct")[,1:15]
```

We then plot the heatmap showing the annotation about literature on T cell exhaustion.

```{r heatmap-exhaustion, fig.width=19, fig.height=5}
#sort data by clustering from heatmap before
order <- colnames(data_sep_scale)[h@column_order] #row_order is wrong - don'T know why
df_ordered <- data_annotated[order, ]
annot <- annot[order]

increased_cytokines <- df_ordered$threshold==1
df_increased <- df_ordered[increased_cytokines,]
  
#Annotation
column_ha <- HeatmapAnnotation(assay = annot[increased_cytokines],
                               threshold = as.factor(df_increased$threshold),
                               T_cell_exhaustion = as.factor(df_increased$T.cell.exhaustion),
                               col=list(assay=col_vector))

#Separate scaling per cel line
col_filter_chla <- c("I", "PBMC_I", "CHLA", "CHLA_D")
col_filter_clb <- c("R", "PBMC_R", "CLB", "CLB_D")
data_sep_scale <- rbind(scale(t(df_increased[,col_filter_chla])), scale(t(df_increased[,col_filter_clb])))

Heatmap(data_sep_scale, 
        column_title = "all assays-all- CHLA90 and CLB-MA",
        row_order = c("I", "R", "PBMC_I", "PBMC_R", "CHLA", "CLB", "CHLA_D", "CLB_D"),
        cluster_columns =F,
        top_annotation = column_ha,
        col=viridis(100)
        )
```

## Session Info

<details>
   <summary>SessionInfo</summary>
   
```{r, echo = FALSE}
sessionInfo()
```
</details>
